<!--
The network analysis itself. This file is the single source for three
notebooks, which differ only in which predictability nodes they keep:

  network-analysis.qmd                     drop_var <- NULL         (15 nodes)
  sensitivity/network-no-varloss.qmd       drop_var <- "var_loss"   (14 nodes)
  sensitivity/network-no-meanloss.qmd      drop_var <- "mean_loss"  (14 nodes)

Each parent sets `drop_var` and then includes this file. Nothing below names
mean_loss or var_loss directly; the code works from `pred_vars` (retained) and
`focal` (the first retained node). See the `resolve-drop` chunk.

The leading underscore keeps Quarto's project render from treating this as a
document in its own right, and include paths are relative to the parent, so the
sensitivity notebooks include `../_network-body.qmd`.
-->

# Section 1. Setup and Data Preparation

```{r}
#| label: setup
library(tidyverse)  # always
library(psych)
library(corrplot)
library(qgraph)      # visualization
library(bootnet)     # edge-weight accuracy CIs + group viz networks
library(igraph)
library(EGAnet)      # estimation, communities, network loadings, network.compare
library(lavaan)

# NOTE: EGAnet's signatures have shifted across 2.x releases. Using EGAnet 2.4.1.
```

```{r}
#| label: resolve-drop

# `drop_var` is set by the parent notebook before this body is included. NULL
# keeps the full battery (the main analysis); a variable name holds that one
# predictability node out (a sensitivity run). Everything downstream works from
# `pred_vars` (what is retained) and `focal` (the first retained node) instead
# of naming mean_loss or var_loss, so the three notebooks differ only here.
stopifnot(exists("drop_var"))
pred_all <- c("mean_loss", "var_loss")
stopifnot(is.null(drop_var) ||
          (length(drop_var) == 1 && drop_var %in% pred_all))

pred_vars <- setdiff(pred_all, drop_var)
focal <- pred_vars[1]

# Sensitivity runs write to their own directory under a run-specific prefix, so
# they can never overwrite the figures the main analysis produces.
if (is.null(drop_var)) {
  fig_dir <- "fig"
  fig_prefix <- ""
} else {
  fig_dir <- "fig/sensitivity"
  fig_prefix <- paste0("no-", gsub("_", "", drop_var), "_")
}
dir.create(fig_dir, showWarnings = FALSE, recursive = TRUE)
fig_path <- function(stem) file.path(fig_dir, paste0(fig_prefix, stem))

# A parent may pin `n_cores` so that concurrent renders do not oversubscribe
# the machine. Left alone, this is what the analysis has always used.
if (!exists("n_cores")) n_cores <- parallel::detectCores() - 1

cat("Held out:", if (is.null(drop_var)) "nothing (full battery)" else drop_var, "\n")
cat("Predictability nodes retained:", paste(pred_vars, collapse = ", "), "\n")
cat("Figures ->", fig_path("*.pdf"), "| bootstrap cores:", n_cores, "\n")
```

```{r}
#| label: load-data
# ellipse_metrics.csv carries the 16 linguistic features + ELLIPSE scores +
# predictability (mean_loss, var_loss) from the Study 1 benchmark model
# (Llama-3.1-8B base). All networks below use Spearman correlations.
df_raw <- read.csv("data/ellipse/ellipse_metrics.csv")

cat("Raw data:", nrow(df_raw), "rows x", ncol(df_raw), "columns\n")
```

## A note on clauses per T-unit

`clauses_per_tunit` (C/T) is computed and retained in the metrics table but is
**not entered into the network**. On this corpus it does not measure
subordination; it measures punctuation.

The T-unit is the only denominator in the battery that depends on sentence
segmentation, and spaCy segments on punctuation. An unpunctuated run-on has
multiple independent clauses that are treated as subordinate by the parser.

- C/T correlates −0.205 with Overall proficiency across all essays, but −0.014
  among well-punctuated essays only (8–25 words per mark, *n* = 3,590). The
  entire association is the confound.


```{r}
#| label: select-features

# Assemble the feature set: 15 variables, less any held-out predictability node.
#
# The three syntactic indices (src/features/syntactic.py) sample distinct
# subconstructs rather than accumulating redundant omnibus measures (Norris &
# Ortega, 2009; Bulté & Housen, 2012): length at the clausal level (MLC), and
# phrasal elaboration and phrasal coordination at the NP level (CN/C, CP/C).
# All three are L2SCA indices (Lu, 2010, 2011), so the battery is directly
# comparable to the large body of work using that tool, and one clause
# definition serves all three denominators.
#
# Three further indices are in the CSV but deliberately NOT entered:
#   - `clauses_per_tunit`  -- a punctuation index here; see the note above.
#   - `deps_per_clause`    -- (Kyle & Crossley, 2018) indexes the same
#                             clausal-length subconstruct as MLC, which is the
#                             far more widely reported measure.
#   - `deps_per_nominal`   -- (Kyle & Crossley, 2018) behaved as an island,
#                             loading -.098 on its own community and
#                             correlating ~.00 with every ELLIPSE subscore.
feature_vars <- c(
  # Lexical (4)
  "mtld", "lexical_density", "log_mean_token_freq", "mean_word_length",
  # Syntactic (3)
  "mean_length_of_clause",
  "complex_nominals_per_clause", "coord_phrases_per_clause",
  # Phraseological (3)
  "amod_mi", "dobj_mi", "advmod_mi",
  # Cohesion (3)
  "content_word_overlap", "connective_density", "sentence_similarity",
  # Predictability (2, less whatever `drop_var` holds out)
  pred_vars
)

df_features <- df_raw[, feature_vars]

# Variance check: a constant variable cannot enter a network. (An early version
# of the syntactic battery produced an index that was constant at 1 and had to
# be caught here, so every battery since is checked rather than assumed.)
var_check <- data.frame(
  Variable = feature_vars,
  Unique = sapply(df_features, function(x) length(unique(na.omit(x)))),
  SD = round(sapply(df_features, sd, na.rm = TRUE), 4)
)
degenerate <- var_check$Variable[var_check$SD == 0 | is.na(var_check$SD)]
if (length(degenerate) > 0) {
  cat("WARNING: zero-variance variables:", paste(degenerate, collapse = ", "), "\n")
  print(var_check[var_check$Variable %in% degenerate, ], row.names = FALSE)
} else {
  cat("All", length(feature_vars), "variables have non-zero variance.\n")
}
```

```{r}
#| label: missing-data

# Report missingness. The few missing values are structural: dependency-MI
# features (amod/dobj/advmod) are undefined for bigrams unseen in the Dolma
# reference corpus, and content_word_overlap / sentence_similarity are undefined
# for single-sentence essays. Rather than listwise-delete (which biases the
# sample toward "typical" essays), we retain every essay and use
# pairwise-complete correlations throughout (Isvoranu & Epskamp, 2023).
missing_counts <- colSums(is.na(df_features))
cat("Missing values per variable (retained; handled pairwise):\n")
print(missing_counts[missing_counts > 0])
cat("\nEssays retained:", nrow(df_features),
    "(", sum(!complete.cases(df_features)), "have >=1 pairwise-handled NA)\n")
```

```{r}
#| label: residualize-on-prompt

# Residualize every variable on writing prompt to remove systematic
# prompt-driven variance. na.action = na.exclude keeps each residual vector
# full length with NA left in the originally-missing positions, so the
# downstream pairwise-complete correlations can use every available essay.
df_net <- df_features
r2_values <- numeric(ncol(df_features))
names(r2_values) <- names(df_features)

for (var in names(df_features)) {
  fit <- lm(df_features[[var]] ~ factor(df_raw$prompt), na.action = na.exclude)
  df_net[[var]] <- residuals(fit)
  r2_values[var] <- summary(fit)$r.squared
}

cat("R² (variance explained by prompt) per variable:\n")
r2_df <- data.frame(Variable = names(r2_values), R2 = round(r2_values, 4))
r2_df <- r2_df[order(-r2_df$R2), ]
print(r2_df, row.names = FALSE)
```

## Local dependence screening (UVA)

The syntactic battery is not independent by construction. All three indices share
the clause as a denominator, and MLC's numerator (the word count) necessarily
includes the postmodifiers that make a nominal complex and the conjuncts that
make a coordinate phrase, so MLC is structurally entangled with CN/C and CP/C.
Local dependence is therefore worth measuring.

Unique Variable Analysis (Christensen, Garrido & Golino, 2023) quantifies it via
weighted topological overlap. This section **reports** wTO and does not act on
it. No variable is removed or combined here, and the conventional .25 cutoff is
printed only as a reference line. Local dependence is evidence to weigh rather
than a decision procedure. Which indices belong in the network is a question
about constructs, and a topological-overlap statistic cannot settle it. Where a
pair looks dependent, the reading of that pair's shared community is qualified in
the discussion instead of the pair being pruned.

```{r}
#| label: uva-screen

# Screen on the residualized data, since that is what every network below uses,
# with the same pairwise missing-data handling as every other estimate here.
# reduce = FALSE: report the overlap only. Nothing downstream consumes this.
uva <- UVA(data = df_net, cut.off = 0.25, reduce = FALSE,
           na.data = "pairwise", verbose = FALSE)

wto_pairs <- uva$wto$pairwise
wto_pairs <- wto_pairs[order(-wto_pairs$wto), ]

cat("Highest weighted topological overlap:", round(max(wto_pairs$wto), 3), "\n")
cat("Pairs at or above the .25 reference line:",
    sum(wto_pairs$wto >= 0.25), "of", nrow(wto_pairs), "\n\n")

cat("Top 12 pairs by wTO (reported, not acted on):\n")
print(head(wto_pairs, 12), row.names = FALSE)
```

```{r}
#| label: define-battery

# Every variable in feature_vars enters the network; UVA above is diagnostic.
syntactic_battery <- c("mean_length_of_clause",
                       "complex_nominals_per_clause", "coord_phrases_per_clause")

cat("Variables entering the network:", ncol(df_net), "\n")
```

```{r}
#| label: define-groups

# Theoretical groupings (construct membership).
construct_members <- list(
  Lexical        = c("mtld", "lexical_density", "log_mean_token_freq", "mean_word_length"),
  Syntactic      = syntactic_battery,
  Phraseological = c("amod_mi", "dobj_mi", "advmod_mi"),
  Cohesion       = c("content_word_overlap", "connective_density", "sentence_similarity"),
  Predictability = pred_vars
)

# Canonical node order, shared by the full-sample and proficiency-group networks.
node_order <- feature_vars

# Short labels for plotting
short_labels <- c(
  mtld = "MTLD",
  lexical_density = "LexDen",
  log_mean_token_freq = "TokFreq",
  mean_word_length = "WordLen",
  clauses_per_tunit = "C/T",
  mean_length_of_clause = "MLC",
  complex_nominals_per_clause = "CN/C",
  coord_phrases_per_clause = "CP/C",
  amod_mi = "AmodMI",
  dobj_mi = "DobjMI",
  advmod_mi = "AdvMI",
  content_word_overlap = "CWO",
  connective_density = "ConnDen",
  sentence_similarity = "SentSim",
  mean_loss = "Loss",
  var_loss = "VarLoss"
)

net_labels <- short_labels[node_order]

# --- Visual encoding: COLOUR = theoretical construct ------------------------
# Shape is reserved for the empirically derived EGA communities (Section 3),
# which is the reading the figures should give up first and the only channel
# that survives black-and-white printing intact. The construct palette is
# therefore deliberately subdued: low chroma so it never competes with the
# silhouettes, but with the five lightness values spread evenly so the hues
# still separate into distinguishable greys when the colour is thrown away.
construct_colors <- setNames(
  hcl(h = c(70, 250, 140, 320, 20), c = 20, l = c(94, 85, 76, 67, 58)),
  names(construct_members)
)

node_construct <- rep(NA_character_, length(node_order))
names(node_construct) <- node_order
for (g in names(construct_members)) {
  node_construct[construct_members[[g]]] <- g
}
node_construct <- node_construct[node_order]
node_colors <- unname(construct_colors[node_construct])

# Greyscale check: relative luminance of each construct colour, on 0-100.
# Neighbouring constructs should stay ~8-9 points apart.
cat("Construct palette and its greyscale equivalent:\n")
print(data.frame(
  Construct = names(construct_colors),
  Hex = unname(construct_colors),
  Grey = sapply(construct_colors, function(x) {
    rgb <- col2rgb(x) / 255
    round(100 * (0.2126 * rgb[1] + 0.7152 * rgb[2] + 0.0722 * rgb[3]), 1)
  })
), row.names = FALSE)
```

```{r}
#| label: define-shapes

# --- Visual encoding: SHAPE = empirical EGA community -----------------------
# Communities are assigned shapes from this pool in order, so the encoding
# still works if a re-run returns a different number of communities.
#
# qgraph ships only circle/square/triangle/diamond as built-in shapes, and its
# built-in "diamond" draws the fill (pch 18) smaller than the border (pch 5),
# leaving a white ring. Both the rhombus and the 5-pointed star are therefore
# supplied as custom polygons (coordinates on [-1, 1], scaled to the node box).
shape_polygons <- list(
  rhombus = list(x = c(0, 1, 0, -1), y = c(1, 0, -1, 0)),
  star = local({
    ang <- seq(pi / 2, pi / 2 + 2 * pi, length.out = 11)[1:10]
    list(x = rep(c(1, 0.6), 5) * cos(ang), y = rep(c(1, 0.6), 5) * sin(ang))
  })
)

# Ordered most- to least-distinguishable at small sizes
shape_pool <- c("circle", "square", "triangle", "rhombus", "star")

# Shapes with less usable interior than a circle are drawn larger so that the
# labels stay legible and the *apparent* node areas stay comparable. Size
# carries no information here.
shape_scale <- c(circle = 1, square = 1, triangle = 1.3, rhombus = 1.15, star = 1.4)

# Plotting-character equivalents, for the shape half of the legend
shape_pch <- c(circle = 21, square = 22, triangle = 24, rhombus = 23, star = 8)
```


# Section 2. Descriptive Statistics and Correlation Matrix

```{r}
#| label: descriptive-stats
#| tbl-cap: "Descriptive statistics for all network variables"

desc_stats <- describe(df_net)
print(desc_stats[, c("n", "mean", "sd", "min", "max", "skew", "kurtosis")], digits = 3)
```

```{r}
#| label: correlation-heatmap
#| fig-cap: "Spearman correlation matrix of all network variables"
#| fig-width: 10
#| fig-height: 9

# Spearman to match the network estimation input (Isvoranu & Epskamp, 2023)
cor_mat <- cor(df_net, use = "pairwise.complete.obs", method = "spearman")

# Use short labels for readability
colnames(cor_mat) <- short_labels[colnames(cor_mat)]
rownames(cor_mat) <- short_labels[rownames(cor_mat)]

corrplot(cor_mat, method = "color", type = "lower",
         tl.col = "black", tl.cex = 0.9,
         addCoef.col = "black", number.cex = 0.6,
         col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
         diag = FALSE)
```

```{r}
#| label: normality-check
#| fig-cap: "Distribution of all network variables"
#| fig-width: 12
#| fig-height: 10

df_net %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(variable = short_labels[variable]) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", linewidth = 0.2) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  theme_minimal(base_size = 11) +
  labs(x = NULL, y = "Count")
```


# Section 3. Network Estimation (EBICglasso) and Community Detection

```{r}
#| label: estimate-network

# Estimate Gaussian Graphical Model (EBICglasso) on Spearman correlations
# (Isvoranu & Epskamp, 2023). This bootnet object is used only for
# visualization (Section 4) and edge-weight bootstrapping (Section 7).
network <- estimateNetwork(df_net, default = "EBICglasso", tuning = 0.5,
                           corMethod = "spearman", missing = "pairwise")

cat("Network estimated with", ncol(network$graph), "nodes\n")
cat("Non-zero edges:", sum(network$graph[upper.tri(network$graph)] != 0),
    "out of", sum(upper.tri(network$graph)), "possible\n")
cat("Sparsity:", round(1 - mean(network$graph[upper.tri(network$graph)] != 0), 3), "\n")
```

```{r}
#| label: adjacency-matrix
#| tbl-cap: "Weighted adjacency matrix (partial correlations)"

adj_mat <- round(network$graph, 3)
colnames(adj_mat) <- short_labels[colnames(adj_mat)]
rownames(adj_mat) <- short_labels[rownames(adj_mat)]
print(adj_mat)
```

```{r}
#| label: ega

# Exploratory Graph Analysis: the empirically derived community structure.
# Estimated here rather than in Section 6 because every network plot from
# Section 4 onward colours its nodes by community. Sections 6 and 8 report and
# interrogate this same `ega_result` object; nothing is re-estimated.
ega_result <- EGA(df_net, corr = "spearman", model = "glasso",
                  algorithm = "walktrap", na.data = "pairwise",
                  plot.EGA = FALSE)

cat("Number of communities detected:", ega_result$n.dim, "\n")
```

```{r}
#| label: community-aesthetics

# The full-sample community assignment is reused for every network plot,
# including the proficiency-group panels in Section 9, so that a shape means
# the same thing in each figure and the panels stay comparable.
comm_ids <- sort(unique(ega_result$wc))
comm_n <- as.integer(table(factor(ega_result$wc, levels = comm_ids)))
community_shapes <- setNames(
  shape_pool[seq_along(comm_ids)],
  paste0("Community ", comm_ids, " (", comm_n, ")")
)
node_shapes <- unname(community_shapes[match(ega_result$wc[node_order], comm_ids)])

# Node sizes are fixed here (not per-figure) because the overlap relaxation
# below works in the same [-1, 1] coordinate frame that qgraph plots into.
node_vsize <- unname(7.5 * shape_scale[node_shapes])

cat("Construct (colour) x EGA community (shape):\n")
print(table(Construct = factor(node_construct, levels = names(construct_members)),
            Community = ega_result$wc[node_order]))
cat("\nPer-node assignment:\n")
print(data.frame(
  Variable  = unname(net_labels),
  Construct = unname(node_construct),
  Community = unname(ega_result$wc[node_order])
), row.names = FALSE)
```

```{r}
#| label: plot-helpers

# A spring layout places strongly connected nodes almost on top of each other.
# With the full battery, Loss and VarLoss have a partial correlation near .7 and
# land within a node width of each other, so the glyphs collide and the
# community shapes stop being readable. This nudges any overlapping pair apart
# while holding every node
# inside the plotting box, which preserves the spring structure but guarantees
# the silhouettes stay separable. `radius_scale` converts qgraph's `vsize` into
# the half-width of a node in plot coordinates.
relax_overlap <- function(co, vsize, iter = 600, pad = 1.10, radius_scale = 0.019) {
  r <- vsize * radius_scale
  lim <- 1 - r
  n <- nrow(co)
  for (it in seq_len(iter)) {
    moved <- FALSE
    for (i in seq_len(n - 1)) {
      for (j in (i + 1):n) {
        d <- co[j, ] - co[i, ]
        dist <- sqrt(sum(d^2))
        need <- pad * (r[i] + r[j])
        if (dist < need) {
          if (dist < 1e-8) { d <- c(1e-3, 0); dist <- 1e-3 }
          push <- 0.55 * (need - dist) / 2
          u <- d / dist
          co[i, ] <- co[i, ] - u * push
          co[j, ] <- co[j, ] + u * push
          moved <- TRUE
        }
      }
    }
    co[, 1] <- pmax(pmin(co[, 1], lim), -lim)
    co[, 2] <- pmax(pmin(co[, 2], lim), -lim)
    if (!moved) break
  }
  # Centre, then expand by a single factor so the graph fills the frame instead
  # of sitting in a band. One factor for both axes: a per-axis rescale would
  # stretch the layout and misrepresent the distances the spring layout encodes.
  co <- sweep(co, 2, c(mean(range(co[, 1])), mean(range(co[, 2]))))
  headroom <- 1 - max(r)
  scale <- min(headroom / max(abs(co[, 1])), headroom / max(abs(co[, 2])))
  co * max(1, scale)
}

# Spring coordinates for one graph, already de-overlapped
spring_coords <- function(graph, seed = 42) {
  set.seed(seed)
  qg <- qgraph(graph, layout = "spring", vsize = node_vsize, DoNotPlot = TRUE)
  relax_overlap(qg$layout, node_vsize)
}

# Legend in its own panel so it can never overlap the graph. Three blocks:
# shape (the primary reading), colour (secondary), edge style (sign).
draw_network_legend <- function(cex = 0.82, swatches = construct_colors) {
  op <- par(mar = c(0, 0, 0, 0))
  on.exit(par(op))
  plot.new()
  plot.window(xlim = c(0, 1), ylim = c(0, 1))
  legend(0, 0.98, legend = names(community_shapes),
         pch = unname(shape_pch[community_shapes]), pt.bg = "grey92",
         col = "grey10", pt.lwd = 1.6, pt.cex = 1.8, cex = cex, bty = "n",
         title = "EGA community (shape)", title.adj = 0, y.intersp = 1.3)
  legend(0, 0.70, legend = names(construct_colors), pch = 22,
         pt.bg = unname(swatches), col = "grey45", pt.cex = 1.8,
         cex = cex, bty = "n", title = "Construct (colour)",
         title.adj = 0, y.intersp = 1.3)
  legend(0, 0.34, legend = c("positive", "negative"), lty = c(1, 2),
         lwd = 2.4, col = "grey25", cex = cex, bty = "n",
         title = "Partial correlation", title.adj = 0, y.intersp = 1.3)
}

# Single entry point for every network figure, so the encoding cannot drift.
#
# `minimum` hides edges below |.05| and `cut` gives full width and saturation to
# edges above |.10|. The network is dense (105 possible edges on the full
# battery, half of them below |.03|) and on the full battery one edge,
# Loss-VarLoss at .70, otherwise compresses everything else into invisible
# hairlines. The complete weighted adjacency matrix is printed above, so
# nothing is lost, only decluttered.
# `negDashed` makes edge sign readable in black and white.
plot_network <- function(graph, coords, title = "", legend = TRUE,
                         minimum = 0.05, cut = 0.10, color = node_colors,
                         posCol = "#2166AC", negCol = "#B2182B",
                         swatches = construct_colors, ...) {
  if (legend) graphics::layout(matrix(1:2, nrow = 1), widths = c(1, 0.32))
  qg <- qgraph(graph,
               layout = coords,
               rescale = FALSE,
               labels = net_labels,
               color = color,
               shape = node_shapes,
               polygonList = shape_polygons,
               vsize = node_vsize,
               borders = TRUE,
               border.width = 2.6,
               border.color = "grey10",
               label.scale.equal = TRUE,
               label.color = "grey5",
               negDashed = TRUE,
               minimum = minimum,
               cut = cut,
               esize = 10,
               legend = FALSE,
               posCol = posCol,
               negCol = negCol,
               title = title,
               title.cex = 1.2,
               # qgraph's `mar` pads the [-1, 1] box in tenths; keep it tight so
               # the graph fills the panel, with extra headroom for the title.
               mar = c(2, 2, 4, 2),
               ...)
  if (legend) {
    draw_network_legend(swatches = swatches)
    graphics::layout(1)
  }
  invisible(qg)
}
```


# Section 4. Network Visualization

Node **shape** encodes the empirically derived EGA community. That is the primary
reading, and the one channel that survives black-and-white printing. Node
**colour** is the a priori theoretical construct, deliberately desaturated so it
sits behind the shapes. Nodes whose colour disagrees with the shapes around them
are the cases where the data depart from the a priori grouping. Solid edges are
positive partial correlations, dashed edges negative.

```{r}
#| label: network-spring
#| fig-cap: "GGM network (spring layout, Fruchterman-Reingold). Shape = EGA community, colour = theoretical construct; edges below |.05| omitted."
#| fig-width: 13
#| fig-height: 8.5

spring_layout <- spring_coords(network$graph)

plot_network(network$graph, spring_layout,
             title = "ELLIPSE Linguistic Features (GGM Network)")
```

```{r}
#| label: save-spring-plot
#| include: false

pdf(fig_path("network_spring.pdf"), width = 13, height = 8.5)
plot_network(network$graph, spring_layout,
             title = "ELLIPSE Linguistic Features (GGM Network)")
dev.off()
cat("Saved to", fig_path("network_spring.pdf"), "\n")
```

```{r}
#| label: network-circle
#| fig-cap: "GGM network (circle layout). Shape = EGA community, colour = theoretical construct; edges below |.05| omitted."
#| fig-width: 12
#| fig-height: 10

circle_layout <- local({
  ang <- seq(pi / 2, pi / 2 - 2 * pi, length.out = length(node_order) + 1)[-1]
  cbind(0.82 * cos(ang), 0.82 * sin(ang))
})

plot_network(network$graph, circle_layout,
             title = "ELLIPSE Linguistic Features (GGM Network, Circle Layout)")
```

```{r}
#| label: network-spring-greyscale
#| fig-cap: "Greyscale proof of the spring-layout network: community (shape) and edge sign (dash) remain readable without colour."
#| fig-width: 13
#| fig-height: 8.5

# Print check. Constructs collapse onto the grey ramp reported in Section 1;
# communities and edge sign are unaffected because neither is carried by hue.
grey_construct <- sapply(construct_colors, function(x) {
  rgb <- col2rgb(x) / 255
  grey(0.2126 * rgb[1] + 0.7152 * rgb[2] + 0.0722 * rgb[3])
})

plot_network(network$graph, spring_layout,
             title = "GGM Network, Greyscale Proof",
             color = unname(grey_construct[node_construct]),
             swatches = grey_construct,
             posCol = "grey20", negCol = "grey20")
```


# Section 5. Centrality Analysis

```{r}
#| label: centrality-plot
#| fig-cap: "Centrality indices for all nodes"
#| fig-width: 10
#| fig-height: 8

# Strength (primary) + signed ExpectedInfluence. Network loadings
# (Section 6) are the confound-adjusted node-importance metric and supersede
# raw strength for interpretation.
centralityPlot(network$graph, labels = net_labels,
               include = c("Strength", "ExpectedInfluence"),
               orderBy = "Strength")
```

```{r}
#| label: centrality-table
#| tbl-cap: "Centrality values for all nodes"

cent <- centrality_auto(network$graph)$node.centrality
cent$Variable <- short_labels[rownames(cent)]
# Keep Strength + ExpectedInfluence only (drop Betweenness/Closeness)
cent <- cent %>% select(Variable, Strength, any_of("ExpectedInfluence"))
print(cent %>% arrange(desc(Strength)), digits = 3)
```

```{r}
#| label: predictability-edges
#| tbl-cap: "Edge weights involving predictability nodes"

# One column per retained predictability node, ordered by the focal node's
# edge magnitude so the columns stay comparable across runs.
adj <- network$graph
partners <- setdiff(colnames(adj), pred_vars)

pred_edges <- data.frame(Partner = unname(short_labels[partners]))
for (v in pred_vars) {
  pred_edges[[unname(short_labels[[v]])]] <- round(adj[v, partners], 3)
}
pred_edges <- pred_edges[order(-abs(adj[focal, partners])), ]

cat("Strongest connections of the predictability node(s):\n")
print(pred_edges, row.names = FALSE)
```


# Section 6. Community Structure (EGA)

`ega_result` is estimated in Section 3 (its communities drive the node colours
throughout); this section reports and interrogates it.

```{r}
#| label: ega-assignments

cat("Number of communities detected:", ega_result$n.dim, "\n")
cat("\nCommunity assignments:\n")

# Show assignments with short labels
assignments <- data.frame(
  Variable = short_labels[names(ega_result$wc)],
  Full_name = names(ega_result$wc),
  Community = ega_result$wc
)
print(assignments %>% arrange(Community))
```

```{r}
#| label: ega-predictability

# Where each retained predictability node ended up. This is the whole point of
# the sensitivity runs, so it is pulled out rather than left buried in the full
# assignment table above.
for (v in pred_vars) {
  comm <- unname(ega_result$wc[v])
  members <- names(ega_result$wc)[ega_result$wc == comm]
  cat(unname(short_labels[[v]]), "-> Community", comm, "\n")
  cat("  Members:", paste(unname(short_labels[members]), collapse = ", "), "\n")
  cat("  With the dependency-MI indices:",
      all(c("amod_mi", "dobj_mi", "advmod_mi") %in% members), "\n\n")
}
```

```{r}
#| label: net-loadings
#| tbl-cap: "Revised network loadings (standardized)"

# Network loadings (Christensen, Golino, Abad & Garrido, 2025): confound-adjusted
# node importance that splits strength by community and adjusts for community
# size and sign.
# Effect-size guides: ~0.20 small, ~0.35 moderate, ~0.50 large.

nl <- net.loads(ega_result, loading.method = "revised")$std

nl_df <- as.data.frame(round(nl, 3))
nl_df$Variable <- short_labels[rownames(nl)]
nl_df <- nl_df %>% select(Variable, everything())
cat("Standardized network loadings (nodes x communities):\n")
print(nl_df, row.names = FALSE)

cat("\nPredictability network loadings:\n")
print(nl_df %>% filter(Variable %in% short_labels[pred_vars]), row.names = FALSE)
```

```{r}
#| label: boot-ega
#| fig-cap: "Bootstrap EGA: stability of community assignments"

# NOT REPRODUCIBLE ACROSS RENDERS, despite the seed. bootEGA and the two
# bootnet calls below run their resampling in parallel worker processes, and
# neither seeds those workers from the master's RNG state, so set.seed() here
# fixes nothing that happens inside them. Four renders of identical code during
# development gave structural consistency for the 4th (cohesion) community of
# .690, .674, .644 and .658; one of those renders differed from the previous
# only in markdown prose. Everything the notebook derives from the point
# estimates (communities, loadings, edge weights, centrality, the group
# comparisons) IS stable and reproduces exactly.
#
# So quote the stability figures as approximate, and do not read a change of a
# few points between runs as a finding. Making them reproducible would mean
# forcing single-core execution or seeding the workers with
# parallel::clusterSetRNGStream(), neither of which is wired up here.
set.seed(42)
boot_ega <- bootEGA(df_net, iter = 500,
                    corr = "spearman",
                    model = "glasso",
                    algorithm = "walktrap",
                    na.data = "pairwise",
                    type = "parametric",
                    verbose = FALSE)

cat("Modal number of communities across bootstraps:",
    as.numeric(names(sort(table(boot_ega$n.dim), decreasing = TRUE))[1]), "\n")

# boot_ega$frequency is the distribution of the NUMBER of dimensions across
# bootstraps -- not a per-item quantity. Item and dimension stability come from
# dimensionStability(), which is what actually answers "does this node keep
# landing in the same community?".
cat("\nDistribution of the number of communities across bootstraps:\n")
print(round(boot_ega$frequency, 3))

dim_stab <- dimensionStability(boot_ega, verbose = FALSE)

cat("\nStructural consistency (proportion of bootstraps in which each",
    "\nempirical community is exactly recovered):\n")
print(round(dim_stab$dimension.stability$structural.consistency, 3))

cat("\nItem stability (proportion of bootstraps in which each node is",
    "\nassigned to its empirical community):\n")
item_stab <- dim_stab$item.stability$item.stability$empirical.dimensions
names(item_stab) <- short_labels[names(item_stab)]
print(round(sort(item_stab), 3))
```


# Section 7. Network Stability (bootnet)

```{r}
#| label: boot-edges
#| fig-cap: "Bootstrap confidence intervals for edge weights"
#| fig-width: 12
#| fig-height: 10

set.seed(42)
boot_nonparam <- bootnet(network, nBoots = 1000, type = "nonparametric",
                         nCores = n_cores,
                         verbose = FALSE)

plot(boot_nonparam, labels = net_labels,
     order = "sample", CIstyle = "quantiles")
```

```{r}
#| label: save-edge-boot
#| include: false

pdf(fig_path("network_edge_bootstrap.pdf"), width = 12, height = 10)
plot(boot_nonparam, labels = net_labels,
     order = "sample", CIstyle = "quantiles")
dev.off()
cat("Saved to", fig_path("network_edge_bootstrap.pdf"), "\n")
```

```{r}
#| label: boot-centrality
#| fig-cap: "Case-dropping bootstrap for centrality stability"
#| fig-width: 10
#| fig-height: 8

set.seed(42)
boot_case <- bootnet(network, nBoots = 1000, type = "case",
                     statistics = c("strength", "expectedInfluence"),
                     nCores = n_cores,
                     verbose = FALSE)

plot(boot_case, statistics = c("strength", "expectedInfluence"))
```

```{r}
#| label: cs-coefficients
#| tbl-cap: "CS-coefficients (correlation stability)"

cs_coefs <- corStability(boot_case)
print(cs_coefs)
```

**Interpretation guidelines** (Epskamp et al., 2018)

- CS ≥ 0.50 means centrality is stable and interpretable
- CS ≥ 0.25 is the acceptable minimum
- CS < 0.25 means centrality should **not** be interpreted


# Section 8. Between-Community Loadings (Bridge Behavior)

Bridge behavior is operationalized via **between-community (cross-)loadings**
from the revised network loadings (Christensen et al., 2025). Cross-loadings
distinguish a genuine bridge from between-community correlation.

```{r}
#| label: cross-loadings
#| tbl-cap: "Between-community (cross-)loadings"

# `nl` (standardized network loadings, nodes x communities) is computed in
# Section 6. What matters is a node's loading on its OWN community against its
# loadings on the others. High and roughly-equal cross-loadings mean
# integrative bridge behavior, a single dominant loading means the node
# belongs to one community.
# nl's columns are named by community id ("1","2",...), matching ega_result$wc.

assigned <- ega_result$wc  # community per node

cross_load_summary <- function(node) {
  row <- nl[node, ]
  own <- as.character(assigned[node])
  others <- setdiff(names(row), own)
  data.frame(
    Variable = short_labels[node],
    Own_community = own,
    Own_loading = round(row[[own]], 3),
    Max_cross = round(max(row[others]), 3),
    Mean_abs_cross = round(mean(abs(row[others])), 3)
  )
}

cat("Standardized loadings (nodes x communities):\n")
print(round(nl, 3))

cat("\nPredictability nodes, in-community vs cross-community loading:\n")
print(do.call(rbind, lapply(pred_vars, cross_load_summary)), row.names = FALSE)
```

# Section 9. Network Comparison by Proficiency Level

```{r}
#| label: proficiency-split

# Use single Overall score from public ELLIPSE data.
df_grp <- df_raw
prompt_grp <- df_grp$prompt

# Three proficiency groups in (approximately) equal bins. `Overall` is discrete,
# with 9 levels in 0.5 steps and heavy ties, so exact equal-n tertiles would
# split essays with identical scores across groups. Instead the two cutpoints
# are placed at the observed score levels closest to the 1/3 and 2/3 quantiles,
# which keeps every tied score intact and still yields near-equal bins.
score_counts <- table(df_grp$Overall)
score_levels <- as.numeric(names(score_counts))
cum_prop <- cumsum(score_counts) / nrow(df_grp)
cut_low  <- score_levels[which.min(abs(cum_prop - 1 / 3))]
cut_high <- score_levels[which.min(abs(cum_prop - 2 / 3))]

df_grp$prof_group <- factor(
  ifelse(df_grp$Overall <= cut_low, "Low",
         ifelse(df_grp$Overall <= cut_high, "Mid", "High")),
  levels = c("Low", "Mid", "High")
)
prof_levels <- levels(df_grp$prof_group)

cat("Cutpoints on Overall: Low <=", cut_low, "| Mid <=", cut_high, "| High >", cut_high, "\n\n")
cat("Proficiency group sizes and score ranges:\n")
print(data.frame(
  Group = prof_levels,
  n = as.integer(table(df_grp$prof_group)),
  Min = tapply(df_grp$Overall, df_grp$prof_group, min),
  Max = tapply(df_grp$Overall, df_grp$prof_group, max),
  Pct = round(100 * as.numeric(table(df_grp$prof_group)) / nrow(df_grp), 1)
), row.names = FALSE)

# Same variable set as the full-sample network
keep_vars <- node_order

# Residualize on prompt within each group, exactly as for the full sample
residualize_group <- function(g) {
  idx <- df_grp$prof_group == g
  d <- df_grp[idx, keep_vars]
  pr <- prompt_grp[idx]
  for (var in names(d)) {
    d[[var]] <- residuals(lm(d[[var]] ~ factor(pr), na.action = na.exclude))
  }
  d
}
df_prof <- lapply(setNames(prof_levels, prof_levels), residualize_group)
```

```{r}
#| label: estimate-group-networks

nets_prof <- lapply(df_prof, function(d) {
  estimateNetwork(d, default = "EBICglasso", tuning = 0.5,
                  corMethod = "spearman", missing = "pairwise")
})

cat("Non-zero edges per group (of", sum(upper.tri(nets_prof$Low$graph)), "possible):\n")
print(sapply(nets_prof, function(n) sum(n$graph[upper.tri(n$graph)] != 0)))
```

```{r}
#| label: group-network-plots
#| fig-cap: "Network comparison across three equal-bin proficiency groups. Shape = full-sample EGA community, colour = theoretical construct; layout, edge scaling and the |.05| display threshold are shared across panels."
#| fig-width: 18
#| fig-height: 6

# One layout and one edge-width scale for all three panels, so differences in
# the pictures are differences in the networks.
avg_layout <- relax_overlap(
  do.call(averageLayout, lapply(nets_prof, function(n) n$graph)),
  node_vsize
)
prof_max <- max(abs(unlist(lapply(nets_prof, function(n) n$graph))))

plot_proficiency_panels <- function() {
  graphics::layout(matrix(1:4, nrow = 1), widths = c(1, 1, 1, 0.42))
  for (g in prof_levels) {
    plot_network(nets_prof[[g]]$graph, avg_layout,
                 title = paste0(g, " proficiency (n = ", nrow(df_prof[[g]]), ")"),
                 legend = FALSE,
                 maximum = prof_max)
  }
  draw_network_legend()
  graphics::layout(1)
}

plot_proficiency_panels()
```

```{r}
#| label: save-group-plots
#| include: false

pdf(fig_path("network_group_comparison.pdf"), width = 18, height = 6)
plot_proficiency_panels()
dev.off()
cat("Saved to", fig_path("network_group_comparison.pdf"), "\n")
```

```{r}
#| label: network-compare
#| output: false

# EGAnet permutation network comparison, run for all three pairs of groups.
# Feed the residualized group DATA FRAMES.
# nc$network is a matrix with columns sF (signed Frobenius),
# JSS (Jensen-Shannon similarity), and Total Strength.
# In each comparison the statistic is oriented as `comparison - base`.
prof_pairs <- list(
  c("Low", "Mid"),
  c("Mid", "High"),
  c("Low", "High")
)
names(prof_pairs) <- sapply(prof_pairs, paste, collapse = " vs ")

set.seed(42)
nc_list <- lapply(prof_pairs, function(pair) {
  network.compare(
    base = df_prof[[pair[1]]], comparison = df_prof[[pair[2]]],
    corr = "spearman", model = "glasso",
    na.data = "pairwise",
    iter = 1000, seed = 42
  )
})

cat("=== Permutation Network Comparison (network.compare) ===\n")
for (nm in names(nc_list)) {
  cat("\n---", nm, "(statistic oriented as second - first) ---\n")
  print(nc_list[[nm]]$network)   # cols: sF (signed Frobenius), JSS, Total Strength
}
```

```{r}
#| label: network-compare-edges
#| output: false

# nc$edges is a list of node-by-node matrices: statistic (comparison - base), p.value,
# p.adjusted (BH), M_permutated, SD_permutated. Report edges whose BH-adjusted
# permutation p < 0.05 (upper triangle only), for each pair of groups.
for (nm in names(nc_list)) {
  padj <- nc_list[[nm]]$edges$p.adjusted
  stat <- nc_list[[nm]]$edges$statistic
  ut <- which(upper.tri(padj) & padj < 0.05, arr.ind = TRUE)
  cat("\n---", nm, "per-edge differences (BH-adjusted p < 0.05) ---\n")
  if (nrow(ut) > 0) {
    sig_edges <- data.frame(
      Node1 = short_labels[rownames(padj)[ut[, "row"]]],
      Node2 = short_labels[colnames(padj)[ut[, "col"]]],
      Statistic = round(stat[ut], 3),
      p_adjusted = round(padj[ut], 3)
    ) %>% arrange(p_adjusted)
    print(sig_edges, row.names = FALSE)
  } else {
    cat("No individual edges differ after BH correction.\n")
  }
}
```

```{r}
#| label: invariance
#| eval: true

# loadings-based permutation invariance test (Jamison, Christensen & Golino, 2024)
# Alternative/supplement to network.compare above if RQ4 is framed as invariance.
# invariance() takes all three groups at once and reports every pairwise contrast.

inv <- invariance(
  data = do.call(rbind, df_prof),
  groups = rep(prof_levels, sapply(df_prof, nrow)),
  corr = "spearman", na.data = "pairwise", iter = 1000, seed = 42
)
print(inv$results)   # per-node loading differences + permutation p-values
```

```{r}
#| label: compare-predictability

# Predictability-focused edge comparison: partial correlation in each of the
# three groups (Spearman group networks) plus the Low-vs-High permutation
# difference test from network.compare (statistic = High - Low, p.value, BH
# p.adjusted). All matrices indexed by node name to keep columns aligned.
nc_lh <- nc_list[["Low vs High"]]

pred_edge_table <- function(node) {
  partners <- colnames(nets_prof$Low$graph)
  data.frame(
    Partner = short_labels[partners],
    Low = round(nets_prof$Low$graph[node, partners], 3),
    Mid = round(nets_prof$Mid$graph[node, partners], 3),
    High = round(nets_prof$High$graph[node, partners], 3),
    Diff_HL = round(nc_lh$edges$statistic[node, partners], 3),
    p_value = round(nc_lh$edges$p.value[node, partners], 3),
    p_adjusted = round(nc_lh$edges$p.adjusted[node, partners], 3)
  ) %>%
    filter(!Partner %in% short_labels[pred_vars]) %>%
    arrange(desc(abs(Diff_HL)))
}

for (v in pred_vars) {
  cat(v, "edge weight by proficiency group + Low-vs-High test:\n")
  print(pred_edge_table(v), row.names = FALSE)
  cat("\n")
}
```


# Summary and Interpretation

```{r}
#| label: summary-values
#| echo: false

# Derived quantities for the narrative below (names stripped so inline R prints
# clean scalars).
focal_label    <- unname(short_labels[[focal]])
pred_comm      <- unname(ega_result$wc[focal])
same_comm      <- names(ega_result$wc)[ega_result$wc == pred_comm]
loss_own       <- as.character(ega_result$wc[focal])
loss_own_load  <- round(nl[focal, loss_own], 3)
loss_max_cross <- round(max(nl[focal, setdiff(colnames(nl), loss_own)]), 3)
cent_sorted    <- cent %>% arrange(desc(Strength))
loss_rank      <- which(cent_sorted$Variable == focal_label)
n_nodes        <- nrow(cent_sorted)
mi_same_comm   <- unname(pred_comm == ega_result$wc["amod_mi"])
```

## 1. Community structure (EGA)

EGA detected `r ega_result$n.dim` communities.

```{r}
#| label: summary-communities
#| echo: false
#| output: asis

if (!is.null(drop_var)) {
  cat("This run holds `", drop_var, "` out of the battery.\n\n", sep = "")
}
for (v in pred_vars) {
  cm <- unname(ega_result$wc[v])
  mem <- setdiff(names(ega_result$wc)[ega_result$wc == cm], v)
  cat(sprintf("- `%s` is in community %s, alongside %s\n", v, cm,
              paste(unname(short_labels[mem]), collapse = ", ")))
}
```

## 2. Node importance

`r focal` loads `r loss_own_load` on its own community, with a maximum
cross-community loading of `r loss_max_cross`. As a secondary check, its raw
strength rank is `r loss_rank` of `r n_nodes`.

## 3. Bridge behavior (cross-loadings)

Bridge behavior is read from between-community cross-loadings (Section 8). High
and roughly-equal cross-loadings indicate an integrative bridge node, whereas a
single dominant loading means the node belongs to one community.

## 4. Interpretation relative to CFA hypotheses

| CFA Model | Supporting Evidence |
|-----------|---------------------|
| Model 1 (Phraseological) | Predictability shares the MI community: `r mi_same_comm` |
| Model 2 (General factor)  | `r focal` strength rank = `r loss_rank` of `r n_nodes` |
| Novel finding             | See community membership in Section 6 |

## 5. Stability

CS-coefficients and edge-accuracy CIs are reported in Section 7.

## 6. Proficiency group differences (network.compare)

Three equal-bin proficiency groups (Low / Mid / High), all three pairwise
comparisons. Omnibus structural comparison: signed Frobenius (sF),
Jensen-Shannon similarity (JSS), and total strength with permutation
*p*-values. The statistic is oriented as (second group − first group).

```{r}
#| label: summary-network-compare
#| echo: false
nc_summary <- do.call(rbind, lapply(names(nc_list), function(nm) {
  m <- round(nc_list[[nm]]$network, 3)
  data.frame(Comparison = nm, Quantity = rownames(m), m, row.names = NULL,
             check.names = FALSE)
}))
knitr::kable(nc_summary)
```
