---
title: "Commonality Analysis of Word Predictability, Phraseology and Grammar"
subtitle: "Is surprisal more phraseological than grammatical?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

The network analysis places `mean_loss` in a community with the dependency-MI
features. This notebook asks a related but **distinct** question, of the human
ratings rather than the computed indices: **of the variance in word
predictability that the ELLIPSE ratings explain, how much is uniquely
phraseological, how much is uniquely grammatical, and how much is shared?**

The distinction matters for reading the result — a rater's holistic
`Phraseology` judgment and a corpus-derived MI index are different
operationalizations, so this notebook does not adjudicate the network finding.
See the Interpretation section.

Commonality analysis (Newton & Spurrell, 1967; Nimon & Oswald, 2013) partitions
$R^2$ into unique and common components. For two predictors:

$$U_{\text{phr}} = R^2_{pg} - R^2_{g} \qquad
  U_{\text{gram}} = R^2_{pg} - R^2_{p} \qquad
  C = R^2_{p} + R^2_{g} - R^2_{pg}$$

The framing matters here because `Phraseology` and `Grammar` are rated by the
same humans on the same essays and carry a large halo: most of what they
explain, they explain jointly. Partial correlations would report that shared
part as though it had vanished. Commonality reports it as a quantity, which is
the more honest description of a rating set like this one.

**Sign convention.** All analyses use `neg_loss = -mean_loss`, so that higher
values mean *more predictable* text and every coefficient reads in the same
direction as the proficiency scores.

```{r}
#| label: setup
library(tidyverse)
library(boot)

set.seed(42)
```

```{r}
#| label: load-data

d_raw <- read_csv("../../data/ellipse/ellipse_metrics.csv",
                  show_col_types = FALSE)

d <- d_raw |>
  transmute(
    text_id_kaggle,
    prompt,
    neg_loss    = -mean_loss,
    cohesion    = Cohesion,
    syntax      = Syntax,
    vocabulary  = Vocabulary,
    phraseology = Phraseology,
    grammar     = Grammar,
    conventions = Conventions,
    n_tokens    = n_tokens
  )

# The six analytic criteria, in the order ELLIPSE reports them.
crit <- c("cohesion", "syntax", "vocabulary",
          "phraseology", "grammar", "conventions")

cat("Rows loaded:", nrow(d), "\n")
na_counts <- colSums(is.na(d))
if (any(na_counts > 0)) {
  cat("Missing values:\n"); print(na_counts[na_counts > 0])
}

d <- drop_na(d)
cat("Complete cases used:", nrow(d), "\n\n")

cat("Rating scales (ELLIPSE ratings are ordinal, 1-5 in 0.5 steps):\n")
print(map_dfr(d[crit],
              \(x) tibble(min = min(x), max = max(x),
                          levels = n_distinct(x), sd = sd(x)),
              .id = "variable"))
```

## Rank transformation

Everything below is computed on **ranks**. With only ~9 distinct rating levels
and a bounded scale, Pearson correlations on the raw scores are not the right
summary; ranking first makes every simple $R^2$ equal to a squared Spearman
correlation and keeps the whole decomposition on one consistent footing.

Ranks are computed **once**, on the full sample, and the bootstrap then
resamples rows of the ranked data. This treats the ranks as the data rather than
re-ranking inside each resample. Re-ranking within resamples is the stricter
analogue of a Spearman bootstrap, but with *n* > 6,000 the two are
indistinguishable and re-ranking is far slower.

```{r}
#| label: rank-transform

vars <- c("neg_loss", crit, "n_tokens")
dr <- d |> mutate(across(all_of(vars), \(x) rank(x)))
```

## Zero-order correlations

Reported first for context, and because the phraseology–grammar correlation *is*
the halo estimate — it sets an upper bound on how much unique variance either
rating can possibly have.

```{r}
#| label: zero-order

zo <- tribble(
  ~pair,                      ~rho,
  "neg_loss ~ phraseology",   cor(dr$neg_loss, dr$phraseology),
  "neg_loss ~ grammar",       cor(dr$neg_loss, dr$grammar),
  "neg_loss ~ vocabulary",    cor(dr$neg_loss, dr$vocabulary),
  "phraseology ~ grammar",    cor(dr$phraseology, dr$grammar),
  "phraseology ~ vocabulary", cor(dr$phraseology, dr$vocabulary),
  "grammar ~ vocabulary",     cor(dr$grammar, dr$vocabulary)
) |> mutate(rho = round(rho, 3), r_squared = round(rho^2, 3))

print(as.data.frame(zo), row.names = FALSE)
```

# Two-predictor decomposition

```{r}
#| label: r2-helper

# R^2 of a multiple regression, straight from the correlation matrix:
#   R^2 = r_yx' Rxx^-1 r_yx
# Algebraically identical to summary(lm(...))$r.squared on standardized data,
# but it lets the bootstrap compute every sub-model from ONE cor() call per
# resample instead of fitting a fresh lm for each. Validated against lm below.
r2_from_cor <- function(R, y, xs) {
  rxy <- R[xs, y, drop = FALSE]
  Rxx <- R[xs, xs, drop = FALSE]
  as.numeric(crossprod(rxy, solve(Rxx, rxy)))
}
```

```{r}
#| label: validate-helper

R_full <- cor(dr[vars])
check <- tibble(
  model = c("phr", "gram", "phr + gram"),
  via_cor = c(
    r2_from_cor(R_full, "neg_loss", "phraseology"),
    r2_from_cor(R_full, "neg_loss", "grammar"),
    r2_from_cor(R_full, "neg_loss", c("phraseology", "grammar"))
  ),
  via_lm = c(
    summary(lm(neg_loss ~ phraseology, dr))$r.squared,
    summary(lm(neg_loss ~ grammar, dr))$r.squared,
    summary(lm(neg_loss ~ phraseology + grammar, dr))$r.squared
  )
) |> mutate(abs_diff = abs(via_cor - via_lm))

print(as.data.frame(check), row.names = FALSE)
stopifnot(all(check$abs_diff < 1e-10))
cat("\nHelper matches lm() to within 1e-10.\n")
```

```{r}
#| label: common2

common2 <- function(dd, i) {
  R <- cor(dd[i, c("neg_loss", "phraseology", "grammar")])
  full <- r2_from_cor(R, "neg_loss", c("phraseology", "grammar"))
  p    <- r2_from_cor(R, "neg_loss", "phraseology")
  g    <- r2_from_cor(R, "neg_loss", "grammar")
  c(U_phr  = full - g,
    U_gram = full - p,
    Common = p + g - full,
    R2     = full,
    Diff   = p - g)          # U_phr - U_gram, algebraically
}

est2 <- common2(dr, seq_len(nrow(dr)))
print(round(est2, 5))
```

Note that $U_{\text{phr}} - U_{\text{gram}} = (R^2_{pg} - R^2_g) - (R^2_{pg} -
R^2_p) = R^2_p - R^2_g$. The comparison of unique contributions reduces exactly
to the comparison of the two zero-order $R^2$ values, which is worth knowing
before over-interpreting it: it is the *same* question as "which rating
correlates more strongly with predictability", asked in variance units.

```{r}
#| label: common2-boot

b2 <- boot(dr, common2, R = 5000)

ci_table <- function(b, est) {
  map_dfr(seq_along(est), \(k) {
    ci <- boot.ci(b, type = "perc", index = k)$percent[4:5]
    tibble(component = names(est)[k],
           estimate  = est[k],
           lower     = ci[1],
           upper     = ci[2],
           boot_se   = sd(b$t[, k]))
  })
}

res2 <- ci_table(b2, est2) |>
  mutate(across(c(estimate, lower, upper, boot_se), \(x) round(x, 5)))
print(as.data.frame(res2), row.names = FALSE)
```

```{r}
#| label: common2-proportions

prop <- res2 |>
  filter(component %in% c("U_phr", "U_gram", "Common")) |>
  mutate(pct_of_R2 = round(100 * estimate / est2[["R2"]], 1)) |>
  dplyr::select(component, estimate, pct_of_R2)

cat("Share of the explained variance (R2 =", round(est2[["R2"]], 4), "):\n")
print(as.data.frame(prop), row.names = FALSE)
```

## The focal test

`Diff` is the direct answer to "is predictability more phraseological than
grammatical?". Its bootstrap percentile interval replaces the analytic test for
comparing dependent correlations: no normal-theory assumption, and it inherits
the rank transformation without further argument.

```{r}
#| label: focal-test

diff_ci <- res2 |> filter(component == "Diff")
cat("U_phr - U_gram =", diff_ci$estimate,
    "  95% CI [", diff_ci$lower, ",", diff_ci$upper, "]\n\n")

p_boot <- mean(b2$t[, 5] <= 0)
cat("Proportion of bootstrap replicates with Diff <= 0:", round(p_boot, 4), "\n")
cat(if (diff_ci$lower > 0) {
  "CI excludes zero: phraseology contributes more unique variance.\n"
} else if (diff_ci$upper < 0) {
  "CI excludes zero: grammar contributes more unique variance.\n"
} else {
  "CI includes zero: no reliable difference in unique contribution.\n"
})
```

# Three-predictor decomposition

Adding `vocabulary` gives all seven commonality components. The pairwise and
triple terms have closed forms, so nothing here needs to be assembled by hand:

$$C_{pg} = R^2_{pv} + R^2_{gv} - R^2_{v} - R^2_{pgv}$$
$$C_{pgv} = R^2_{p} + R^2_{g} + R^2_{v} - R^2_{pg} - R^2_{pv} - R^2_{gv} + R^2_{pgv}$$

The seven components sum to $R^2_{pgv}$ by construction, which the code asserts.

```{r}
#| label: common3

common3 <- function(dd, i) {
  R <- cor(dd[i, c("neg_loss", "phraseology", "grammar", "vocabulary")])
  r2 <- \(xs) r2_from_cor(R, "neg_loss", xs)

  p <- r2("phraseology"); g <- r2("grammar"); v <- r2("vocabulary")
  pg <- r2(c("phraseology", "grammar"))
  pv <- r2(c("phraseology", "vocabulary"))
  gv <- r2(c("grammar", "vocabulary"))
  pgv <- r2(c("phraseology", "grammar", "vocabulary"))

  c(U_phr   = pgv - gv,
    U_gram  = pgv - pv,
    U_vocab = pgv - pg,
    C_pg    = pv + gv - v - pgv,
    C_pv    = pg + gv - g - pgv,
    C_gv    = pg + pv - p - pgv,
    C_pgv   = p + g + v - pg - pv - gv + pgv,
    R2      = pgv)
}

est3 <- common3(dr, seq_len(nrow(dr)))
print(round(est3, 5))

# The seven components must sum to the full model R^2.
stopifnot(abs(sum(est3[1:7]) - est3[["R2"]]) < 1e-10)
cat("\nComponents sum to R2 (difference <", 1e-10, ").\n")
```

```{r}
#| label: common3-boot

b3 <- boot(dr, common3, R = 5000)

res3 <- ci_table(b3, est3) |>
  mutate(pct_of_R2 = round(100 * estimate / est3[["R2"]], 1),
         across(c(estimate, lower, upper, boot_se), \(x) round(x, 5)))
print(as.data.frame(res3), row.names = FALSE)
```

Commonality components can legitimately be **negative**, which indicates
suppression rather than an error. Any negative component above should be read
that way, not as a computational problem.

# Relative importance across all six criteria

Commonality analysis exposes the problem but does not solve it: with ~70% of the
explained variance shared, the *unique* components understate every predictor,
and they understate the correlated ones most. Ranking criteria by unique
variance therefore rewards whichever rating happens to be least redundant with
the others, which is not the same as importance.

**LMG** (Lindeman, Merenda & Gold, 1980), the Shapley-value decomposition,
answers the question the unique components cannot. It averages each predictor's
sequential $R^2$ contribution over **all $6! = 720$ orderings** of entry, so
shared variance is *allocated* among the predictors that share it rather than
discarded. The weights are non-negative and sum exactly to $R^2$, which makes
them readable as "this criterion's share of the explained variance."

Both metrics are reported below, because the contrast between them is
informative in itself:

- `lmg` — Shapley share, shared variance allocated.
- `last` — the variance a criterion adds when entered last. This is exactly the
  squared semipartial, and exactly the *unique* commonality component from the
  sections above, generalized to six predictors.

A criterion with a large `lmg` but a near-zero `last` contributes almost
entirely through variance it shares with the other ratings.

```{r}
#| label: relimp-setup
library(relaimpo)

m_all <- lm(reformulate(crit, "neg_loss"), data = dr)
cat("Full model R2 =", round(summary(m_all)$r.squared, 5), "\n")
```

```{r}
#| label: relimp-coherence

# `last` from relaimpo must reproduce the unique commonality components
# computed by hand earlier. Checked on the two-predictor model so the two
# sections are demonstrably the same decomposition, not merely similar.
m2 <- lm(neg_loss ~ phraseology + grammar, data = dr)
last2 <- calc.relimp(m2, type = "last", rela = FALSE)$last

coh <- tibble(
  component   = c("U_phr", "U_gram"),
  commonality = c(est2[["U_phr"]], est2[["U_gram"]]),
  relaimpo_last = c(last2[["phraseology"]], last2[["grammar"]])
) |> mutate(abs_diff = abs(commonality - relaimpo_last))

print(as.data.frame(coh), row.names = FALSE)
stopifnot(all(coh$abs_diff < 1e-10))
cat("\nrelaimpo's `last` reproduces the commonality unique components exactly.\n")
```

```{r}
#| label: relimp-point

ri_raw  <- calc.relimp(m_all, type = c("lmg", "last"), rela = FALSE)
ri_rela <- calc.relimp(m_all, type = c("lmg", "last"), rela = TRUE)

ri_tab <- tibble(
  criterion = crit,
  lmg       = as.numeric(ri_raw$lmg[crit]),
  lmg_pct   = 100 * as.numeric(ri_rela$lmg[crit]),
  last      = as.numeric(ri_raw$last[crit]),
  last_pct  = 100 * as.numeric(ri_rela$last[crit])
) |>
  mutate(shared_share = round(100 * (1 - last / lmg), 1)) |>
  arrange(desc(lmg)) |>
  mutate(across(c(lmg, last), \(x) round(x, 5)),
         across(c(lmg_pct, last_pct), \(x) round(x, 1)))

print(as.data.frame(ri_tab), row.names = FALSE)
cat("\nlmg sums to", round(sum(ri_raw$lmg), 5),
    "= R2 =", round(summary(m_all)$r.squared, 5), "\n")
cat("shared_share = percentage of a criterion's LMG weight that is NOT unique.\n")
```

## Bootstrap intervals

`rela = FALSE` is used for the bootstrap so the weights stay in variance units
and the intervals are not distorted by the sum-to-100% constraint that
normalization imposes. `diff = TRUE` yields intervals on every pairwise
difference, which is the focal test; `rank = TRUE` yields intervals on each
criterion's rank position.

```{r}
#| label: relimp-boot

set.seed(42)
bb <- boot.relimp(m_all, b = 2000, type = "lmg",
                  rank = TRUE, diff = TRUE, rela = FALSE)
ev <- booteval.relimp(bb, level = 0.95)
```

```{r}
#| label: relimp-boot-table

nm <- ev@namen[-1]
boot_tab <- tibble(
  criterion = nm,
  lmg       = as.numeric(ev@lmg),
  lower     = as.numeric(ev@lmg.lower[1, ]),
  upper     = as.numeric(ev@lmg.upper[1, ])
) |>
  arrange(desc(lmg)) |>
  mutate(across(c(lmg, lower, upper), \(x) round(x, 5)))

print(as.data.frame(boot_tab), row.names = FALSE)
```

```{r}
#| label: relimp-plot
#| fig-height: 3.5
#| fig-width: 7
#| fig-cap: "LMG relative importance with 95% bootstrap intervals (2,000 replicates)"

boot_tab |>
  mutate(criterion = factor(criterion, levels = rev(criterion))) |>
  ggplot(aes(x = lmg, y = criterion)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.18,
                 colour = "grey40") +
  geom_point(size = 2.6) +
  labs(x = expression("LMG share of " * R^2), y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

## Pairwise differences

relaimpo's own formatted output, which carries the rank letters:

```{r}
#| label: relimp-print
print(ev)
```

```{r}
#| label: relimp-diffs

# The @diffnam slot comes back empty from booteval.relimp, but the difference
# vectors are ordered exactly as combn() enumerates the pairs, so the labels
# are reconstructed rather than read off. The stopifnot guards that assumption.
pair_mat <- combn(nm, 2)
pair_names <- paste0(pair_mat[1, ], " - ", pair_mat[2, ])
stopifnot(length(pair_names) == length(as.numeric(ev@lmg.diff)))

diff_tab <- tibble(
  pair  = pair_names,
  diff  = as.numeric(ev@lmg.diff),
  lower = as.numeric(ev@lmg.diff.lower[1, ]),
  upper = as.numeric(ev@lmg.diff.upper[1, ])
) |>
  mutate(excludes_zero = ifelse(lower > 0 | upper < 0, "yes", "no"),
         across(c(diff, lower, upper), \(x) round(x, 5))) |>
  arrange(desc(abs(diff)))

print(as.data.frame(diff_tab), row.names = FALSE)

cat("\nDifferences whose 95% interval excludes zero:",
    sum(diff_tab$excludes_zero == "yes"), "of", nrow(diff_tab), "\n")
```

```{r}
#| label: relimp-focal

focal <- diff_tab |>
  filter(str_detect(pair, "phraseology"), str_detect(pair, "grammar"))
cat("Focal comparison (phraseology vs grammar):\n")
print(as.data.frame(focal), row.names = FALSE)

cat("\nInterpretation: a negative difference means grammar carries the larger",
    "\nLMG share. The comparison is reliable only if the interval excludes 0.\n")
```

## Rank stability

```{r}
#| label: relimp-ranks

rank_tab <- tibble(
  criterion  = nm,
  rank_est   = as.numeric(ev@lmg.rank),
  rank_lower = as.numeric(ev@lmg.rank.lower[1, ]),
  rank_upper = as.numeric(ev@lmg.rank.upper[1, ])
) |> arrange(rank_est)
stopifnot(nrow(rank_tab) == length(nm))

print(as.data.frame(rank_tab), row.names = FALSE)
cat("\nRank 1 = largest LMG weight. Intervals are over 2,000 bootstrap samples.\n")
```

# Robustness

## Controlling for essay length

Word predictability is computed per token but essay length still varies widely,
and longer essays earn higher ratings. If the phraseological signal is really a
length effect, adding `n_tokens` should absorb it.

```{r}
#| label: length-control

common2_len <- function(dd, i) {
  R <- cor(dd[i, c("neg_loss", "phraseology", "grammar", "n_tokens")])
  r2 <- \(xs) r2_from_cor(R, "neg_loss", xs)
  full <- r2(c("phraseology", "grammar", "n_tokens"))
  c(U_phr  = full - r2(c("grammar", "n_tokens")),
    U_gram = full - r2(c("phraseology", "n_tokens")),
    U_len  = full - r2(c("phraseology", "grammar")),
    R2     = full)
}

est_len <- common2_len(dr, seq_len(nrow(dr)))
b_len <- boot(dr, common2_len, R = 5000)

res_len <- ci_table(b_len, est_len) |>
  mutate(across(c(estimate, lower, upper, boot_se), \(x) round(x, 5)))
print(as.data.frame(res_len), row.names = FALSE)

cat("\nU_phr without length control:", round(est2[["U_phr"]], 5),
    "\nU_phr with length control   :", round(est_len[["U_phr"]], 5), "\n")
```

## Residualizing on prompt

The network analysis residualizes every variable on writing prompt before
estimation. Repeating the two-predictor decomposition on prompt-residualized
ranks checks that the result is not driven by prompt composition.

```{r}
#| label: prompt-residualized

dr_resid <- dr |>
  mutate(across(all_of(c("neg_loss", "phraseology", "grammar")),
                \(x) residuals(lm(x ~ factor(d$prompt), na.action = na.exclude))))

est2_resid <- common2(dr_resid, seq_len(nrow(dr_resid)))
b2_resid <- boot(dr_resid, common2, R = 5000)

res2_resid <- ci_table(b2_resid, est2_resid) |>
  mutate(across(c(estimate, lower, upper, boot_se), \(x) round(x, 5)))
print(as.data.frame(res2_resid), row.names = FALSE)
```

```{r}
#| label: comparison-table

comparison <- bind_rows(
  res2       |> mutate(model = "raw ranks"),
  res2_resid |> mutate(model = "prompt-residualized")
) |>
  filter(component %in% c("U_phr", "U_gram", "Common", "R2", "Diff")) |>
  dplyr::select(model, component, estimate, lower, upper) |>
  pivot_wider(names_from = model, values_from = c(estimate, lower, upper))

print(as.data.frame(comparison), row.names = FALSE)
```

# Summary

```{r}
#| label: summary
#| output: asis

cat(sprintf(
"- Phraseology and grammar together explain **%.1f%%** of the rank variance in
  word predictability.
- Of that, **%.1f%%** is common to both ratings, **%.1f%%** is unique to
  phraseology, and **%.1f%%** is unique to grammar.
- U_phr - U_gram = **%.4f**, 95%% CI [%.4f, %.4f].
- Adding vocabulary raises the total to **%.1f%%**.
- Controlling essay length moves U_phr from %.4f to %.4f.
",
100 * est2[["R2"]],
100 * est2[["Common"]] / est2[["R2"]],
100 * est2[["U_phr"]] / est2[["R2"]],
100 * est2[["U_gram"]] / est2[["R2"]],
est2[["Diff"]], diff_ci$lower, diff_ci$upper,
100 * est3[["R2"]],
est2[["U_phr"]], est_len[["U_phr"]]
))
```

# Interpretation

**The directional hypothesis is not supported.** Grammar contributes roughly
1.7 times the unique variance that phraseology does, the bootstrap interval on
the difference excludes zero, and every one of the 5,000 replicates agrees in
sign. Neither the length control nor prompt residualization overturns it. If
predictability were distinctively phraseological among the human ratings, this
is where it would show, and it does not.

**But both unique components are small, and that is the larger finding.** Nearly
70% of what phraseology and grammar jointly explain is common to them. With
vocabulary added, the three-way component alone is over half the total $R^2$,
while unique vocabulary is under 1%. What predicts word predictability in this
rating set is overwhelmingly *general proficiency as perceived by raters*, not
any separable analytic dimension. Reporting the unique components without the
common one would badly misdescribe that.

**This does not directly contradict the network analysis, because the two use
different operationalizations.** The network placed `mean_loss` in a community
with the *dependency-MI indices* (`amod_mi`, `dobj_mi`, `advmod_mi`) — corpus-
derived association strengths computed from the Dolma reference corpus. The
predictors here are *human ratings* that happen to be named Phraseology and
Grammar. A rater's holistic phraseology judgment and a pointwise MI statistic
are not the same construct, and the rating carries a halo the MI indices do not.
The defensible summary is that surprisal aligns with corpus-derived formulaicity
at the level of measured indices, while at the level of human ratings it tracks
general proficiency with a modest grammatical edge. Treating the two as rival
answers to one question would be a category error.

# References

Newton, R. G., & Spurrell, D. J. (1967). A development of multiple regression
for the analysis of routine data. *Applied Statistics, 16*(1), 51–64.

Nimon, K. F., & Oswald, F. L. (2013). Understanding the results of multiple
linear regression: Beyond standardized regression coefficients.
*Organizational Research Methods, 16*(4), 650–674.
