---
title: "Network Psychometric Analysis of ELLIPSE Metrics"
subtitle: "Where Does Word Predictability Fit in L2 Writing Proficiency Structure?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

This analysis complements the CFA approach by using a data-driven
Gaussian Graphical Model (GGM) to empirically discover how word predictability relates
to other linguistic features — without imposing a factor structure.

**Key question:** Does word predictability (`mean_loss`) cluster with
phraseological, lexical, or other features — or does it form its own community?

# Section 1 — Setup & Data Preparation

```{r}
#| label: setup
library(tidyverse)
library(psych)
library(corrplot)
library(qgraph)      # visualization only
library(bootnet)     # edge-weight accuracy CIs + group viz networks only
library(igraph)
library(EGAnet)      # estimation, communities, network loadings, network.compare
library(lavaan)

# NOTE: EGAnet 2.x function signatures have changed across releases. The
# net.loads(), network.compare(), EGA(corr=), and invariance() calls below
# follow the currently documented signatures — verify argument names against
# the INSTALLED version (?net.loads, ?network.compare, ?EGA) before rendering.
```

```{r}
#| label: load-data
# ellipse_metrics.csv carries the 13 linguistic features + ELLIPSE scores +
# predictability (mean_loss, var_loss) from the Study 1 benchmark model
# (Llama-3.1-8B base). All networks below use Spearman correlations.
df_raw <- read.csv("../../data/ellipse/ellipse_metrics.csv")

cat("Raw data:", nrow(df_raw), "rows x", ncol(df_raw), "columns\n")
```

```{r}
#| label: select-features

# Select the 16 feature variables
feature_vars <- c(
  # Lexical (4)
  "mtld", "lexical_density", "log_mean_token_freq", "mean_word_length",
  # Syntactic (4 planned, but clauses_per_tunit is constant)
  "words_per_sentence", "clauses_per_tunit", "mod_per_nom", "mean_dep_distance",
  # Phraseological (3)
  "amod_mi", "dobj_mi", "advmod_mi",
  # Cohesion (3)
  "content_word_overlap", "connective_density", "sentence_similarity",
  # Predictability (2)
  "mean_loss", "var_loss"
)

df_features <- df_raw[, feature_vars]

# Check clauses_per_tunit variance
cat("clauses_per_tunit — unique values:", length(unique(df_features$clauses_per_tunit)),
    "| sd:", sd(df_features$clauses_per_tunit, na.rm = TRUE), "\n")

# Drop clauses_per_tunit — zero variance (all values = 1)
df_features$clauses_per_tunit <- NULL
cat("Dropped clauses_per_tunit (constant). Remaining:", ncol(df_features), "variables\n")
```

```{r}
#| label: check-collinearity

# mean_entropy removed due to near-perfect correlation with mean_loss (r = 0.97)
# which created a trivial island in the network
cat("Note: mean_entropy excluded (r = 0.97 with mean_loss). Using mean_loss + var_loss.\n")
cat("var_loss captures consistency of predictability (variance of token-level loss).\n")
```

```{r}
#| label: missing-data

# Report missingness
missing_counts <- colSums(is.na(df_features))
cat("Missing values per variable:\n")
print(missing_counts[missing_counts > 0])
cat("\nTotal rows before listwise deletion:", nrow(df_features), "\n")

# Listwise deletion
complete_idx <- complete.cases(df_features)
df_complete <- df_features[complete_idx, ]
prompt_complete <- df_raw$prompt[complete_idx]
cat("Total rows after listwise deletion:", nrow(df_complete),
    "(dropped", nrow(df_features) - nrow(df_complete), "rows)\n")
```

```{r}
#| label: winsorize

# Winsorize at 1st/99th percentiles. Retained for continuity; under Spearman
# (rank-based) estimation this has negligible effect and could be dropped.
winsorize <- function(x, probs = c(0.01, 0.99)) {
  q <- quantile(x, probs = probs, na.rm = TRUE)
  x[x < q[1]] <- q[1]
  x[x > q[2]] <- q[2]
  x
}

# Show pre-winsorization extremes for key variables
cat("Pre-winsorization extremes:\n")
cat("  words_per_sentence: max =", max(df_complete$words_per_sentence), "\n")
cat("  connective_density: max =", max(df_complete$connective_density), "\n")

df_win <- df_complete %>%
  mutate(across(everything(), winsorize))

cat("\nPost-winsorization extremes:\n")
cat("  words_per_sentence: max =", max(df_win$words_per_sentence), "\n")
cat("  connective_density: max =", max(df_win$connective_density), "\n")
```

```{r}
#| label: residualize-on-prompt

# Residualize all variables on prompt to remove systematic prompt-driven variance
df_resid <- df_win
r2_values <- numeric(ncol(df_win))
names(r2_values) <- names(df_win)

for (var in names(df_win)) {
  fit <- lm(df_win[[var]] ~ factor(prompt_complete))
  df_resid[[var]] <- residuals(fit)
  r2_values[var] <- summary(fit)$r.squared
}

df_win <- df_resid  # all downstream code uses residualized data

cat("R² (variance explained by prompt) per variable:\n")
r2_df <- data.frame(Variable = names(r2_values), R2 = round(r2_values, 4))
r2_df <- r2_df[order(-r2_df$R2), ]
print(r2_df, row.names = FALSE)
```

```{r}
#| label: define-groups

# Theoretical groupings for coloring (after dropping clauses_per_tunit)
groups <- list(
  Lexical       = c("mtld", "lexical_density", "log_mean_token_freq", "mean_word_length"),
  Syntactic     = c("words_per_sentence", "mod_per_nom", "mean_dep_distance"),
  Phraseological = c("amod_mi", "dobj_mi", "advmod_mi"),
  Cohesion      = c("content_word_overlap", "connective_density", "sentence_similarity"),
  Predictability = c("mean_loss", "var_loss")
)

# Short labels for plotting
short_labels <- c(
  mtld = "MTLD",
  lexical_density = "LexDen",
  log_mean_token_freq = "TokFreq",
  mean_word_length = "WordLen",
  words_per_sentence = "WPS",
  mod_per_nom = "ModNom",
  mean_dep_distance = "DepDist",
  amod_mi = "AmodMI",
  dobj_mi = "DobjMI",
  advmod_mi = "AdvMI",
  content_word_overlap = "CWO",
  connective_density = "ConnDen",
  sentence_similarity = "SentSim",
  mean_loss = "Loss",
  var_loss = "VarLoss"
)

# Color scheme
group_colors <- c(
  Lexical = "#E41A1C",        # red
  Syntactic = "#377EB8",      # blue
  Phraseological = "#4DAF4A", # green
  Cohesion = "#FF7F00",       # orange
  Predictability = "#984EA3"  # purple
)

# Map each variable to its group color
node_colors <- character(ncol(df_win))
names(node_colors) <- names(df_win)
for (g in names(groups)) {
  node_colors[groups[[g]]] <- group_colors[g]
}
```


# Section 2 — Descriptive Statistics & Correlation Matrix

```{r}
#| label: descriptive-stats
#| tbl-cap: "Descriptive statistics for all network variables"

desc_stats <- describe(df_win)
print(desc_stats[, c("n", "mean", "sd", "min", "max", "skew", "kurtosis")], digits = 3)
```

```{r}
#| label: correlation-heatmap
#| fig-cap: "Spearman correlation matrix of all network variables"
#| fig-width: 10
#| fig-height: 9

# Spearman to match the network estimation input (Isvoranu & Epskamp, 2023)
cor_mat <- cor(df_win, use = "pairwise.complete.obs", method = "spearman")

# Use short labels for readability
colnames(cor_mat) <- short_labels[colnames(cor_mat)]
rownames(cor_mat) <- short_labels[rownames(cor_mat)]

corrplot(cor_mat, method = "color", type = "lower",
         tl.col = "black", tl.cex = 0.9,
         addCoef.col = "black", number.cex = 0.6,
         col = colorRampPalette(c("#2166AC", "white", "#B2182B"))(200),
         diag = FALSE)
```

```{r}
#| label: normality-check
#| fig-cap: "Distribution of all network variables"
#| fig-width: 12
#| fig-height: 10

df_win %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  mutate(variable = short_labels[variable]) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white", linewidth = 0.2) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  theme_minimal(base_size = 11) +
  labs(x = NULL, y = "Count")
```


# Section 3 — Network Estimation (EBICglasso)

```{r}
#| label: estimate-network

# Estimate Gaussian Graphical Model (EBICglasso) on Spearman correlations
# (Isvoranu & Epskamp, 2023). This bootnet object is used only for
# visualization (Section 4) and edge-weight bootstrapping (Section 7).
network <- estimateNetwork(df_win, default = "EBICglasso", tuning = 0.5,
                           corMethod = "spearman")

cat("Network estimated with", ncol(network$graph), "nodes\n")
cat("Non-zero edges:", sum(network$graph[upper.tri(network$graph)] != 0),
    "out of", sum(upper.tri(network$graph)), "possible\n")
cat("Sparsity:", round(1 - mean(network$graph[upper.tri(network$graph)] != 0), 3), "\n")
```

```{r}
#| label: adjacency-matrix
#| tbl-cap: "Weighted adjacency matrix (partial correlations)"

adj_mat <- round(network$graph, 3)
colnames(adj_mat) <- short_labels[colnames(adj_mat)]
rownames(adj_mat) <- short_labels[rownames(adj_mat)]
print(adj_mat)
```


# Section 4 — Network Visualization

```{r}
#| label: network-spring
#| fig-cap: "GGM network (spring layout, Fruchterman-Reingold)"
#| fig-width: 10
#| fig-height: 10

# Assign short labels for the plot
net_labels <- short_labels[colnames(network$graph)]

spring_layout <- qgraph(network$graph,
       layout = "spring",
       labels = net_labels,
       groups = groups,
       color = group_colors,
       legend = TRUE,
       legend.cex = 0.45,
       borders = TRUE,
       border.width = 1.5,
       vsize = 8,
       esize = 8,
       edge.color = NULL,
       posCol = "#2166AC",
       negCol = "#B2182B",
       title = "ELLIPSE Linguistic Features — GGM Network",
       title.cex = 1.2)
```

```{r}
#| label: save-spring-plot
#| include: false

pdf("../../fig/network_spring.pdf", width = 10, height = 10)
qgraph(network$graph,
       layout = "spring",
       labels = net_labels,
       groups = groups,
       color = group_colors,
       legend = TRUE,
       legend.cex = 0.45,
       borders = TRUE,
       border.width = 1.5,
       vsize = 8,
       esize = 8,
       posCol = "#2166AC",
       negCol = "#B2182B",
       title = "ELLIPSE Linguistic Features — GGM Network")
dev.off()
cat("Saved to fig/network_spring.pdf\n")
```

```{r}
#| label: network-circle
#| fig-cap: "GGM network (circle layout)"
#| fig-width: 10
#| fig-height: 10

qgraph(network$graph,
       layout = "circle",
       labels = net_labels,
       groups = groups,
       color = group_colors,
       legend = TRUE,
       legend.cex = 0.45,
       borders = TRUE,
       border.width = 1.5,
       vsize = 8,
       esize = 8,
       posCol = "#2166AC",
       negCol = "#B2182B",
       title = "ELLIPSE Linguistic Features — GGM Network (Circle)")
```


# Section 5 — Centrality Analysis

```{r}
#| label: centrality-plot
#| fig-cap: "Centrality indices for all nodes"
#| fig-width: 10
#| fig-height: 8

# Strength (primary) + signed ExpectedInfluence. Betweenness/Closeness dropped:
# non-informative in cross-sectional psychometric networks. Network loadings
# (Section 6) are the confound-adjusted node-importance metric and supersede
# raw strength for interpretation.
centralityPlot(network$graph, labels = net_labels,
               include = c("Strength", "ExpectedInfluence"),
               orderBy = "Strength")
```

```{r}
#| label: centrality-table
#| tbl-cap: "Centrality values for all nodes"

cent <- centrality_auto(network$graph)$node.centrality
cent$Variable <- short_labels[rownames(cent)]
# Keep Strength + ExpectedInfluence only (drop Betweenness/Closeness)
cent <- cent %>% select(Variable, Strength, any_of("ExpectedInfluence"))
print(cent %>% arrange(desc(Strength)), digits = 3)
```

```{r}
#| label: predictability-edges
#| tbl-cap: "Edge weights involving predictability nodes"

# Extract edges for mean_loss and var_loss
adj <- network$graph
loss_edges <- adj["mean_loss", ]
var_loss_edges <- adj["var_loss", ]

pred_edges <- data.frame(
  Partner = short_labels[names(loss_edges)],
  Loss_edge = round(loss_edges, 3),
  VarLoss_edge = round(var_loss_edges, 3)
) %>%
  filter(!Partner %in% c("Loss", "VarLoss")) %>%
  arrange(desc(abs(Loss_edge)))

cat("Strongest connections of predictability nodes:\n")
print(pred_edges)
```


# Section 6 — Community Detection (EGA)

```{r}
#| label: ega
#| fig-cap: "Exploratory Graph Analysis — community detection"
#| fig-width: 10
#| fig-height: 10

ega_result <- EGA(df_win, corr = "spearman", model = "glasso",
                  algorithm = "walktrap",
                  plot.EGA = TRUE)

cat("\nNumber of communities detected:", ega_result$n.dim, "\n")
cat("\nCommunity assignments:\n")

# Show assignments with short labels
assignments <- data.frame(
  Variable = short_labels[names(ega_result$wc)],
  Full_name = names(ega_result$wc),
  Community = ega_result$wc
)
print(assignments %>% arrange(Community))
```

```{r}
#| label: ega-predictability

# Highlight where predictability ended up
cat("Predictability node assignments:\n")
cat("  mean_loss -> Community", ega_result$wc["mean_loss"], "\n")
cat("  var_loss  -> Community", ega_result$wc["var_loss"], "\n")

# What else is in the same community as mean_loss?
comm <- ega_result$wc["mean_loss"]
members <- names(ega_result$wc)[ega_result$wc == comm]
cat("\nCommunity", comm, "(containing Loss):\n")
cat("  Members:", paste(short_labels[members], collapse = ", "), "\n")

# What else is in the same community as var_loss?
comm_var <- ega_result$wc["var_loss"]
members_var <- names(ega_result$wc)[ega_result$wc == comm_var]
cat("\nCommunity", comm_var, "(containing VarLoss):\n")
cat("  Members:", paste(short_labels[members_var], collapse = ", "), "\n")
```

```{r}
#| label: net-loadings
#| tbl-cap: "Revised network loadings (standardized) — primary node importance"

# Network loadings (Christensen, Golino, Abad & Garrido, 2025): confound-adjusted
# node importance that splits strength by community and, in the revised
# formulation, adjusts for community size and sign. These supersede raw strength.
# Effect-size guides: ~0.20 small, ~0.35 moderate, ~0.50 large.
# VERIFY loading.method value + return element ($std) against installed EGAnet.
nl <- net.loads(ega_result, loading.method = "revised")$std

nl_df <- as.data.frame(round(nl, 3))
nl_df$Variable <- short_labels[rownames(nl)]
nl_df <- nl_df %>% select(Variable, everything())
cat("Standardized network loadings (nodes x communities):\n")
print(nl_df, row.names = FALSE)

cat("\nPredictability network loadings:\n")
print(nl_df %>% filter(Variable %in% c("Loss", "VarLoss")))
```

```{r}
#| label: boot-ega
#| fig-cap: "Bootstrap EGA — stability of community assignments"

set.seed(42)
boot_ega <- bootEGA(df_win, iter = 500,
                    corr = "spearman",
                    model = "glasso",
                    algorithm = "walktrap",
                    type = "parametric",
                    verbose = FALSE)

cat("Modal number of communities across bootstraps:",
    as.numeric(names(sort(table(boot_ega$n.dim), decreasing = TRUE))[1]), "\n")

# Structural consistency
cat("\nItem stability (proportion of bootstraps in modal community):\n")
stab <- boot_ega$frequency
print(round(stab, 3))
```


# Section 7 — Network Stability (bootnet)

```{r}
#| label: boot-edges
#| fig-cap: "Bootstrap confidence intervals for edge weights"
#| fig-width: 12
#| fig-height: 10

set.seed(42)
boot_nonparam <- bootnet(network, nBoots = 1000, type = "nonparametric",
                         nCores = parallel::detectCores() - 1,
                         verbose = FALSE)

plot(boot_nonparam, labels = net_labels,
     order = "sample", CIstyle = "quantiles")
```

```{r}
#| label: save-edge-boot
#| include: false

pdf("../../fig/network_edge_bootstrap.pdf", width = 12, height = 10)
plot(boot_nonparam, labels = net_labels,
     order = "sample", CIstyle = "quantiles")
dev.off()
cat("Saved to fig/network_edge_bootstrap.pdf\n")
```

```{r}
#| label: boot-centrality
#| fig-cap: "Case-dropping bootstrap for centrality stability"
#| fig-width: 10
#| fig-height: 8

set.seed(42)
boot_case <- bootnet(network, nBoots = 1000, type = "case",
                     statistics = c("strength", "expectedInfluence"),
                     nCores = parallel::detectCores() - 1,
                     verbose = FALSE)

plot(boot_case, statistics = c("strength", "expectedInfluence"))
```

```{r}
#| label: cs-coefficients

cs_coefs <- corStability(boot_case)
cat("CS-coefficients (correlation stability):\n")
print(cs_coefs)

cat("\nInterpretation guidelines:\n")
cat("  CS >= 0.50: Good, centrality is stable and interpretable\n")
cat("  CS >= 0.25: Acceptable minimum\n")
cat("  CS < 0.25: Centrality should NOT be interpreted\n")
```


# Section 8 — Between-Community Loadings (Bridge Behavior)

Bridge behavior is operationalized via **between-community (cross-)loadings**
from the revised network loadings (Christensen et al., 2025), not Jones et al.
(2021) bridge strength. Cross-loadings distinguish a genuine bridge (a node that
loads on multiple communities) from mere between-community correlation.

```{r}
#| label: cross-loadings
#| tbl-cap: "Between-community (cross-)loadings — bridge behavior"

# `nl` (standardized network loadings, nodes x communities) is computed in
# Section 6. A node's loading on its OWN community vs. its loadings on the
# others: high & roughly-equal cross-loadings = integrative/bridge behavior;
# a single dominant loading = the node belongs to one community.
# VERIFY that nl's columns are named by community id (matches ega_result$wc).
assigned <- ega_result$wc  # community per node

cross_load_summary <- function(node) {
  row <- nl[node, ]
  own <- as.character(assigned[node])
  others <- setdiff(names(row), own)
  data.frame(
    Variable = short_labels[node],
    Own_community = own,
    Own_loading = round(row[[own]], 3),
    Max_cross = round(max(row[others]), 3),
    Mean_abs_cross = round(mean(abs(row[others])), 3)
  )
}

cat("Standardized loadings (nodes x communities):\n")
print(round(nl, 3))

cat("\nPredictability nodes — in-community vs cross-community loading:\n")
print(do.call(rbind, lapply(c("mean_loss", "var_loss"), cross_load_summary)),
      row.names = FALSE)
```

# Section 9 — Network Comparison by Proficiency Level

```{r}
#| label: proficiency-split

# Use single Overall score from public ELLIPSE data
df_raw_complete <- df_raw[complete.cases(df_raw[, feature_vars[feature_vars != "clauses_per_tunit"]]), ]
prompt_raw_complete <- df_raw_complete$prompt

# Define proficiency groups
df_raw_complete$prof_group <- case_when(
  df_raw_complete$Overall <= 2.5 ~ "Low",
  df_raw_complete$Overall >= 4.0 ~ "High",
  TRUE ~ "Mid"
)

cat("Proficiency group sizes:\n")
print(table(df_raw_complete$prof_group))

# Select feature columns (without clauses_per_tunit)
keep_vars <- setdiff(feature_vars, "clauses_per_tunit")

# Low proficiency: winsorize then residualize on prompt
low_idx <- df_raw_complete$prof_group == "Low"
df_low <- df_raw_complete[low_idx, keep_vars] %>%
  mutate(across(everything(), winsorize))
prompt_low <- prompt_raw_complete[low_idx]

for (var in names(df_low)) {
  fit <- lm(df_low[[var]] ~ factor(prompt_low))
  df_low[[var]] <- residuals(fit)
}

# High proficiency: winsorize then residualize on prompt
high_idx <- df_raw_complete$prof_group == "High"
df_high <- df_raw_complete[high_idx, keep_vars] %>%
  mutate(across(everything(), winsorize))
prompt_high <- prompt_raw_complete[high_idx]

for (var in names(df_high)) {
  fit <- lm(df_high[[var]] ~ factor(prompt_high))
  df_high[[var]] <- residuals(fit)
}

cat("\nLow proficiency n:", nrow(df_low), "\n")
cat("High proficiency n:", nrow(df_high), "\n")
```

```{r}
#| label: estimate-group-networks

net_low <- estimateNetwork(df_low, default = "EBICglasso", tuning = 0.5,
                           corMethod = "spearman")
net_high <- estimateNetwork(df_high, default = "EBICglasso", tuning = 0.5,
                            corMethod = "spearman")
```

```{r}
#| label: group-network-plots
#| fig-cap: "Network comparison: Low vs High proficiency"
#| fig-width: 14
#| fig-height: 7

# Use the same layout for both (average of spring layouts)
avg_layout <- averageLayout(net_low$graph, net_high$graph)

par(mfrow = c(1, 2))

qgraph(net_low$graph,
       layout = avg_layout,
       labels = net_labels,
       groups = groups,
       color = group_colors,
       legend = FALSE,
       borders = TRUE,
       border.width = 1.5,
       vsize = 7,
       esize = 8,
       posCol = "#2166AC",
       negCol = "#B2182B",
       title = "Low Proficiency",
       title.cex = 1.3,
       maximum = max(abs(c(net_low$graph, net_high$graph))))

qgraph(net_high$graph,
       layout = avg_layout,
       labels = net_labels,
       groups = groups,
       color = group_colors,
       legend = FALSE,
       borders = TRUE,
       border.width = 1.5,
       vsize = 7,
       esize = 8,
       posCol = "#2166AC",
       negCol = "#B2182B",
       title = "High Proficiency",
       title.cex = 1.3,
       maximum = max(abs(c(net_low$graph, net_high$graph))))

par(mfrow = c(1, 1))
```

```{r}
#| label: save-group-plots
#| include: false

pdf("../../fig/network_group_comparison.pdf", width = 14, height = 7)
par(mfrow = c(1, 2))
qgraph(net_low$graph,
       layout = avg_layout, labels = net_labels,
       groups = groups, color = group_colors, legend = FALSE,
       borders = TRUE, border.width = 1.5, vsize = 7, esize = 8,
       posCol = "#2166AC", negCol = "#B2182B",
       title = "Low Proficiency", title.cex = 1.3,
       maximum = max(abs(c(net_low$graph, net_high$graph))))
qgraph(net_high$graph,
       layout = avg_layout, labels = net_labels,
       groups = groups, color = group_colors, legend = FALSE,
       borders = TRUE, border.width = 1.5, vsize = 7, esize = 8,
       posCol = "#2166AC", negCol = "#B2182B",
       title = "High Proficiency", title.cex = 1.3,
       maximum = max(abs(c(net_low$graph, net_high$graph))))
par(mfrow = c(1, 1))
dev.off()
cat("Saved to fig/network_group_comparison.pdf\n")
```

```{r}
#| label: network-compare
#| output: false

# EGAnet permutation network comparison replaces NetworkComparisonTest::NCT.
# Feed the residualized group DATA FRAMES (not the network objects).
# Returns per-measure permutation tests: Jensen-Shannon distance, Frobenius-norm
# ("SF") similarity, and total network strength (van Borkulo aspect), plus
# per-edge statistics. VERIFY arg + return names against installed EGAnet.
set.seed(42)
nc <- network.compare(
  base = df_low, comparison = df_high,
  corr = "spearman", model = "glasso",
  iter = 1000, seed = 42
)

cat("=== Permutation Network Comparison (network.compare) ===\n\n")
cat("Omnibus structural comparison (statistic + permutation p per measure):\n")
print(nc$network)   # Jensen-Shannon, Frobenius/SF, total strength
```

```{r}
#| label: network-compare-edges
#| output: false

# Per-edge permutation differences (BH-adjusted). Column names may vary by
# EGAnet version — inspect nc$edges and adjust the filter if needed.
cat("Per-edge differences (p < 0.05):\n")
sig_edges <- nc$edges %>% filter(p.value < 0.05)
if (nrow(sig_edges) > 0) {
  print(sig_edges)
} else {
  cat("No individual edges significantly differ between groups.\n")
}
```

```{r}
#| label: invariance-optional
#| eval: false

# OPTIONAL (author decision — Open Item 3): if RQ4 is "are the network LOADINGS
# invariant across proficiency" rather than "does the structure differ", use the
# loadings-based permutation invariance test (Jamison, Christensen & Golino,
# 2024) instead of / in addition to network.compare above. Left un-run by
# default; flip `eval` to true if the RQ is framed as invariance.
inv <- invariance(
  data = rbind(df_low, df_high),
  groups = c(rep("Low", nrow(df_low)), rep("High", nrow(df_high))),
  corr = "spearman", iter = 1000, seed = 42
)
print(inv$results)   # per-node loading differences + permutation p-values
```

```{r}
#| label: nct-predictability

# Predictability-focused edge-weight differences (effect sizes from the two
# Spearman group networks). Per-edge significance comes from nc$edges above.
cat("Edge weight differences involving predictability nodes:\n\n")

adj_low <- net_low$graph
adj_high <- net_high$graph

pred_comparison <- data.frame(
  Partner = short_labels[colnames(adj_low)],
  Low_Loss = round(adj_low["mean_loss", ], 3),
  High_Loss = round(adj_high["mean_loss", ], 3),
  Diff_Loss = round(adj_high["mean_loss", ] - adj_low["mean_loss", ], 3),
  Low_VarLoss = round(adj_low["var_loss", ], 3),
  High_VarLoss = round(adj_high["var_loss", ], 3),
  Diff_VarLoss = round(adj_high["var_loss", ] - adj_low["var_loss", ], 3)
) %>%
  filter(!Partner %in% c("Loss", "VarLoss")) %>%
  arrange(desc(abs(Diff_Loss)))

print(pred_comparison)
```


# Summary & Interpretation

```{r}
#| label: summary

cat("=" %>% strrep(60), "\n")
cat("SUMMARY OF NETWORK ANALYSIS RESULTS\n")
cat("=" %>% strrep(60), "\n\n")

# 1. Community structure
cat("1. COMMUNITY STRUCTURE (EGA)\n")
cat("   Number of communities:", ega_result$n.dim, "\n")
cat("   mean_loss assigned to community:", ega_result$wc["mean_loss"], "\n")
cat("   var_loss assigned to community:", ega_result$wc["var_loss"], "\n\n")

# What's in the same community as predictability?
pred_comm <- ega_result$wc["mean_loss"]
same_comm <- names(ega_result$wc)[ega_result$wc == pred_comm]
cat("   Variables in same community as mean_loss:\n")
cat("  ", paste(short_labels[same_comm], collapse = ", "), "\n")

varloss_comm <- ega_result$wc["var_loss"]
same_comm_var <- names(ega_result$wc)[ega_result$wc == varloss_comm]
cat("   Variables in same community as var_loss:\n")
cat("  ", paste(short_labels[same_comm_var], collapse = ", "), "\n\n")

# 2. Node importance — network loadings (primary), strength (secondary)
cat("2. NODE IMPORTANCE (network loadings primary)\n")
loss_own <- as.character(ega_result$wc["mean_loss"])
cat("   mean_loss loading on own community:", round(nl["mean_loss", loss_own], 3), "\n")
cat("   mean_loss max cross-community loading:",
    round(max(nl["mean_loss", setdiff(colnames(nl), loss_own)]), 3), "\n")
cent_sorted <- cent %>% arrange(desc(Strength))
loss_rank <- which(cent_sorted$Variable == "Loss")
cat("   (secondary) mean_loss raw-strength rank:", loss_rank, "of", nrow(cent_sorted), "\n\n")

# 3. Bridge behavior — between-community cross-loadings
cat("3. BRIDGE BEHAVIOR (cross-loadings; see Section 8)\n")
cat("   High & roughly-equal cross-loadings = integrative/bridge;\n")
cat("   a single dominant loading = belongs to one community.\n\n")

# 4. Interpretation framework
cat("4. INTERPRETATION RELATIVE TO CFA HYPOTHESES\n")
cat("   | CFA Model | Supporting Evidence |\n")
cat("   |-----------|--------------------|\n")
cat("   | Model 1 (Phraseological) | Predictability in same community as MI? ",
    pred_comm == ega_result$wc["amod_mi"], "|\n")
cat("   | Model 2 (General factor)  | High centrality? Loss rank =", loss_rank, "|\n")
cat("   | Novel finding             | Check community membership above |\n\n")

# 5. Stability
cat("5. STABILITY\n")
cat("   CS-coefficients reported in Section 7\n")
cat("   Edge accuracy CIs reported in Section 7\n\n")

# 6. Proficiency differences — permutation network.compare (nc$network)
cat("6. PROFICIENCY GROUP DIFFERENCES (network.compare)\n")
cat("   Omnibus structural comparison (Jensen-Shannon, Frobenius/SF,\n")
cat("   total strength) with permutation p-values:\n")
print(nc$network)
```
