---
title: "Relative Importance and Criterion Profiles"
subtitle: "Which ELLIPSE criteria does word predictability track?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

The network analysis places `mean_loss` in a community with the dependency-MI
features. This notebook asks a related but **distinct** question, of the human
ratings rather than the computed indices: **across the six ELLIPSE analytic
criteria, which ones does word predictability track, and by how much?**

Two complementary answers are given.

1. **Relative importance (LMG).** The six criteria jointly predict
   predictability; LMG partitions that $R^2$ into a non-negative share per
   criterion. This is a *variance-explained* answer.
2. **Criterion profile similarity.** Every measured index has a profile of
   correlations across the six criteria. Comparing predictability's profile to
   the profile of the phraseological family and of the syntactic-elaboration
   family asks whether predictability *behaves like* one family or the other.
   This is a *pattern* answer, and it does not depend on how much variance
   anything explains.

The distinction from the network analysis matters for reading either result — a
rater's holistic `Phraseology` judgment and a corpus-derived MI index are
different operationalizations, so this notebook does not adjudicate the network
finding. See the Interpretation section.

**Sign convention.** All analyses use `neg_loss = -mean_loss`, so that higher
values mean *more predictable* text and every coefficient reads in the same
direction as the proficiency scores.

```{r}
#| label: setup
library(tidyverse)
library(boot)
library(relaimpo)

set.seed(42)
```

`relaimpo` attaches `MASS`, which masks `dplyr::select()`. Loading it here rather
than mid-document means any masking problem surfaces at the top of the render;
`select()` is qualified where it is used.

```{r}
#| label: load-data

d_raw <- read_csv("data/ellipse/ellipse_metrics.csv",
                  show_col_types = FALSE)

# The six analytic criteria, in the order ELLIPSE reports them.
crit <- c("cohesion", "syntax", "vocabulary",
          "phraseology", "grammar", "conventions")

# The two index families compared against predictability in the profile
# analysis. `mi_vars` is the phraseological family (pointwise MI of dependency
# bigrams against Dolma); `syn_vars` is the syntactic-elaboration family (the
# L2SCA battery, minus C/T -- see network-analysis.qmd for why C/T is excluded).
mi_vars  <- c("amod_mi", "dobj_mi", "advmod_mi")
syn_vars <- c("mean_length_of_clause", "complex_nominals_per_clause",
              "coord_phrases_per_clause")

d <- d_raw |>
  transmute(
    text_id_kaggle,
    prompt,
    neg_loss    = -mean_loss,
    cohesion    = Cohesion,
    syntax      = Syntax,
    vocabulary  = Vocabulary,
    phraseology = Phraseology,
    grammar     = Grammar,
    conventions = Conventions,
    n_tokens    = n_tokens,
    across(all_of(c(mi_vars, syn_vars)))
  )

cat("Rows loaded:", nrow(d), "\n")
na_counts <- colSums(is.na(d))
if (any(na_counts > 0)) {
  cat("Missing values:\n"); print(na_counts[na_counts > 0])
}

d <- drop_na(d)
cat("Complete cases used:", nrow(d), "\n\n")

cat("Rating scales (ELLIPSE ratings are ordinal, 1-5 in 0.5 steps):\n")
print(map_dfr(d[crit],
              \(x) tibble(min = min(x), max = max(x),
                          levels = n_distinct(x), sd = sd(x)),
              .id = "variable"))
```

## Rank transformation

Everything below is computed on **ranks**. With only ~9 distinct rating levels
and a bounded scale, Pearson correlations on the raw scores are not the right
summary; ranking first makes every correlation a Spearman correlation and every
simple $R^2$ a squared Spearman correlation, which keeps the regression and the
profile analysis on one consistent footing.

Ranks are computed **once**, on the full sample, and the bootstrap then
resamples rows of the ranked data. This treats the ranks as the data rather than
re-ranking inside each resample. Re-ranking within resamples is the stricter
analogue of a Spearman bootstrap, but with *n* > 6,000 the two are
indistinguishable and re-ranking is far slower.

```{r}
#| label: rank-transform

vars <- c("neg_loss", crit, "n_tokens", mi_vars, syn_vars)
dr <- d |> mutate(across(all_of(vars), \(x) rank(x)))

# Pearson on average ranks is Spearman on the raw values, exactly. Verified
# rather than asserted, because the whole notebook rests on it.
stopifnot(all(abs(
  cor(dr$neg_loss, dr[crit]) -
  cor(d$neg_loss, d[crit], method = "spearman")
) < 1e-12))
cat("Rank-Pearson reproduces Spearman to within 1e-12.\n")
```

## Zero-order correlations

Reported first for context, and because the criterion intercorrelations *are*
the halo estimate — they set an upper bound on how much unique variance any one
rating can possibly have.

```{r}
#| label: zero-order

cat("Spearman correlation of neg_loss with each criterion:\n")
print(round(sort(cor(dr$neg_loss, dr[crit])[1, ], decreasing = TRUE), 3))

cat("\nCriterion intercorrelations:\n")
print(round(cor(dr[crit]), 3))

cat("\nMean off-diagonal criterion intercorrelation:",
    round(mean(cor(dr[crit])[lower.tri(diag(6))]), 3), "\n")
```

# Relative importance across the six criteria

With the criteria this heavily intercorrelated, the *unique* variance of each
one (its squared semipartial) understates every predictor, and understates the
most redundant ones most. Ranking criteria by unique variance therefore rewards
whichever rating happens to be least redundant with the others, which is not the
same thing as importance.

**LMG** (Lindeman, Merenda & Gold, 1980), the Shapley-value decomposition,
answers the question the unique components cannot. It averages each predictor's
sequential $R^2$ contribution over **all $6! = 720$ orderings** of entry, so
shared variance is *allocated* among the predictors that share it rather than
discarded. The weights are non-negative and sum exactly to $R^2$, which makes
them readable as "this criterion's share of the explained variance."

Both metrics are reported below, because the contrast between them is
informative in itself:

- `lmg` — Shapley share, shared variance allocated.
- `last` — the variance a criterion adds when entered last. This is exactly the
  squared semipartial, i.e. the criterion's *unique* contribution.

A criterion with a large `lmg` but a near-zero `last` contributes almost
entirely through variance it shares with the other ratings.

```{r}
#| label: relimp-setup

m_all <- lm(reformulate(crit, "neg_loss"), data = dr)
r2_full <- summary(m_all)$r.squared
cat("Full model R2 =", round(r2_full, 5), "\n")
```

```{r}
#| label: relimp-validate

# `last` must equal R2(full) - R2(full minus that criterion), by definition.
# Checked against refitted models so the decomposition is demonstrably the
# quantity claimed for it, not merely a number relaimpo returned.
drop_last <- map_dbl(crit, \(s)
  r2_full - summary(lm(reformulate(setdiff(crit, s), "neg_loss"), dr))$r.squared)
last_ri <- calc.relimp(m_all, type = "last", rela = FALSE)$last[crit]

val <- tibble(criterion     = crit,
              relaimpo_last = as.numeric(last_ri),
              by_refit      = drop_last) |>
  mutate(abs_diff = abs(relaimpo_last - by_refit))

print(as.data.frame(val), row.names = FALSE)
stopifnot(all(val$abs_diff < 1e-10))
cat("\n`last` reproduces the drop-one R2 loss exactly.\n")
```

```{r}
#| label: relimp-point

ri_raw  <- calc.relimp(m_all, type = c("lmg", "last"), rela = FALSE)
ri_rela <- calc.relimp(m_all, type = c("lmg", "last"), rela = TRUE)

ri_tab <- tibble(
  criterion = crit,
  lmg       = as.numeric(ri_raw$lmg[crit]),
  lmg_pct   = 100 * as.numeric(ri_rela$lmg[crit]),
  last      = as.numeric(ri_raw$last[crit]),
  last_pct  = 100 * as.numeric(ri_rela$last[crit])
) |>
  mutate(shared_share = round(100 * (1 - last / lmg), 1)) |>
  arrange(desc(lmg)) |>
  mutate(across(c(lmg, last), \(x) round(x, 5)),
         across(c(lmg_pct, last_pct), \(x) round(x, 1)))

print(as.data.frame(ri_tab), row.names = FALSE)
cat("\nlmg sums to", round(sum(ri_raw$lmg), 5),
    "= R2 =", round(r2_full, 5), "\n")
cat("shared_share = percentage of a criterion's LMG weight that is NOT unique.\n")
```

## Bootstrap intervals

`rela = FALSE` is used for the bootstrap so the weights stay in variance units
and the intervals are not distorted by the sum-to-100% constraint that
normalization imposes. `diff = TRUE` yields intervals on every pairwise
difference, which is the focal test; `rank = TRUE` yields intervals on each
criterion's rank position.

```{r}
#| label: relimp-boot

set.seed(42)
bb <- boot.relimp(m_all, b = 2000, type = "lmg",
                  rank = TRUE, diff = TRUE, rela = FALSE)
ev <- booteval.relimp(bb, level = 0.95)
```

```{r}
#| label: relimp-boot-table

nm <- ev@namen[-1]
boot_tab <- tibble(
  criterion = nm,
  lmg       = as.numeric(ev@lmg),
  lower     = as.numeric(ev@lmg.lower[1, ]),
  upper     = as.numeric(ev@lmg.upper[1, ])
) |>
  arrange(desc(lmg)) |>
  mutate(across(c(lmg, lower, upper), \(x) round(x, 5)))

print(as.data.frame(boot_tab), row.names = FALSE)
```

```{r}
#| label: relimp-plot
#| fig-height: 3.5
#| fig-width: 7
#| fig-cap: "LMG relative importance with 95% bootstrap intervals (2,000 replicates)"

boot_tab |>
  mutate(criterion = factor(criterion, levels = rev(criterion))) |>
  ggplot(aes(x = lmg, y = criterion)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.18,
                 colour = "grey40") +
  geom_point(size = 2.6) +
  labs(x = expression("LMG share of " * R^2), y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

ggsave("fig/lmg_relative_importance.pdf", width = 7, height = 3.5)
```

## Pairwise differences

relaimpo's own formatted output, which carries the rank letters:

```{r}
#| label: relimp-print
print(ev)
```

```{r}
#| label: relimp-diffs

# The @diffnam slot comes back empty from booteval.relimp, but the difference
# vectors are ordered exactly as combn() enumerates the pairs, so the labels
# are reconstructed rather than read off. The stopifnot guards that assumption.
pair_mat <- combn(nm, 2)
pair_names <- paste0(pair_mat[1, ], " - ", pair_mat[2, ])
stopifnot(length(pair_names) == length(as.numeric(ev@lmg.diff)))

diff_tab <- tibble(
  pair  = pair_names,
  diff  = as.numeric(ev@lmg.diff),
  lower = as.numeric(ev@lmg.diff.lower[1, ]),
  upper = as.numeric(ev@lmg.diff.upper[1, ])
) |>
  mutate(excludes_zero = ifelse(lower > 0 | upper < 0, "yes", "no"),
         across(c(diff, lower, upper), \(x) round(x, 5))) |>
  arrange(desc(abs(diff)))

print(as.data.frame(diff_tab), row.names = FALSE)

cat("\nDifferences whose 95% interval excludes zero:",
    sum(diff_tab$excludes_zero == "yes"), "of", nrow(diff_tab), "\n")
```

```{r}
#| label: relimp-focal

focal <- diff_tab |>
  filter(str_detect(pair, "phraseology"), str_detect(pair, "grammar"))
cat("Focal comparison (phraseology vs grammar):\n")
print(as.data.frame(focal), row.names = FALSE)

cat("\nInterpretation: a negative difference means grammar carries the larger",
    "\nLMG share. The comparison is reliable only if the interval excludes 0.\n")
```

## Rank stability

```{r}
#| label: relimp-ranks

rank_tab <- tibble(
  criterion  = nm,
  rank_est   = as.numeric(ev@lmg.rank),
  rank_lower = as.numeric(ev@lmg.rank.lower[1, ]),
  rank_upper = as.numeric(ev@lmg.rank.upper[1, ])
) |> arrange(rank_est)
stopifnot(nrow(rank_tab) == length(nm))

print(as.data.frame(rank_tab), row.names = FALSE)
cat("\nRank 1 = largest LMG weight. Intervals are over 2,000 bootstrap samples.\n")
```

## Robustness

Point estimates only. The bootstrap above establishes the sampling uncertainty
on the LMG weights; these checks ask whether the *point* estimates move under
two plausible confounds, which does not require re-running 2,000 replicates
twice.

**Essay length.** Predictability is computed per token, but essay length still
varies widely and longer essays earn higher ratings. Entering `n_tokens` as a
seventh predictor lets it take whatever share is really length.

**Prompt.** The network analysis residualizes every variable on writing prompt
before estimation; repeating that here checks that the ordering is not driven by
prompt composition.

```{r}
#| label: robustness

# Length control: n_tokens enters as a 7th predictor (5,040 orderings).
m_len <- lm(reformulate(c(crit, "n_tokens"), "neg_loss"), data = dr)
lmg_len <- calc.relimp(m_len, type = "lmg", rela = FALSE)$lmg

# Prompt residualization: every analysis variable regressed on prompt first.
dr_resid <- dr |>
  mutate(across(all_of(c("neg_loss", crit)),
                \(x) residuals(lm(x ~ factor(d$prompt), na.action = na.exclude))))
m_resid <- lm(reformulate(crit, "neg_loss"), data = dr_resid)
lmg_resid <- calc.relimp(m_resid, type = "lmg", rela = FALSE)$lmg

robust <- tibble(
  criterion    = crit,
  base         = as.numeric(ri_raw$lmg[crit]),
  with_length  = as.numeric(lmg_len[crit]),
  prompt_resid = as.numeric(lmg_resid[crit])
) |>
  arrange(desc(base)) |>
  mutate(across(where(is.numeric), \(x) round(x, 5)))

print(as.data.frame(robust), row.names = FALSE)

cat("\nR2:  base =", round(r2_full, 4),
    " with length =", round(summary(m_len)$r.squared, 4),
    " prompt-residualized =", round(summary(m_resid)$r.squared, 4), "\n")
cat("LMG share taken by n_tokens itself:",
    round(as.numeric(lmg_len[["n_tokens"]]), 5), "\n")
```

# Criterion profile analysis

LMG answers "how much variance". It cannot answer "does predictability behave
like a phraseological index or like a syntactic-elaboration index", because the
MI and L2SCA indices are not in that regression at all — only the human ratings
are.

The profile approach compares *shapes*. Each measured index has a profile: its
Spearman correlation with each of the six criteria. Two indices that tap the
same construct should rise and fall across the criteria together, whatever their
overall magnitude. Family profiles are the mean of the member indices' profiles,
which averages away index-specific noise.

**What a profile correlation does and does not measure.** Correlating two
six-element profiles centres each one first, so it discards *elevation* — how
strongly an index correlates with the ratings on average — and reports only
*pattern*. That is the intended quantity: a weak index and a strong index that
distribute their association across the criteria identically should score 1.0.
Elevation is reported separately below, because it is exactly what the
correlation throws away.

**The n = 6 caveat.** A profile correlation is computed over six points, and
those six criteria are themselves heavily intercorrelated, so the profiles are
compressed into a narrow band and small shifts move the correlation a lot. The
bootstrap resamples essays and recomputes the whole pipeline, so the intervals
carry the sampling uncertainty in the profile estimates; they do **not** turn
n = 6 into a larger effective sample. The intervals are wide by construction and
should be read that way. A jackknife over the six criteria is reported below as
the complement to the bootstrap: it asks whether any one criterion carries a
similarity on its own, which resampling essays cannot detect.

**Low-elevation profiles are the fragile case.** Centring rescales every profile
to the same footing, so an index whose correlations with all six criteria sit
near zero contributes its *shape* at full weight even though that shape is a
small difference between small numbers. The elaboration indices are in this
regime. The per-index table at the end of the section is the check: a family
result that survives only through one near-null member should not be reported as
a family result.

```{r}
#| label: profiles

# One correlation matrix per resample; every profile is read off it.
profile_stat <- function(dd, i) {
  R <- cor(dd[i, c("neg_loss", crit, mi_vars, syn_vars)])
  loss <- R[crit, "neg_loss"]
  phr  <- rowMeans(R[crit, mi_vars,  drop = FALSE])
  syn  <- rowMeans(R[crit, syn_vars, drop = FALSE])
  c(loss, phr, syn,
    sim_phr     = cor(loss, phr),
    sim_syn     = cor(loss, syn),
    sim_phr_syn = cor(phr,  syn),
    sim_diff    = cor(loss, phr) - cor(loss, syn))
}

est_p <- profile_stat(dr, seq_len(nrow(dr)))

prof <- tibble(
  criterion   = crit,
  loss        = est_p[1:6],
  phraseology = est_p[7:12],
  elaboration = est_p[13:18]
)

cat("Profiles (Spearman correlation with each criterion):\n")
print(as.data.frame(prof |> mutate(across(-criterion, \(x) round(x, 3)))),
      row.names = FALSE)

cat("\nElevation (mean correlation across the six criteria):\n")
print(round(colMeans(prof[-1]), 3))

cat("\nProfile similarity:\n")
print(round(est_p[c("sim_phr", "sim_syn", "sim_phr_syn", "sim_diff")], 3))
```

```{r}
#| label: profile-boot

b_p <- boot(dr, profile_stat, R = 2000)

sim_idx <- match(c("sim_phr", "sim_syn", "sim_phr_syn", "sim_diff"),
                 names(est_p))
sim_tab <- map_dfr(sim_idx, \(k) {
  ci <- boot.ci(b_p, type = "perc", index = k)$percent[4:5]
  tibble(quantity = names(est_p)[k], estimate = est_p[k],
         lower = ci[1], upper = ci[2], boot_se = sd(b_p$t[, k]))
}) |>
  mutate(across(c(estimate, lower, upper, boot_se), \(x) round(x, 3)))

print(as.data.frame(sim_tab), row.names = FALSE)

cat("\nsim_diff = sim_phr - sim_syn. Positive means predictability's profile",
    "\nresembles the phraseological family more closely than the elaboration",
    "\nfamily. Proportion of replicates with sim_diff <= 0:",
    round(mean(b_p$t[, sim_idx[4]] <= 0), 4), "\n")
```

```{r}
#| label: profile-plot
#| fig-height: 4.2
#| fig-width: 8
#| fig-cap: "Criterion profiles for word predictability, the phraseological (MI) family and the syntactic-elaboration family. Left: raw Spearman correlations. Right: each profile mean-centred, which is the scale the profile-similarity correlations are computed on."

prof_long <- prof |>
  pivot_longer(-criterion, names_to = "profile", values_to = "rho") |>
  group_by(profile) |>
  mutate(centred = rho - mean(rho)) |>
  ungroup() |>
  pivot_longer(c(rho, centred), names_to = "scale", values_to = "value") |>
  mutate(
    criterion = factor(criterion,
                       levels = prof$criterion[order(prof$loss)]),
    profile   = factor(profile,
                       levels = c("loss", "phraseology", "elaboration"),
                       labels = c("predictability", "phraseology (MI)",
                                  "elaboration (L2SCA)")),
    scale     = factor(scale, levels = c("rho", "centred"),
                       labels = c("raw correlation", "mean-centred"))
  )

ggplot(prof_long, aes(x = value, y = criterion,
                      colour = profile, shape = profile)) +
  geom_vline(xintercept = 0, colour = "grey85", linewidth = 0.4) +
  geom_point(size = 2.8, alpha = 0.9) +
  facet_wrap(~ scale, scales = "free_x") +
  scale_colour_manual(values = c("predictability"      = "#1b4965",
                                 "phraseology (MI)"    = "#c1666b",
                                 "elaboration (L2SCA)" = "#5e8c61")) +
  scale_shape_manual(values = c(16, 17, 15)) +
  labs(x = "Spearman correlation with criterion", y = NULL,
       colour = NULL, shape = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        legend.position = "bottom",
        strip.text = element_text(face = "bold"))

ggsave("fig/criterion_profiles.pdf", width = 8, height = 4.2)
```

```{r}
#| label: profile-jackknife

# With six points, one criterion can carry a profile correlation on its own.
# Recomputing each similarity with each criterion held out shows whether it
# does. The bootstrap above cannot see this: it resamples essays, not criteria.
jack <- map_dfr(crit, \(s) {
  k <- which(crit != s)
  tibble(dropped = s,
         sim_phr = cor(prof$loss[k], prof$phraseology[k]),
         sim_syn = cor(prof$loss[k], prof$elaboration[k]))
}) |>
  mutate(sim_diff = sim_phr - sim_syn,
         across(where(is.numeric), \(x) round(x, 3)))

print(as.data.frame(jack), row.names = FALSE)
cat("\nFull-profile values for comparison: sim_phr =",
    round(est_p[["sim_phr"]], 3), " sim_syn =", round(est_p[["sim_syn"]], 3),
    " sim_diff =", round(est_p[["sim_diff"]], 3), "\n")
```

```{r}
#| label: profile-members

# A family mean can hide disagreement among its members. The per-index profiles
# show whether the family is a coherent thing to average in the first place.
member <- cor(dr[c(mi_vars, syn_vars)], dr[crit])
cat("Per-index profiles:\n")
print(round(member, 3))

cat("\nWithin-family profile agreement (mean pairwise correlation):\n")
agree <- \(vs) mean(cor(t(member[vs, , drop = FALSE]))[lower.tri(diag(length(vs)))])
cat("  phraseology (MI):    ", round(agree(mi_vars), 3), "\n")
cat("  elaboration (L2SCA): ", round(agree(syn_vars), 3), "\n")

cat("\nSimilarity of each index's profile to the predictability profile:\n")
print(round(sort(apply(member, 1, \(p) cor(p, est_p[1:6])), decreasing = TRUE), 3))
```

# Summary

```{r}
#| label: summary
#| output: asis

cat(sprintf(
"- The six criteria jointly explain **%.1f%%** of the rank variance in word
  predictability.
- The largest LMG share is **%s** (%.4f, 95%% CI [%.4f, %.4f]); **%d of %d**
  pairwise differences have intervals excluding zero.
- Unique variance is small throughout: the median criterion holds **%.0f%%** of
  its LMG weight in variance shared with the other ratings.
- Predictability's criterion profile correlates **%.2f** with the phraseological
  family and **%.2f** with the elaboration family (difference %.2f, 95%% CI
  [%.2f, %.2f]).
- Under the criterion jackknife the negative similarity holds (%.2f to %.2f)
  but the positive one does not: dropping **%s** moves it to **%.2f**.
",
100 * r2_full,
boot_tab$criterion[1], boot_tab$lmg[1], boot_tab$lower[1], boot_tab$upper[1],
sum(diff_tab$excludes_zero == "yes"), nrow(diff_tab),
median(ri_tab$shared_share),
est_p[["sim_phr"]], est_p[["sim_syn"]], est_p[["sim_diff"]],
sim_tab$lower[sim_tab$quantity == "sim_diff"],
sim_tab$upper[sim_tab$quantity == "sim_diff"],
max(jack$sim_syn), min(jack$sim_syn),
jack$dropped[which.min(jack$sim_phr)], min(jack$sim_phr)
))
```

# Interpretation

**Read the LMG weights, not the unique components.** The criteria are rated by
the same humans on the same essays and carry a large halo — the mean
intercorrelation is reported above. Under that much redundancy the squared
semipartials shrink toward zero for every criterion, and shrink most for the
ones most entangled with the rest, so a ranking by unique variance measures
distinctiveness rather than importance. The `shared_share` column makes the
problem concrete: for the criteria with the smallest unique components, nearly
all of the LMG weight is variance held in common. LMG allocates that shared
variance instead of discarding it, which is why it is the reported quantity.

**Among the ratings, the directional hypothesis is not supported.** Grammar takes
the largest LMG share, ranks first in every one of the 2,000 bootstrap samples,
and beats phraseology by an interval that excludes zero. Neither the length
control nor prompt residualization disturbs the ordering. If predictability were
distinctively phraseological *as raters perceive it*, this is where it would
show, and it does not. The one ambiguity is phraseology versus conventions,
which is the single pair of fifteen whose interval covers zero — a criterion
about spelling and punctuation is holding its own against a criterion about
formulaic language, which is worth pre-empting in the write-up rather than
leaving for a reader to notice.

**The pattern question gives a different answer, but only half of it is robust.**
LMG asks which *ratings* predictability tracks; the profile analysis asks which
*family of measured indices* it resembles in shape. Taken at face value the
second answer flips sign: predictability's profile correlates +0.58 with the
phraseological family and −0.62 with the elaboration family, a gap of 1.20 that
no bootstrap replicate contradicts. The jackknife does not let both halves
stand.

- **The negative half is robust.** Dropping any single criterion leaves
  `sim_syn` between −0.48 and −0.74, and the per-index table shows it is not one
  index's doing either: MLC (−0.67) and CN/C (−0.95) point that way
  independently, with CP/C near zero. Predictability's profile is genuinely
  *unlike* the elaboration family's.
- **The positive half is not.** Dropping cohesion sends `sim_phr` from +0.58 to
  −0.08. The resemblance to the MI family rests almost entirely on cohesion
  being the low point of both profiles; over the other five criteria the two
  profiles do not agree, and they disagree most on vocabulary, which is near the
  top of the MI profile and near the bottom of predictability's. Report this as
  "predictability and the MI indices are both least associated with cohesion",
  not as a general profile match.

`sim_diff` stays positive under every jackknife (0.66 to 1.34), so the *ordering*
survives — predictability is closer to the MI family than to the elaboration
family. But that ordering is carried by the elaboration term, not by a positive
match to MI. The dissimilarity is the finding; the similarity is one shared low
point.

What predictability shares with the *ratings*, meanwhile, is mostly general
proficiency: the median criterion holds around 90% of its LMG weight in variance
common to the others, and grammar leads because it is the rating closest to that
general factor. Elevation and pattern answer different questions, and only the
pattern comparison is free of the halo, because centring removes exactly the
elevation the halo inflates.

**Read alongside the network analysis, this is weak convergence at best.** The
network placed `mean_loss` in a community with the dependency-MI indices. The
profile analysis is consistent with that but does not independently corroborate
it, because the part of the profile result that would corroborate it — a
positive match to the MI family — is the part that fails the jackknife. What the
profile analysis contributes on its own evidence is the *exclusion*: whatever
predictability is tracking, it is not syntactic elaboration. The LMG result
neither supports nor contradicts the network finding, because its predictors are
*human ratings* that happen to be named Phraseology and Grammar — a rater's
holistic phraseology judgment and a pointwise MI statistic are not the same
construct. The defensible summary is that surprisal is clearly dissociated from
syntactic elaboration, plausibly but not decisively aligned with corpus-derived
formulaicity, and at the level of human ratings tracks general proficiency with
a grammatical edge.

# References

Davison, M. L., & Davenport, E. C. (2002). Identifying criterion-related
patterns of predictor scores using multiple regression. *Psychological Methods,
7*(4), 468–484.

Grömping, U. (2006). Relative importance for linear regression in R: The package
relaimpo. *Journal of Statistical Software, 17*(1), 1–27.

Lindeman, R. H., Merenda, P. F., & Gold, R. Z. (1980). *Introduction to
bivariate and multivariate analysis.* Scott, Foresman.

Lu, X. (2010). Automatic analysis of syntactic complexity in second language
writing. *International Journal of Corpus Linguistics, 15*(4), 474–496.
