---
title: "Relative Importance of the ELLIPSE Criteria"
subtitle: "Which ratings does word predictability track?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

The network analysis puts `mean_loss` in a community with the dependency-MI
features. This notebook asks a different question, about the human ratings
rather than the computed indices. Across the six ELLIPSE analytic criteria,
which ones does word predictability track, and by how much?

All six criteria go into one regression, and LMG splits the resulting $R^2$ into
a non-negative share for each criterion.

NOTE: A rater's holistic `Phraseology` judgment and a
corpus-derived MI index are different operationalizations, so nothing here
adjudicates the network finding. The Interpretation section comes back to that.

**Sign convention.** Everything uses `neg_loss = -mean_loss`, so higher values
mean more predictable text and every coefficient reads in the same direction as
the proficiency scores.

```{r}
#| label: setup
library(tidyverse)
library(relaimpo)

set.seed(42)
```

```{r}
#| label: load-data

d_raw <- read_csv("data/ellipse/ellipse_metrics.csv",
                  show_col_types = FALSE)

# The six analytic criteria, in the order ELLIPSE reports them.
crit <- c("cohesion", "syntax", "vocabulary",
          "phraseology", "grammar", "conventions")

d <- d_raw |>
  transmute(
    text_id_kaggle,
    prompt,
    neg_loss    = -mean_loss,
    cohesion    = Cohesion,
    syntax      = Syntax,
    vocabulary  = Vocabulary,
    phraseology = Phraseology,
    grammar     = Grammar,
    conventions = Conventions,
    n_tokens    = n_tokens
  )

cat("Rows loaded:", nrow(d), "\n")
na_counts <- colSums(is.na(d))
if (any(na_counts > 0)) {
  cat("Missing values:\n"); print(na_counts[na_counts > 0])
}

d <- drop_na(d)
cat("Complete cases used:", nrow(d), "\n\n")

cat("Rating scales (ELLIPSE ratings are ordinal, 1-5 in 0.5 steps):\n")
print(map_dfr(d[crit],
              \(x) tibble(min = min(x), max = max(x),
                          levels = n_distinct(x), sd = sd(x)),
              .id = "variable"))
```

## Rank transformation

Everything below runs on **ranks**. The ratings have about nine distinct levels
on a bounded scale, so Pearson correlations of the raw scores aren't the right
summary. Ranking first makes every correlation a Spearman correlation and every
simple $R^2$ a squared Spearman correlation.

Ranks get computed once, on the full sample, and the bootstrap then resamples
rows of the ranked data. That treats the ranks as the data rather than
re-ranking inside each resample.

```{r}
#| label: rank-transform

vars <- c("neg_loss", crit, "n_tokens")
dr <- d |> mutate(across(all_of(vars), \(x) rank(x)))

# Pearson on average ranks is Spearman on the raw values, exactly. Verified
# rather than asserted, since the whole notebook rests on it.
stopifnot(all(abs(
  cor(dr$neg_loss, dr[crit]) -
  cor(d$neg_loss, d[crit], method = "spearman")
) < 1e-12))
cat("Rank-Pearson reproduces Spearman to within 1e-12.\n")
```

## Zero-order correlations

Reported first for context, and because the criterion intercorrelations represent
the halo estimate. They set an upper bound on how much unique variance any one
rating can possibly have.

```{r}
#| label: zero-order

cat("Spearman correlation of neg_loss with each criterion:\n")
print(round(sort(cor(dr$neg_loss, dr[crit])[1, ], decreasing = TRUE), 3))

cat("\nCriterion intercorrelations:\n")
print(round(cor(dr[crit]), 3))

cat("\nMean off-diagonal criterion intercorrelation:",
    round(mean(cor(dr[crit])[lower.tri(diag(6))]), 3), "\n")
```

# Relative importance across the six criteria

The criteria are heavily intercorrelated, so the unique variance of each one
(its squared semipartial) understates every predictor, and understates the most
redundant ones worst. Ranking criteria by unique variance ends up rewarding
whichever rating happens to be least redundant with the others, which isn't the
same thing as importance.

**LMG** (Lindeman, Merenda & Gold, 1980), the Shapley-value decomposition,
averages each predictor's sequential $R^2$ contribution over all $6! = 720$
orderings of entry, so shared variance gets allocated among the predictors.
The weights are non-negative and sum exactly to $R^2$, so they read as
"this criterion's share of the explained variance".

Quantities reported:

- `lmg` is the Shapley share, with shared variance allocated. Mainly reported
  for congruence with the bootstrap intervals, which are in variance units.
- `lmg_pct` is the same share expressed as a percentage of $R^2$.
- `last` is the variance a criterion adds when it enters last. That's exactly
  the squared semipartial, the criterion's unique contribution.
- `last_pct` is the same unique contribution expressed as a percentage of $R^2$.
- `shared_share` is the percentage of a criterion's LMG weight that is NOT unique.

A criterion with a large `lmg` and a near-zero `last` is contributing almost
entirely through variance it shares with the other ratings.

```{r}
#| label: relimp-setup

m_all <- lm(reformulate(crit, "neg_loss"), data = dr)
r2_full <- summary(m_all)$r.squared
cat("Full model R2 =", round(r2_full, 5), "\n")
```

```{r}
#| label: relimp-validate

drop_last <- map_dbl(crit, \(s)
  r2_full - summary(lm(reformulate(setdiff(crit, s), "neg_loss"), dr))$r.squared)
last_ri <- calc.relimp(m_all, type = "last", rela = FALSE)$last[crit]

val <- tibble(criterion     = crit,
              relaimpo_last = as.numeric(last_ri),
              by_refit      = drop_last) |>
  mutate(abs_diff = abs(relaimpo_last - by_refit))

print(as.data.frame(val), row.names = FALSE)
stopifnot(all(val$abs_diff < 1e-10))
cat("\n`last` reproduces the drop-one R2 loss exactly.\n")
```

```{r}
#| label: relimp-point

ri_raw  <- calc.relimp(m_all, type = c("lmg", "last"), rela = FALSE)
ri_rela <- calc.relimp(m_all, type = c("lmg", "last"), rela = TRUE)

ri_tab <- tibble(
  criterion = crit,
  lmg       = as.numeric(ri_raw$lmg[crit]),
  lmg_pct   = 100 * as.numeric(ri_rela$lmg[crit]),
  last      = as.numeric(ri_raw$last[crit]),
  last_pct  = 100 * as.numeric(ri_rela$last[crit])
) |>
  mutate(shared_share = round(100 * (1 - last / lmg), 1)) |>
  arrange(desc(lmg)) |>
  mutate(across(c(lmg, last), \(x) round(x, 5)),
         across(c(lmg_pct, last_pct), \(x) round(x, 1)))

print(as.data.frame(ri_tab), row.names = FALSE)
cat("\nlmg sums to", round(sum(ri_raw$lmg), 5),
    "= R2 =", round(r2_full, 5), "\n")
cat("shared_share = percentage of a criterion's LMG weight that is NOT unique.\n")
```

## Bootstrap intervals

`rela = FALSE` keeps the weights in variance units, so the intervals aren't
distorted by the sum-to-100% constraint that normalization imposes.
`diff = TRUE` gives intervals on every pairwise difference, which is the focal
test, and `rank = TRUE` gives intervals on each criterion's rank position.

```{r}
#| label: relimp-boot

set.seed(42)
bb <- boot.relimp(m_all, b = 2000, type = "lmg",
                  rank = TRUE, diff = TRUE, rela = FALSE)
ev <- booteval.relimp(bb, level = 0.95)
```

```{r}
#| label: relimp-boot-table

nm <- ev@namen[-1]
boot_tab <- tibble(
  criterion = nm,
  lmg       = as.numeric(ev@lmg),
  lower     = as.numeric(ev@lmg.lower[1, ]),
  upper     = as.numeric(ev@lmg.upper[1, ])
) |>
  arrange(desc(lmg)) |>
  mutate(across(c(lmg, lower, upper), \(x) round(x, 5)))

print(as.data.frame(boot_tab), row.names = FALSE)
```

```{r}
#| label: relimp-plot
#| fig-height: 3.5
#| fig-width: 7
#| fig-cap: "LMG relative importance with 95% bootstrap intervals (2,000 replicates)"

boot_tab |>
  mutate(criterion = factor(criterion, levels = rev(criterion))) |>
  ggplot(aes(x = lmg, y = criterion)) +
  geom_errorbarh(aes(xmin = lower, xmax = upper), height = 0.18,
                 colour = "grey40") +
  geom_point(size = 2.6) +
  labs(x = expression("LMG share of " * R^2), y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

ggsave("fig/lmg_relative_importance.pdf", width = 7, height = 3.5)
```

## Pairwise differences

relaimpo's own formatted output, which carries the rank letters.

```{r}
#| label: relimp-print
print(ev)
```

```{r}
#| label: relimp-diffs

# The @diffnam slot comes back empty from booteval.relimp, but the difference
# vectors are ordered exactly as combn() enumerates the pairs, so the labels
# are reconstructed rather than read off.
pair_mat <- combn(nm, 2)
pair_names <- paste0(pair_mat[1, ], " - ", pair_mat[2, ])
stopifnot(length(pair_names) == length(as.numeric(ev@lmg.diff)))

diff_tab <- tibble(
  pair  = pair_names,
  diff  = as.numeric(ev@lmg.diff),
  lower = as.numeric(ev@lmg.diff.lower[1, ]),
  upper = as.numeric(ev@lmg.diff.upper[1, ])
) |>
  mutate(excludes_zero = ifelse(lower > 0 | upper < 0, "yes", "no"),
         across(c(diff, lower, upper), \(x) round(x, 5))) |>
  arrange(desc(abs(diff)))

print(as.data.frame(diff_tab), row.names = FALSE)

cat("\nDifferences whose 95% interval excludes zero:",
    sum(diff_tab$excludes_zero == "yes"), "of", nrow(diff_tab), "\n")
```

```{r}
#| label: relimp-focal

focal <- diff_tab |>
  filter(str_detect(pair, "phraseology"), str_detect(pair, "grammar"))
cat("Focal comparison (phraseology vs grammar):\n")
print(as.data.frame(focal), row.names = FALSE)

cat("\nA negative difference means grammar carries the larger LMG share.",
    "\nThe comparison is only reliable if the interval excludes 0.\n")
```

## Rank stability

```{r}
#| label: relimp-ranks

rank_tab <- tibble(
  criterion  = nm,
  rank_est   = as.numeric(ev@lmg.rank),
  rank_lower = as.numeric(ev@lmg.rank.lower[1, ]),
  rank_upper = as.numeric(ev@lmg.rank.upper[1, ])
) |> arrange(rank_est)
stopifnot(nrow(rank_tab) == length(nm))

print(as.data.frame(rank_tab), row.names = FALSE)
cat("\nRank 1 = largest LMG weight. Intervals are over 2,000 bootstrap samples.\n")
```

## Robustness

Point estimates only. The bootstrap above already establishes sampling
uncertainty on the weights, and these two checks only ask whether the point
estimates move under a couple of plausible confounds, which doesn't need 2,000
replicates run twice.

**Essay length.** Predictability is computed per token, but length still varies
a lot and longer essays earn higher ratings. Putting `n_tokens` in as a seventh
predictor lets it take whatever share is really length.

**Prompt.** The network analysis residualizes every variable on writing prompt
before estimation. Doing the same here checks that the ordering isn't driven by
prompt composition.

```{r}
#| label: robustness

# Length control: n_tokens enters as a 7th predictor (5,040 orderings).
m_len <- lm(reformulate(c(crit, "n_tokens"), "neg_loss"), data = dr)
lmg_len <- calc.relimp(m_len, type = "lmg", rela = FALSE)$lmg

# Prompt residualization: every analysis variable regressed on prompt first.
dr_resid <- dr |>
  mutate(across(all_of(c("neg_loss", crit)),
                \(x) residuals(lm(x ~ factor(d$prompt), na.action = na.exclude))))
m_resid <- lm(reformulate(crit, "neg_loss"), data = dr_resid)
lmg_resid <- calc.relimp(m_resid, type = "lmg", rela = FALSE)$lmg

robust <- tibble(
  criterion    = crit,
  base         = as.numeric(ri_raw$lmg[crit]),
  with_length  = as.numeric(lmg_len[crit]),
  prompt_resid = as.numeric(lmg_resid[crit])
) |>
  arrange(desc(base)) |>
  mutate(across(where(is.numeric), \(x) round(x, 5)))

print(as.data.frame(robust), row.names = FALSE)

cat("\nR2:  base =", round(r2_full, 4),
    " with length =", round(summary(m_len)$r.squared, 4),
    " prompt-residualized =", round(summary(m_resid)$r.squared, 4), "\n")
cat("LMG share taken by n_tokens itself:",
    round(as.numeric(lmg_len[["n_tokens"]]), 5), "\n")
```

# Summary

```{r}
#| label: summary
#| output: asis

cat(sprintf(
"- The six criteria jointly explain **%.1f%%** of the rank variance in word
  predictability.
- The largest LMG share is **%s** (%.4f, 95%% CI [%.4f, %.4f]), and **%d of %d**
  pairwise differences have intervals excluding zero.
- Unique variance is small throughout. The median criterion holds **%.0f%%** of
  its LMG weight in variance shared with the other ratings.
- Neither the length control nor prompt residualization changes the ordering,
  and `n_tokens` takes only %.4f of the variance on its own.
",
100 * r2_full,
boot_tab$criterion[1], boot_tab$lmg[1], boot_tab$lower[1], boot_tab$upper[1],
sum(diff_tab$excludes_zero == "yes"), nrow(diff_tab),
median(ri_tab$shared_share),
as.numeric(lmg_len[["n_tokens"]])
))
```

# Interpretation

**The directional hypothesis doesn't hold up.** Grammar takes the largest LMG
share, ranks first in all 2,000 bootstrap samples, and beats phraseology by an
interval that excludes zero. Neither the length control nor prompt
residualization disturbs the ordering. If predictability were distinctively
phraseological as raters perceive it, this is where it would show, and it
doesn't.

**Phraseology vs. Conventions** Conventions carries the larger point estimate
of the two, .091 against .083, and their difference is one of three among the
fifteen whose interval covers zero. A criterion about spelling and punctuation
is holding its own against a criterion about formulaic language, and on the
point estimate it is slightly ahead. The reading is that conventions is picking
up mechanical accuracy that the language model also finds predictable.

**None of this adjudicates the network analysis.** The network placed
`mean_loss` in a community with the dependency-MI indices (`amod_mi`,
`dobj_mi`, `advmod_mi`), which are corpus-derived association strengths computed
against the Dolma reference corpus. The predictors here are human ratings that
happen to be named Phraseology and Grammar. A rater's holistic phraseology
judgment and a pointwise MI statistic aren't the same construct, and the rating
carries a halo the MI indices don't. The summary is that surprisal
aligns with corpus-derived formulaicity at the level of measured indices of 
Sophistication and Complexity, while at the level of human ratings that include
Accuracy measures, it tracks general proficiency with an accuracy edge (both
grammatical accuracy and mechanical accuracy).

# References

Grömping, U. (2006). Relative importance for linear regression in R: The package
relaimpo. *Journal of Statistical Software, 17*(1), 1–27.

Lindeman, R. H., Merenda, P. F., & Gold, R. Z. (1980). *Introduction to
bivariate and multivariate analysis.* Scott, Foresman.
