---
title: "Study 1 — Fairness–Validity Pareto Front"
subtitle: "Selecting an LLM-Surprisal Configuration that Balances Concurrent Validity and Cross-L1 Fairness"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

For every configuration in the Study 1 matrix — **11 language models × 3 context
windows (8 / 64 / full)** — we place each on a two-axis plane and find the
non-dominated (Pareto) set:

- **Validity (y):** strength of the concurrent correlation between mean token
  surprisal and human holistic proficiency, measured on **ELLIPSE** as
  `|Pearson r|`. (All correlations are negative — more proficient writers produce
  more predictable text — so validity is the *magnitude* `|r|`; higher is better.)
- **Fairness (x):** worst-case cross-L1 score displacement on **TOEFL 11**,
  measured as **max |Cohen's d|** across the 11 L1 groups, computed within
  proficiency strata (DIF logic). Lower is better.

A configuration is **Pareto-optimal** if no other configuration is simultaneously
*more valid* (higher `|r|`) **and** *fairer* (lower max `|d|`). The default for
Studies 2–3 is the **highest-`|r|` point on the front among the pure-pretrained
models** (instruct models are a contrast condition, not eligible defaults).

Inputs are the two long tables written by `assemble_analysis_data.py`
(`results/predictability/{ellipse,toefl}_long.csv.gz`).

# Section 1 — Setup & Data

```{r}
#| label: setup
# Individual tidyverse components (the `tidyverse` meta-package pulls xml2/rvest,
# which need system libxml2 and aren't used here).
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggrepel)

set.seed(42)          # fixed bootstrap seed (reproducibility)
N_BOOT   <- 2000      # bootstrap iterations (plan §3.3)
D_FLAG   <- 0.10      # Williamson et al. (2012) fairness flag threshold

# Locate results/predictability by walking up from the working dir (Quarto runs
# from the .qmd's directory), so this works whether rendered or run standalone.
find_results <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    cand <- file.path(d, "results", "predictability")
    if (dir.exists(cand)) return(cand)
    d <- dirname(d)
  }
  stop("results/predictability not found — run assemble_analysis_data.py first")
}
RESULTS <- find_results()
```

```{r}
#| label: model-meta
# Architecture / post-training status per model (from models.py registry).
model_meta <- tribble(
  ~model,                  ~arch,     ~instruct,
  "bert-base",             "masked",  FALSE,
  "modernbert-base",       "masked",  FALSE,
  "modernbert-large",      "masked",  FALSE,
  "gpt2",                  "causal",  FALSE,
  "gpt2-xl",               "causal",  FALSE,
  "olmo2-1b",              "causal",  FALSE,
  "olmo2-7b",              "causal",  FALSE,
  "qwen2.5-7b",            "causal",  FALSE,
  "qwen2.5-7b-instruct",   "causal",  TRUE,
  "llama3.1-8b",           "causal",  FALSE,
  "llama3.1-8b-instruct",  "causal",  TRUE,
)
WINDOWS <- c("8", "64", "full")
```

```{r}
#| label: load-data
ell <- read_csv(file.path(RESULTS, "ellipse_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

toe <- read_csv(file.path(RESULTS, "toefl_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

# ELLIPSE as an [essay x config] surprisal matrix + aligned criterion vector,
# so every configuration shares the same essays (dependent, essay-level resampling).
ell_wide <- ell |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(ell, text_id, overall), by = "text_id")

config_cols <- setdiff(names(ell_wide), c("text_id", "overall"))
S_ell <- as.matrix(ell_wide[, config_cols])   # essays x configs
y_ell <- ell_wide$overall

# TOEFL as an [essay x config] surprisal matrix + L1 / proficiency-level vectors.
toe_wide <- toe |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(toe, text_id, l1, level), by = "text_id")

stopifnot(identical(config_cols, setdiff(names(toe_wide), c("text_id", "l1", "level"))))
M_toe <- as.matrix(toe_wide[, config_cols])   # essays x configs
l1_toe    <- toe_wide$l1
level_toe <- toe_wide$level

cat(sprintf("configs: %d | ELLIPSE essays: %d | TOEFL essays: %d\n",
            length(config_cols), nrow(S_ell), nrow(M_toe)))
```

# Section 2 — Validity (ELLIPSE)

Per configuration, the Pearson correlation between mean surprisal and the
holistic `overall` score, reported as the magnitude `|r|`.

```{r}
#| label: validity
# cor(S, y) returns one correlation per column of S -> vector over configs.
r_hat   <- cor(S_ell, y_ell)[, 1]
validity <- tibble(config = names(r_hat), r = r_hat, abs_r = abs(r_hat))
```

# Section 3 — Fairness (TOEFL 11)

For each configuration and each focal L1 group *g*: within each proficiency
stratum *s* (low / medium / high), Cohen's *d* in surprisal between *g* and all
remaining learners, standardized by the **pooled within-stratum SD**; the three
stratum *d*'s are combined into a focal-group **weighted average** (weighted by
*g*'s stratum sizes). The configuration's fairness score is **max |d|** across
the 11 groups (worst-case). We also count how many groups exceed the |d| ≥ 0.10
flag.

```{r}
#| label: fairness-fns
# Fast per-column variance (avoids apply): var of each column of X.
col_var <- function(X) {
  n <- nrow(X); m <- colMeans(X)
  (colSums(X * X) - n * m * m) / (n - 1)
}

# Weighted-average Cohen's d per (group) for every config column of M.
# Returns a [group x config] matrix of weighted-average d.
group_d_matrix <- function(M, l1, level,
                            groups = sort(unique(l1)),
                            strata = c("low", "medium", "high")) {
  d_wavg <- matrix(NA_real_, nrow = length(groups), ncol = ncol(M),
                   dimnames = list(groups, colnames(M)))
  for (g in groups) {
    num  <- rep(0, ncol(M)); wsum <- 0
    for (s in strata) {
      in_s  <- level == s
      foc   <- in_s & (l1 == g)
      rst   <- in_s & (l1 != g)
      n1 <- sum(foc); n2 <- sum(rst)
      if (n1 < 2L || n2 < 2L) next
      Mf <- M[foc, , drop = FALSE]; Mr <- M[rst, , drop = FALSE]
      m1 <- colMeans(Mf); m2 <- colMeans(Mr)
      v1 <- col_var(Mf);  v2 <- col_var(Mr)
      sp <- sqrt(((n1 - 1) * v1 + (n2 - 1) * v2) / (n1 + n2 - 2))
      d  <- (m1 - m2) / sp
      num  <- num + n1 * d          # weight stratum by focal-group size
      wsum <- wsum + n1
    }
    d_wavg[g, ] <- num / wsum
  }
  d_wavg
}

# Fairness summaries per config: max|d|, the worst L1, and #groups over the flag.
fairness_summary <- function(M, l1, level) {
  D <- group_d_matrix(M, l1, level)
  absD <- abs(D)
  tibble(
    config    = colnames(M),
    max_abs_d = apply(absD, 2, max, na.rm = TRUE),
    worst_l1  = rownames(D)[apply(absD, 2, which.max)],
    n_flag    = colSums(absD >= D_FLAG, na.rm = TRUE)
  )
}
```

```{r}
#| label: fairness
fairness <- fairness_summary(M_toe, l1_toe, level_toe)
```

# Section 4 — Pareto Front

```{r}
#| label: assemble-front
is_front <- function(abs_r, max_abs_d) {
  # config i is dominated if some j is >= on validity AND <= on fairness,
  # with at least one strict.
  n <- length(abs_r); keep <- rep(TRUE, n)
  for (i in seq_len(n)) {
    dom <- (abs_r >= abs_r[i]) & (max_abs_d <= max_abs_d[i]) &
           ((abs_r > abs_r[i]) | (max_abs_d < max_abs_d[i]))
    if (any(dom)) keep[i] <- FALSE
  }
  keep
}

points <- validity |>
  left_join(fairness, by = "config") |>
  separate(config, into = c("model", "window"), sep = "/", remove = FALSE) |>
  mutate(window = factor(window, levels = WINDOWS)) |>
  left_join(model_meta, by = "model") |>
  mutate(front = is_front(abs_r, max_abs_d))

points |>
  arrange(desc(front), desc(abs_r)) |>
  select(model, window, arch, instruct, r, abs_r, max_abs_d, worst_l1, n_flag, front) |>
  mutate(across(c(r, abs_r, max_abs_d), \(x) round(x, 3))) |>
  knitr::kable(caption = "All 33 configurations (front first, then by validity).")
```

## Bootstrap confidence regions

2,000 iterations. ELLIPSE essays are resampled with replacement for `|r|`; TOEFL
essays are resampled **stratified by proficiency level** for max `|d|` (the two
corpora are independent samples, so they are resampled independently). We report
per-axis 95% percentile intervals and, for each configuration, its **bootstrap
front-membership probability** (share of iterations in which it is non-dominated).

```{r}
#| label: bootstrap
n_ell   <- nrow(S_ell)
lvl_idx <- split(seq_len(nrow(M_toe)), level_toe)   # row indices per stratum

boot_absr <- matrix(NA_real_, N_BOOT, ncol(S_ell), dimnames = list(NULL, colnames(S_ell)))
boot_maxd <- matrix(NA_real_, N_BOOT, ncol(M_toe), dimnames = list(NULL, colnames(M_toe)))
boot_front <- matrix(FALSE,   N_BOOT, ncol(S_ell), dimnames = list(NULL, colnames(S_ell)))

for (b in seq_len(N_BOOT)) {
  # validity: essay-level resample shared across configs
  ei <- sample.int(n_ell, n_ell, replace = TRUE)
  ab <- abs(cor(S_ell[ei, ], y_ell[ei])[, 1])

  # fairness: stratified resample
  ti <- unlist(lapply(lvl_idx, \(ix) sample(ix, length(ix), replace = TRUE)),
               use.names = FALSE)
  D  <- group_d_matrix(M_toe[ti, ], l1_toe[ti], level_toe[ti])
  md <- apply(abs(D), 2, max, na.rm = TRUE)

  boot_absr[b, ] <- ab
  boot_maxd[b, ] <- md
  boot_front[b, ] <- is_front(ab, md)
}

q025 <- \(x) quantile(x, 0.025, names = FALSE)
q975 <- \(x) quantile(x, 0.975, names = FALSE)

boot_ci <- tibble(
  config      = colnames(S_ell),
  absr_lo     = apply(boot_absr, 2, q025),
  absr_hi     = apply(boot_absr, 2, q975),
  maxd_lo     = apply(boot_maxd, 2, q025),
  maxd_hi     = apply(boot_maxd, 2, q975),
  front_prob  = colMeans(boot_front)
)

points <- left_join(points, boot_ci, by = "config")
```

## Selected configuration

```{r}
#| label: selection
front_tbl <- points |> filter(front) |> arrange(desc(abs_r))

selected <- front_tbl |> filter(!instruct) |> slice(1)

front_tbl |>
  select(model, window, arch, instruct, abs_r, max_abs_d, worst_l1, n_flag, front_prob) |>
  mutate(across(c(abs_r, max_abs_d, front_prob), \(x) round(x, 3))) |>
  knitr::kable(caption = "Pareto front (non-dominated configurations).")

cat(sprintf(
  "\nSelected default (highest |r| pure-pretrained on the front):\n  %s / window %s  |  |r| = %.3f  |  max|d| = %.3f  |  worst L1 = %s  |  front prob = %.2f\n",
  selected$model, selected$window, selected$abs_r, selected$max_abs_d,
  selected$worst_l1, selected$front_prob))
```

## Cross-corpus validity check

The Pareto plane draws its validity axis from ELLIPSE and its fairness axis from
TOEFL 11. That split is forced by the data, since ELLIPSE carries no L1 labels,
but it is only defensible if the validity ordering is a property of the
configurations rather than of the corpus that supplied the criterion. Were the
selected configuration not also near the top on TOEFL 11, the selection would be
an artefact of which corpus provided the y-axis.

TOEFL 11 proficiency is a three-level ordinal label, so validity there is a
Spearman correlation against the band rank. Two versions are reported. The
pooled estimate uses every essay. The within-L1 estimate computes the
correlation inside each L1 group and averages by group size, which matters
because band and L1 are strongly confounded in this corpus (see
`proficiency-confound-sensitivity.qmd`).

Magnitudes are **not** comparable across the two corpora. A three-category
criterion with heavy ties attenuates the coefficient relative to ELLIPSE's
continuous score, so only the ordering carries information.

```{r}
#| label: toefl-validity
BANDS     <- c("low", "medium", "high")
level_num <- as.integer(factor(level_toe, levels = BANDS))

# Spearman is Pearson on ranks. Ranking the surprisal matrix once here also lets
# the bootstrap below run on fixed rank scores instead of re-ranking every draw
# (verified identical to cor(..., method = "spearman") on the full sample).
R_toe  <- apply(M_toe, 2, rank)
rl_toe <- rank(level_num)

rho_pooled <- cor(R_toe, rl_toe)[, 1]

# Within-L1, size-weighted. Band and L1 are confounded, so the pooled estimate
# carries some L1 composition with it.
rho_within <- local({
  num <- rep(0, ncol(M_toe)); w <- 0
  for (g in sort(unique(l1_toe))) {
    ix  <- which(l1_toe == g)
    num <- num + length(ix) * cor(M_toe[ix, ], level_num[ix], method = "spearman")[, 1]
    w   <- w + length(ix)
  }
  num / w
})

validity_toefl <- tibble(config = colnames(M_toe),
                         rho = rho_pooled, abs_rho = abs(rho_pooled),
                         rho_within = rho_within, abs_rho_within = abs(rho_within))
points <- left_join(points, validity_toefl, by = "config")

xc <- points |> mutate(rank_ellipse = rank(-abs_r), rank_toefl = rank(-abs_rho))

xc |>
  arrange(rank_ellipse) |>
  select(config, instruct, abs_r, rank_ellipse, abs_rho, abs_rho_within, rank_toefl) |>
  head(12) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Top 12 configurations by ELLIPSE validity, with their TOEFL 11 ranks.")

sel_cfg <- paste(selected$model, selected$window, sep = "/")
sel_row <- xc |> filter(config == sel_cfg)
cat(sprintf("Selected %s: ELLIPSE rank %.0f/%d (|r| = %.3f) | TOEFL rank %.0f/%d (|rho| = %.3f, within-L1 %.3f)\n",
            sel_cfg, sel_row$rank_ellipse, nrow(xc), sel_row$abs_r,
            sel_row$rank_toefl, nrow(xc), sel_row$abs_rho, sel_row$abs_rho_within))
```

```{r}
#| label: cross-corpus-agreement
#| fig-width: 6.5
#| fig-height: 5
# TOEFL is resampled UNstratified here. The stratified resample used for the
# fairness axis holds band composition fixed, which would remove the very
# variation this correlation measures.
n_toe      <- nrow(M_toe)
boot_agree <- numeric(N_BOOT)
for (b in seq_len(N_BOOT)) {
  ei <- sample.int(n_ell, n_ell, replace = TRUE)
  ab <- abs(cor(S_ell[ei, ], y_ell[ei])[, 1])
  ti <- sample.int(n_toe, n_toe, replace = TRUE)
  at <- abs(cor(R_toe[ti, ], rl_toe[ti])[, 1])
  boot_agree[b] <- cor(ab, at, method = "spearman")
}

agree_hat <- cor(points$abs_r, points$abs_rho, method = "spearman")
# TEMPORARY(modernbert-base/full): known-broken configuration at the full
# window, where its surprisal exceeds its own 8-token window, which a working
# model cannot do. See "Known issues" in this directory's README. Drop this
# second reported figure once the configuration is fixed or removed.
keep      <- points$config != "modernbert-base/full"
agree_ex  <- cor(points$abs_r[keep], points$abs_rho[keep], method = "spearman")

cat(sprintf("rank agreement of the two validity orderings (%d configs): rho = %.3f, 95%% CI [%.3f, %.3f]\n",
            nrow(points), agree_hat, q025(boot_agree), q975(boot_agree)))
cat(sprintf("excluding modernbert-base/full (%d configs)              : rho = %.3f\n",
            sum(keep), agree_ex))
cat(sprintf("pooled vs within-L1 TOEFL ordering                       : rho = %.3f\n",
            cor(points$abs_rho, points$abs_rho_within, method = "spearman")))

# TEMPORARY(modernbert-base/full): excluded from this plot only, because at
# (0.034, 0.362) it compresses every other configuration into one corner. It is
# retained in every statistic above. See "Known issues" in this directory's
# README. Remove this filter once the configuration is fixed or removed.
pxc_df <- filter(points, config != "modernbert-base/full")

p_xc <- ggplot(pxc_df, aes(abs_rho, abs_r)) +
  geom_point(aes(colour = window, shape = instruct), size = 2.6) +
  ggrepel::geom_text_repel(
    data = filter(pxc_df, front | config == sel_cfg),
    aes(label = config), size = 2.5, max.overlaps = Inf, box.padding = 0.5,
    min.segment.length = 0, segment.colour = "grey55", seed = 42) +
  geom_point(data = filter(pxc_df, config == sel_cfg),
             shape = 21, size = 5, stroke = 1, colour = "black", fill = NA) +
  labs(x = "TOEFL 11 validity (|Spearman rho| vs proficiency band)",
       y = "ELLIPSE validity (|Pearson r| vs holistic score)",
       colour = "Window", shape = "Instruct",
       title = "Does the validity ordering survive changing the criterion corpus?",
       subtitle = paste("Circled point is the selected default. Axes are not on a common scale.",
                        "Broken config modernbert-base/full omitted from the plot only.",
                        sep = "\n")) +
  theme_minimal(base_size = 10)
p_xc
ggsave(file.path(RESULTS, "validity_cross_corpus.png"), p_xc, width = 6.5, height = 5, dpi = 200)
write_csv(select(xc, config, abs_r, rank_ellipse, abs_rho, abs_rho_within, rank_toefl),
          file.path(RESULTS, "validity_cross_corpus.csv"))
```

The ordering transfers. Rank agreement across the 33 configurations is 0.825
with a 95% interval of [0.796, 0.849], and the selected default sits 6th of 33
on ELLIPSE and 8th of 33 on TOEFL 11. The selection is therefore not an artefact
of which corpus supplied the y-axis, which is the assumption the two-corpus
design rests on.

Three further readings. The pooled and within-L1 TOEFL orderings agree at 0.969,
so the band-by-L1 confounding documented in
`proficiency-confound-sensitivity.qmd` does not drive the validity ordering.
The window effect replicates across corpora, with mean validity falling from
window 8 to 64 to full on both (0.626, 0.574, 0.527 on ELLIPSE; 0.499, 0.462,
0.394 on TOEFL), although it is architecture-dependent, since masked models are
level or better at window 64 while causal models are clearly best at window 8.
Finally, `modernbert-base/full` reaches |rho| = 0.034 on TOEFL 11, near zero,
which is independent corroboration that the configuration is broken rather than
merely weak.

# Section 5 — Pareto Plot

```{r}
#| label: pareto-plot
#| fig-width: 13
#| fig-height: 5
# Shared visual language (palette keyed to model family, theme, save helper).
for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}

pa <- points |>
  mutate(family = factor(model_family(model), levels = FAMILY_LEVELS),
         lbl    = short_model(model),
         window = factor(window, levels = c("8", "64", "full")))
win_labs <- c("8" = "window 8", "64" = "window 64", "full" = "window full")
selected <- selected |> mutate(window = factor(window, levels = c("8", "64", "full")))

# TEMPORARY(modernbert-base/full): this configuration sits far below the zoomed
# range, so it is annotated in the "full" panel rather than allowed to stretch
# every panel's y-axis. The 0.50 threshold selects it and nothing else. See
# "Known issues" in this directory's README. Once the configuration is fixed or
# removed, this block and the geom_text layer that consumes `off` can go.
off <- pa |> filter(abs_r < 0.50) |>
  mutate(lbl = paste0(lbl, " (|r|=", sprintf("%.2f", abs_r), ", off-scale)"),
         abs_r = 0.515)

p <- ggplot(pa, aes(max_abs_d, abs_r)) +
  geom_vline(xintercept = D_FLAG, linetype = "dotted", colour = "grey55") +
  # bootstrap CIs shown only for front configs (uncertainty that changes the pick)
  geom_errorbar(data = filter(pa, front), aes(ymin = absr_lo, ymax = absr_hi),
                width = 0, alpha = 0.35, colour = "grey55") +
  geom_errorbar(data = filter(pa, front), aes(xmin = maxd_lo, xmax = maxd_hi),
                orientation = "y", width = 0, alpha = 0.35, colour = "grey55") +
  geom_point(aes(colour = family, shape = instruct), size = 2.7) +
  geom_point(data = filter(pa, front), shape = 21, size = 4.6, stroke = 1,
             colour = "black", fill = NA) +
  geom_point(data = selected, shape = 8, size = 4.6, colour = "black") +
  ggrepel::geom_text_repel(aes(label = lbl, colour = family), size = 2.5,
                           max.overlaps = Inf, box.padding = 0.35, seed = 42,
                           show.legend = FALSE) +
  geom_text(data = off, aes(label = lbl),
            colour = "grey45", size = 2.5, vjust = 1) +
  annotate("text", x = D_FLAG, y = 0.505, label = "|d| = 0.10", hjust = -0.06,
           size = 2.9, colour = "grey45") +
  facet_wrap(~window, nrow = 1, labeller = labeller(window = win_labs)) +
  scale_colour_manual(values = FAMILY_COLORS, drop = FALSE) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17),
                     labels = c("pretrained", "instruct")) +
  coord_cartesian(xlim = c(D_FLAG, 0.66), ylim = c(0.50, 0.66)) +
  labs(
    x = "Unfairness — max |Cohen's d| across L1 groups (TOEFL 11)   → worse",
    y = "Validity — |Pearson r| with holistic score (ELLIPSE)   → better",
    colour = "Model family", shape = "Post-training",
    title = "Fairness–Validity Pareto Front, faceted by context window",
    subtitle = paste("Circled = globally non-dominated · star = selected pure-pretrained",
                     "default · CIs (front only) = 95% bootstrap · dotted = |d|=0.10 flag")) +
  theme_study1(12)

p
save_fig(p, "pareto_front", width = 13, height = 5)
```

# Section 6 — Outputs & Notes

```{r}
#| label: write-table
out <- points |>
  arrange(desc(front), desc(abs_r)) |>
  select(model, window, arch, instruct, r, abs_r, max_abs_d, worst_l1,
         n_flag, front, front_prob, absr_lo, absr_hi, maxd_lo, maxd_hi)
write_csv(out, file.path(RESULTS, "pareto_points.csv"))
cat("wrote", file.path(RESULTS, "pareto_points.csv"), "and pareto_front.png\n")
```

**Interpretation caveats (for the write-up):**

- **All correlations are negative** (more predictable text ↔ higher proficiency);
  validity is reported as the magnitude `|r|`.
- **Fairness violations are pervasive.** Every configuration flags multiple L1
  groups at |d| ≥ 0.10, so the front reflects *relative*, not absolute, fairness.
  Report the selected configuration's max |d| and flagged-group count explicitly.
- **Thin low-proficiency strata for some L1s** (e.g. GER-low n≈14, HIN-low n≈25)
  make those stratum-level *d*'s noisy; the bootstrap envelopes capture this.
- **Instruct models are a contrast condition**, excluded from default selection
  even when Pareto-optimal.
- **Truncation:** BERT (512) and GPT-2 (1024) truncate some long essays at the
  full window (flagged per-row upstream); relevant only for those specific points.

# Reproducibility

```{r}
#| label: sessioninfo
cat(sprintf("N_BOOT = %d | seed = 42\n", N_BOOT))
sessionInfo()
```
