---
title: "Study 1 — Fairness–Validity Pareto Front, by Proficiency Level"
subtitle: "Does cross-L1 displacement (and the optimal configuration) depend on writer proficiency?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

The aggregate analysis (`pareto-analysis.qmd`) collapses fairness across
proficiency by computing Cohen's *d* within each proficiency stratum and taking a
focal-group-size-weighted average. But **TOEFL 11 proficiency is heavily
imbalanced and correlated with L1** (e.g. GER/HIN/FRE skew high, ARA/JPN skew
low; the `low` band holds only 1,201 of 11,000 essays). This document reruns the
**fairness axis separately within each proficiency level** so we can see *where*
displacement concentrates and whether the selected configuration is stable across
the proficiency range.

Two estimators of per-level fairness, reported side by side:

- **all** — every essay in the level; Cohen's *d* handles unequal L1 *n*, but the
  "rest" comparison group is weighted by each L1's natural prevalence in that band.
- **balanced** — within the level, subsample every L1 down to that band's smallest
  cell (`low` = 14/L1, `medium` = 371/L1, `high` = 176/L1), averaged over
  `K_DRAWS` draws. Equal L1 representation removes the composition confound at the
  cost of power. **`low` is severely underpowered (14/L1) — interpret with caution.**

**Validity (y-axis) is unchanged** — it is measured on ELLIPSE (`|Pearson r|`
between mean surprisal and the continuous holistic score) and has no L1 or
proficiency strata. Only the fairness (x-axis) is recomputed per level.

As in `pareto-analysis.qmd`, domination runs over the **eligible set only**:
post-trained configurations are a contrast condition and may not dominate a
selectable point. The selection rule is the aggregate one applied band by band
— the highest-validity class (generative at window 8), then the smallest model
in it with a publicly released pretraining corpus.

Inputs are the tidy tables from `assemble_analysis_data.py`.

# Section 1 — Setup & Data

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

set.seed(42)
K_DRAWS <- 200        # balanced subsample repeats
N_BOOT  <- 2000       # all-essays bootstrap iterations
D_FLAG  <- 0.10       # Williamson et al. (2012) flag
# Configurations dropped as instrument failures rather than data. See "Known
# issues" in this directory's README for the diagnosis and the evidence.
EXCLUDED_CONFIGS <- c("modernbert-base/full")
LEVELS  <- c("low", "medium", "high")

find_results <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    cand <- file.path(d, "results", "predictability")
    if (dir.exists(cand)) return(cand)
    d <- dirname(d)
  }
  stop("results/predictability not found — run assemble_analysis_data.py first")
}
RESULTS <- find_results()

model_meta <- tribble(
  ~model,                  ~arch,     ~instruct, ~params_b, ~open_data,
  "bert-base",             "masked",  FALSE,     0.110,     FALSE,
  "modernbert-base",       "masked",  FALSE,     0.149,     FALSE,
  "modernbert-large",      "masked",  FALSE,     0.395,     FALSE,
  "gpt2",                  "causal",  FALSE,     0.124,     FALSE,
  "gpt2-xl",               "causal",  FALSE,     1.500,     FALSE,
  "olmo2-1b",              "causal",  FALSE,     1.000,     TRUE,
  "olmo2-7b",              "causal",  FALSE,     7.000,     TRUE,
  "qwen2.5-7b",            "causal",  FALSE,     7.000,     FALSE,
  "qwen2.5-7b-instruct",   "causal",  TRUE,      7.000,     FALSE,
  "llama3.1-8b",           "causal",  FALSE,     8.000,     FALSE,
  "llama3.1-8b-instruct",  "causal",  TRUE,      8.000,     FALSE,
)

# The aggregate report's selection rule, applied unchanged within each band.
SEL_ARCH <- "causal"; SEL_WINDOW <- "8"
```

```{r}
#| label: load-data
ell <- read_csv(file.path(RESULTS, "ellipse_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))
toe <- read_csv(file.path(RESULTS, "toefl_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

ell_wide <- ell |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(ell, text_id, overall), by = "text_id")
cc <- setdiff(names(ell_wide), c("text_id", "overall"))

toe_wide <- toe |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(toe, text_id, l1, level), by = "text_id")
stopifnot(identical(cc, setdiff(names(toe_wide), c("text_id", "l1", "level"))))
# Alignment is verified on the full matrix, then excluded configurations drop.
cc <- setdiff(cc, EXCLUDED_CONFIGS)

S <- as.matrix(ell_wide[, cc]); y <- ell_wide$overall
M <- as.matrix(toe_wide[, cc]); l1 <- toe_wide$l1; level <- toe_wide$level
```

# Section 2 — Validity (ELLIPSE, fixed)

```{r}
#| label: validity
abs_r <- abs(cor(S, y)[, 1])
validity <- tibble(config = cc, abs_r = abs_r)
```

# Section 3 — Per-level Fairness

```{r}
#| label: fairness-fns
col_var <- function(X) { n <- nrow(X); m <- colMeans(X); (colSums(X * X) - n * m * m) / (n - 1) }

# max|d| across L1 within a single-level subset (focal L1 vs the rest, pooled SD).
level_maxd <- function(M, l1) {
  groups <- sort(unique(l1)); C <- ncol(M)
  ad <- matrix(NA_real_, length(groups), C)
  for (gi in seq_along(groups)) {
    foc <- l1 == groups[gi]; rst <- !foc
    n1 <- sum(foc); n2 <- sum(rst)
    if (n1 < 2 || n2 < 2) next
    Mf <- M[foc, , drop = FALSE]; Mr <- M[rst, , drop = FALSE]
    sp <- sqrt(((n1 - 1) * col_var(Mf) + (n2 - 1) * col_var(Mr)) / (n1 + n2 - 2))
    ad[gi, ] <- abs((colMeans(Mf) - colMeans(Mr)) / sp)
  }
  list(maxd  = apply(ad, 2, max, na.rm = TRUE),
       worst = groups[apply(ad, 2, which.max)],
       nflag = colSums(ad >= D_FLAG, na.rm = TRUE))
}
```

```{r}
#| label: fairness-compute
res <- list()
for (lv in LEVELS) {
  rows <- which(level == lv); Ml <- M[rows, , drop = FALSE]; l1l <- l1[rows]
  nmin <- min(table(l1l)); rows_by_g <- split(seq_len(nrow(Ml)), l1l)

  # all essays + essay-level bootstrap CI within the level
  pe <- level_maxd(Ml, l1l)
  bd <- replicate(N_BOOT, {
    bi <- sample.int(nrow(Ml), nrow(Ml), replace = TRUE)
    level_maxd(Ml[bi, ], l1l[bi])$maxd
  })
  all_tbl <- tibble(config = cc, level = lv, method = "all",
                    maxd = pe$maxd, worst_l1 = pe$worst, n_flag = pe$nflag,
                    lo = apply(bd, 1, quantile, .025), hi = apply(bd, 1, quantile, .975),
                    n_per_l1 = NA_integer_)

  # balanced subsample to nmin/L1, averaged over draws (spread across draws = band)
  dd <- replicate(K_DRAWS, {
    idx <- unlist(lapply(rows_by_g, \(ix) sample(ix, nmin)), use.names = FALSE)
    level_maxd(Ml[idx, ], l1l[idx])$maxd
  })
  bal_tbl <- tibble(config = cc, level = lv, method = "balanced",
                    maxd = rowMeans(dd), worst_l1 = NA_character_, n_flag = NA_integer_,
                    lo = apply(dd, 1, quantile, .025), hi = apply(dd, 1, quantile, .975),
                    n_per_l1 = nmin)
  res[[lv]] <- bind_rows(all_tbl, bal_tbl)
}
fairness <- bind_rows(res)
```

# Section 4 — Fronts & Selection

```{r}
#| label: fronts
front_of <- function(d) {
  ar <- d$abs_r; md <- d$maxd; n <- length(ar); keep <- rep(TRUE, n)
  for (i in seq_len(n)) {
    dom <- (ar >= ar[i]) & (md <= md[i]) & ((ar > ar[i]) | (md < md[i]))
    if (any(dom)) keep[i] <- FALSE
  }
  d$front <- keep; d
}

points <- fairness |>
  left_join(validity, by = "config") |>
  separate(config, into = c("model", "window"), sep = "/", remove = FALSE) |>
  mutate(window = factor(window, levels = c("8", "64", "full")),
         level  = factor(level, levels = LEVELS)) |>
  left_join(model_meta, by = "model") |>
  mutate(eligible = !instruct) |>
  # Domination runs over the eligible set only, as in pareto-analysis.qmd:
  # post-trained configurations are a contrast condition and may not dominate a
  # selectable point.
  group_by(level, method) |>
  group_modify(\(d, k) { d$front <- FALSE
    d$front[d$eligible] <- front_of(d[d$eligible, ])$front
    d }) |>
  ungroup()

# The aggregate rule, applied within each band: take the highest-validity class
# (generative at window 8), then the smallest model in it with a publicly
# released pretraining corpus. The class is fixed across bands because validity
# is measured on ELLIPSE and does not vary by TOEFL band; only fairness does.
picks <- points |>
  filter(eligible, arch == SEL_ARCH, window == SEL_WINDOW, open_data) |>
  group_by(level, method) |> slice_min(params_b, n = 1, with_ties = FALSE) |> ungroup()

picks |>
  arrange(level, method) |>
  transmute(level, method, pick = paste0(model, "/", window),
            abs_r = round(abs_r, 3), maxd = round(maxd, 3), worst_l1, n_flag, front) |>
  knitr::kable(caption = paste("Selected default per proficiency level, under the aggregate",
                               "selection rule. `front` is membership in that band's",
                               "eligible-set front."))

# What the front-based rule would have picked instead, band by band, since the
# bands are where the two rules can diverge.
points |> filter(front) |>
  group_by(level, method) |> slice_max(abs_r, n = 1, with_ties = FALSE) |> ungroup() |>
  arrange(level, method) |>
  transmute(level, method, front_leader = paste0(model, "/", window),
            abs_r = round(abs_r, 3), maxd = round(maxd, 3), worst_l1) |>
  knitr::kable(caption = "For contrast: the highest-validity point on each band's eligible front.")

write_csv(points, file.path(RESULTS, "pareto_by_level.csv"))
```

# Section 5 — Faceted Pareto Plot

```{r}
#| label: plot
#| fig-width: 11
#| fig-height: 9.5
# Shared visual language (level palette, theme, save helper).
for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}

# Rows = configs, ordered by medium-band unfairness (best at top). One common x
# axis across all levels and both estimators, so the low->high gradient reads as
# horizontal displacement (the free-x facets in the old figure hid this).
ord <- points |> filter(method == "all", level == "medium") |>
  arrange(desc(maxd)) |> pull(config)
pc <- points |> mutate(config = factor(config, levels = ord),
                       level  = factor(level, levels = LEVEL_LEVELS))
seg <- pc |> filter(level %in% c("low", "high")) |>
  select(config, method, level, maxd) |>
  pivot_wider(names_from = level, values_from = maxd)
meth_labs <- c(all = "all essays", balanced = "balanced (equal L1 per level)")

p <- ggplot(pc, aes(maxd, config, colour = level)) +
  geom_vline(xintercept = D_FLAG, linetype = "dotted", colour = "grey55") +
  geom_segment(data = seg, aes(x = high, xend = low, y = config, yend = config),
               inherit.aes = FALSE, colour = "grey78", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0, alpha = 0.30) +
  geom_point(size = 1.9) +
  facet_wrap(~method, labeller = labeller(method = meth_labs)) +
  scale_colour_manual(values = LEVEL_COLORS, name = "Proficiency",
                      breaks = LEVEL_LEVELS) +
  labs(
    x = "Unfairness — max |Cohen's d| across L1 groups (within level)   → worse",
    y = NULL,
    title = "Cross-L1 Unfairness by Proficiency Level (shared scale)",
    subtitle = paste("Each row = one configuration; dots = low / medium / high;",
                     "grey line spans low↔high. Dotted = |d|=0.10 flag.\nlow is",
                     "severely underpowered (14/L1 balanced) — hence the wide CIs",
                     "and the long rightward reach.")) +
  theme_study1(10) +
  theme(axis.text.y = element_text(size = 6.5),
        panel.grid.major.y = element_line(colour = "grey93"))

p
save_fig(p, "pareto_by_level", width = 8.5, height = 10)
```

# Section 6 — Notes

```{r}
#| label: gradient
# Unfairness gradient across proficiency (all-essays estimator) for context.
points |> filter(method == "all") |>
  group_by(level) |>
  summarise(median_maxd = round(median(maxd), 3),
            min_maxd = round(min(maxd), 3), max_maxd = round(max(maxd), 3),
            .groups = "drop") |>
  knitr::kable(caption = sprintf("Distribution of max|d| across the %d configs, by level (all-essays).",
                                 n_distinct(points$config)))
```

**Takeaways for the write-up:**

- **Displacement concentrates at low proficiency** and is smallest at high — the
  worst-case L1 gap shrinks as writer proficiency rises.
- **Balancing changes little at medium/high** (the aggregate was not a composition
  artifact) but inflates and destabilizes `low`, confirming that band is too thin
  to support a reliable per-L1 effect size.
- **Configuration selection is not proficiency-dependent, once the front is
  computed over the eligible set.** The earlier version of this report picked a
  different configuration at `low` than at `medium`/`high`, but that instability
  was an artefact: the band-wise front was computed over the full matrix and then
  filtered to pure-pretrained configurations, so at `low` the eligible leader was
  displaced by an instruct model it could not be compared against. With the
  contrast configurations excluded from the domination comparison, the band
  fronts have the same eligible leader in every band, and the class-based
  selection rule — which reads only ELLIPSE validity, and so cannot vary by TOEFL
  band — returns `olmo2-1b/8` throughout. What varies across the proficiency
  range is the *magnitude* of displacement, not the choice.
- **The selected default is on no band's front**, as in the aggregate analysis,
  and its displacement is roughly twice as large at `low` and `medium` (max |d| ≈
  .58, worst group GER) as at `high` (.32, SPA). The band-wise numbers are the
  ones to report as the fairness cost of the selection, not the aggregate .422.
- Every configuration still flags multiple L1 groups at |d| ≥ 0.10 in every band;
  the front remains a *least-bad* set, not an absolutely fair one.

# Reproducibility

```{r}
#| label: sessioninfo
cat(sprintf("K_DRAWS = %d | N_BOOT = %d | seed = 42\n", K_DRAWS, N_BOOT))
sessionInfo()
```
