---
title: "Study 1 — Fairness–Validity Pareto Front, by Proficiency Level"
subtitle: "Does cross-L1 displacement (and the optimal configuration) depend on writer proficiency?"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

The aggregate analysis (`pareto-analysis.qmd`) collapses fairness across
proficiency by computing Cohen's *d* within each proficiency stratum and taking a
focal-group-size-weighted average. But **TOEFL 11 proficiency is heavily
imbalanced and correlated with L1** (e.g. GER/HIN/FRE skew high, ARA/JPN skew
low; the `low` band holds only 1,201 of 11,000 essays). This document reruns the
**fairness axis separately within each proficiency level** so we can see *where*
displacement concentrates and whether the selected configuration is stable across
the proficiency range.

Two estimators of per-level fairness, reported side by side:

- **all** — every essay in the level; Cohen's *d* handles unequal L1 *n*, but the
  "rest" comparison group is weighted by each L1's natural prevalence in that band.
- **balanced** — within the level, subsample every L1 down to that band's smallest
  cell (`low` = 14/L1, `medium` = 371/L1, `high` = 176/L1), averaged over
  `K_DRAWS` draws. Equal L1 representation removes the composition confound at the
  cost of power. **`low` is severely underpowered (14/L1) — interpret with caution.**

**Validity (y-axis) is unchanged** — it is measured on ELLIPSE (`|Pearson r|`
between mean surprisal and the continuous holistic score) and has no L1 or
proficiency strata. Only the fairness (x-axis) is recomputed per level.

Inputs are the tidy tables from `assemble_analysis_data.py`.

# Section 1 — Setup & Data

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)

set.seed(42)
K_DRAWS <- 200        # balanced subsample repeats
N_BOOT  <- 2000       # all-essays bootstrap iterations
D_FLAG  <- 0.10       # Williamson et al. (2012) flag
# Configurations dropped as instrument failures rather than data. See "Known
# issues" in this directory's README for the diagnosis and the evidence.
EXCLUDED_CONFIGS <- c("modernbert-base/full")
LEVELS  <- c("low", "medium", "high")

find_results <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    cand <- file.path(d, "results", "predictability")
    if (dir.exists(cand)) return(cand)
    d <- dirname(d)
  }
  stop("results/predictability not found — run assemble_analysis_data.py first")
}
RESULTS <- find_results()

model_meta <- tribble(
  ~model,                  ~arch,     ~instruct,
  "bert-base",             "masked",  FALSE,
  "modernbert-base",       "masked",  FALSE,
  "modernbert-large",      "masked",  FALSE,
  "gpt2",                  "causal",  FALSE,
  "gpt2-xl",               "causal",  FALSE,
  "olmo2-1b",              "causal",  FALSE,
  "olmo2-7b",              "causal",  FALSE,
  "qwen2.5-7b",            "causal",  FALSE,
  "qwen2.5-7b-instruct",   "causal",  TRUE,
  "llama3.1-8b",           "causal",  FALSE,
  "llama3.1-8b-instruct",  "causal",  TRUE,
)
```

```{r}
#| label: load-data
ell <- read_csv(file.path(RESULTS, "ellipse_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))
toe <- read_csv(file.path(RESULTS, "toefl_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

ell_wide <- ell |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(ell, text_id, overall), by = "text_id")
cc <- setdiff(names(ell_wide), c("text_id", "overall"))

toe_wide <- toe |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(toe, text_id, l1, level), by = "text_id")
stopifnot(identical(cc, setdiff(names(toe_wide), c("text_id", "l1", "level"))))
# Alignment is verified on the full matrix, then excluded configurations drop.
cc <- setdiff(cc, EXCLUDED_CONFIGS)

S <- as.matrix(ell_wide[, cc]); y <- ell_wide$overall
M <- as.matrix(toe_wide[, cc]); l1 <- toe_wide$l1; level <- toe_wide$level
```

# Section 2 — Validity (ELLIPSE, fixed)

```{r}
#| label: validity
abs_r <- abs(cor(S, y)[, 1])
validity <- tibble(config = cc, abs_r = abs_r)
```

# Section 3 — Per-level Fairness

```{r}
#| label: fairness-fns
col_var <- function(X) { n <- nrow(X); m <- colMeans(X); (colSums(X * X) - n * m * m) / (n - 1) }

# max|d| across L1 within a single-level subset (focal L1 vs the rest, pooled SD).
level_maxd <- function(M, l1) {
  groups <- sort(unique(l1)); C <- ncol(M)
  ad <- matrix(NA_real_, length(groups), C)
  for (gi in seq_along(groups)) {
    foc <- l1 == groups[gi]; rst <- !foc
    n1 <- sum(foc); n2 <- sum(rst)
    if (n1 < 2 || n2 < 2) next
    Mf <- M[foc, , drop = FALSE]; Mr <- M[rst, , drop = FALSE]
    sp <- sqrt(((n1 - 1) * col_var(Mf) + (n2 - 1) * col_var(Mr)) / (n1 + n2 - 2))
    ad[gi, ] <- abs((colMeans(Mf) - colMeans(Mr)) / sp)
  }
  list(maxd  = apply(ad, 2, max, na.rm = TRUE),
       worst = groups[apply(ad, 2, which.max)],
       nflag = colSums(ad >= D_FLAG, na.rm = TRUE))
}
```

```{r}
#| label: fairness-compute
res <- list()
for (lv in LEVELS) {
  rows <- which(level == lv); Ml <- M[rows, , drop = FALSE]; l1l <- l1[rows]
  nmin <- min(table(l1l)); rows_by_g <- split(seq_len(nrow(Ml)), l1l)

  # all essays + essay-level bootstrap CI within the level
  pe <- level_maxd(Ml, l1l)
  bd <- replicate(N_BOOT, {
    bi <- sample.int(nrow(Ml), nrow(Ml), replace = TRUE)
    level_maxd(Ml[bi, ], l1l[bi])$maxd
  })
  all_tbl <- tibble(config = cc, level = lv, method = "all",
                    maxd = pe$maxd, worst_l1 = pe$worst, n_flag = pe$nflag,
                    lo = apply(bd, 1, quantile, .025), hi = apply(bd, 1, quantile, .975),
                    n_per_l1 = NA_integer_)

  # balanced subsample to nmin/L1, averaged over draws (spread across draws = band)
  dd <- replicate(K_DRAWS, {
    idx <- unlist(lapply(rows_by_g, \(ix) sample(ix, nmin)), use.names = FALSE)
    level_maxd(Ml[idx, ], l1l[idx])$maxd
  })
  bal_tbl <- tibble(config = cc, level = lv, method = "balanced",
                    maxd = rowMeans(dd), worst_l1 = NA_character_, n_flag = NA_integer_,
                    lo = apply(dd, 1, quantile, .025), hi = apply(dd, 1, quantile, .975),
                    n_per_l1 = nmin)
  res[[lv]] <- bind_rows(all_tbl, bal_tbl)
}
fairness <- bind_rows(res)
```

# Section 4 — Fronts & Selection

```{r}
#| label: fronts
front_of <- function(d) {
  ar <- d$abs_r; md <- d$maxd; n <- length(ar); keep <- rep(TRUE, n)
  for (i in seq_len(n)) {
    dom <- (ar >= ar[i]) & (md <= md[i]) & ((ar > ar[i]) | (md < md[i]))
    if (any(dom)) keep[i] <- FALSE
  }
  d$front <- keep; d
}

points <- fairness |>
  left_join(validity, by = "config") |>
  separate(config, into = c("model", "window"), sep = "/", remove = FALSE) |>
  mutate(window = factor(window, levels = c("8", "64", "full")),
         level  = factor(level, levels = LEVELS)) |>
  left_join(model_meta, by = "model") |>
  group_by(level, method) |> group_modify(~front_of(.x)) |> ungroup()

# Selected pure-pretrained default per (level, method): highest |r| on the front.
picks <- points |> filter(front, !instruct) |>
  group_by(level, method) |> slice_max(abs_r, n = 1, with_ties = FALSE) |> ungroup()

picks |>
  arrange(level, method) |>
  transmute(level, method, pick = paste0(model, "/", window),
            abs_r = round(abs_r, 3), maxd = round(maxd, 3), worst_l1) |>
  knitr::kable(caption = "Selected pure-pretrained default per proficiency level (highest |r| on that level's front).")

write_csv(points, file.path(RESULTS, "pareto_by_level.csv"))
```

# Section 5 — Faceted Pareto Plot

```{r}
#| label: plot
#| fig-width: 11
#| fig-height: 9.5
# Shared visual language (level palette, theme, save helper).
for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}

# Rows = configs, ordered by medium-band unfairness (best at top). One common x
# axis across all levels and both estimators, so the low->high gradient reads as
# horizontal displacement (the free-x facets in the old figure hid this).
ord <- points |> filter(method == "all", level == "medium") |>
  arrange(desc(maxd)) |> pull(config)
pc <- points |> mutate(config = factor(config, levels = ord),
                       level  = factor(level, levels = LEVEL_LEVELS))
seg <- pc |> filter(level %in% c("low", "high")) |>
  select(config, method, level, maxd) |>
  pivot_wider(names_from = level, values_from = maxd)
meth_labs <- c(all = "all essays", balanced = "balanced (equal L1 per level)")

p <- ggplot(pc, aes(maxd, config, colour = level)) +
  geom_vline(xintercept = D_FLAG, linetype = "dotted", colour = "grey55") +
  geom_segment(data = seg, aes(x = high, xend = low, y = config, yend = config),
               inherit.aes = FALSE, colour = "grey78", linewidth = 0.4) +
  geom_errorbar(aes(xmin = lo, xmax = hi), orientation = "y",
                width = 0, alpha = 0.30) +
  geom_point(size = 1.9) +
  facet_wrap(~method, labeller = labeller(method = meth_labs)) +
  scale_colour_manual(values = LEVEL_COLORS, name = "Proficiency",
                      breaks = LEVEL_LEVELS) +
  labs(
    x = "Unfairness — max |Cohen's d| across L1 groups (within level)   → worse",
    y = NULL,
    title = "Cross-L1 Unfairness by Proficiency Level (shared scale)",
    subtitle = paste("Each row = one configuration; dots = low / medium / high;",
                     "grey line spans low↔high. Dotted = |d|=0.10 flag.\nlow is",
                     "severely underpowered (14/L1 balanced) — hence the wide CIs",
                     "and the long rightward reach.")) +
  theme_study1(10) +
  theme(axis.text.y = element_text(size = 6.5),
        panel.grid.major.y = element_line(colour = "grey93"))

p
save_fig(p, "pareto_by_level", width = 11, height = 9.5)
```

# Section 6 — Notes

```{r}
#| label: gradient
# Unfairness gradient across proficiency (all-essays estimator) for context.
points |> filter(method == "all") |>
  group_by(level) |>
  summarise(median_maxd = round(median(maxd), 3),
            min_maxd = round(min(maxd), 3), max_maxd = round(max(maxd), 3),
            .groups = "drop") |>
  knitr::kable(caption = "Distribution of max|d| across the 33 configs, by level (all-essays).")
```

**Takeaways for the write-up:**

- **Displacement concentrates at low proficiency** and is smallest at high — the
  worst-case L1 gap shrinks as writer proficiency rises.
- **Balancing changes little at medium/high** (the aggregate was not a composition
  artifact) but inflates and destabilizes `low`, confirming that band is too thin
  to support a reliable per-L1 effect size.
- **Configuration selection is proficiency-dependent.** Report the per-level picks
  alongside the aggregate default rather than assuming one configuration is optimal
  across the whole proficiency range.
- Every configuration still flags multiple L1 groups at |d| ≥ 0.10 in every band;
  the front remains a *least-bad* set, not an absolutely fair one.

# Reproducibility

```{r}
#| label: sessioninfo
cat(sprintf("K_DRAWS = %d | N_BOOT = %d | seed = 42\n", K_DRAWS, N_BOOT))
sessionInfo()
```
