---
title: "Study 1 — Fairness–Validity Pareto Front"
subtitle: "Selecting an LLM-Surprisal Configuration that Balances Concurrent Validity and Cross-L1 Fairness"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

For every configuration in the Study 1 matrix — **11 language models × 3 context
windows (8 / 64 / full)** — we place each on a two-axis plane and find the
non-dominated (Pareto) set:

- **Validity (y):** strength of the concurrent correlation between mean token
  surprisal and human holistic proficiency, measured on **ELLIPSE** as
  `|Pearson r|`. (All correlations are negative — more proficient writers produce
  more predictable text — so validity is the *magnitude* `|r|`; higher is better.)
- **Fairness (x):** worst-case cross-L1 score displacement on **TOEFL 11**,
  measured as **max |Cohen's d|** across the 11 L1 groups, computed within
  proficiency strata (DIF logic). Lower is better.

A configuration is **Pareto-optimal** if no other configuration is simultaneously
*more valid* (higher `|r|`) **and** *fairer* (lower max `|d|`). Post-trained
(instruct) configurations are a contrast condition rather than eligible defaults
(plan §1), so they are excluded from the **domination comparison** as well as
from selection. The front is computed over the 26 eligible configurations; the 6
contrast configurations are plotted but are never allowed to dominate an
eligible point, since a configuration that loses only to a model that cannot be
selected is not dominated for the purpose at hand.

**Selection does not follow from the front alone.** No configuration is fair in
absolute terms — every one flags multiple L1 groups at `|d| ≥ 0.10` — so the
front orders a trade-off rather than identifying an acceptable point on it. The
default for Studies 2–3 is therefore taken from the **highest-validity
configuration class** (generative models at the 8-token window), as the
**smallest model in that class with a publicly released pretraining corpus**.
Reproducibility of the pretraining data and the cost of running the model are
weighted above differences in `|r|` of a few thousandths, which is the scale
that separates configurations within the class.

Inputs are the two long tables written by `assemble_analysis_data.py`
(`results/predictability/{ellipse,toefl}_long.csv.gz`).

# Section 1 — Setup & Data

```{r}
#| label: setup
# Individual tidyverse components (the `tidyverse` meta-package pulls xml2/rvest,
# which need system libxml2 and aren't used here).
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(cocor)     # Zou (2007) intervals for correlation differences

set.seed(42)          # fixed bootstrap seed (reproducibility)
N_BOOT   <- 2000      # bootstrap iterations (plan §3.3)
D_FLAG   <- 0.10      # Williamson et al. (2012) fairness flag threshold

# Configurations dropped as instrument failures rather than data. See "Known
# issues" in this directory's README for the diagnosis and the evidence.
EXCLUDED_CONFIGS <- c("modernbert-base/full")

# Locate results/predictability by walking up from the working dir (Quarto runs
# from the .qmd's directory), so this works whether rendered or run standalone.
find_results <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    cand <- file.path(d, "results", "predictability")
    if (dir.exists(cand)) return(cand)
    d <- dirname(d)
  }
  stop("results/predictability not found — run assemble_analysis_data.py first")
}
RESULTS <- find_results()
```

```{r}
#| label: model-meta
# Architecture / post-training status per model (from models.py registry), plus
# the two attributes the selection rule reads: parameter count and whether the
# pretraining corpus is publicly released as a downloadable dataset.
#   open_data TRUE  — OLMo-2 (Dolma / OLMo-mix-1124, released by AI2).
#   open_data FALSE — GPT-2 (WebText never released), Qwen2.5 and Llama-3.1
#                     (undisclosed), ModernBERT (mix undisclosed), and BERT,
#                     whose BooksCorpus half was withdrawn from distribution.
model_meta <- tribble(
  ~model,                  ~arch,     ~instruct, ~params_b, ~open_data,
  "bert-base",             "masked",  FALSE,     0.110,     FALSE,
  "modernbert-base",       "masked",  FALSE,     0.149,     FALSE,
  "modernbert-large",      "masked",  FALSE,     0.395,     FALSE,
  "gpt2",                  "causal",  FALSE,     0.124,     FALSE,
  "gpt2-xl",               "causal",  FALSE,     1.500,     FALSE,
  "olmo2-1b",              "causal",  FALSE,     1.000,     TRUE,
  "olmo2-7b",              "causal",  FALSE,     7.000,     TRUE,
  "qwen2.5-7b",            "causal",  FALSE,     7.000,     FALSE,
  "qwen2.5-7b-instruct",   "causal",  TRUE,      7.000,     FALSE,
  "llama3.1-8b",           "causal",  FALSE,     8.000,     FALSE,
  "llama3.1-8b-instruct",  "causal",  TRUE,      8.000,     FALSE,
)
WINDOWS <- c("8", "64", "full")
MODELS  <- model_meta$model
# Base/instruct pairs for the post-training contrast (plan §3.4).
INSTRUCT_PAIRS <- list(c("qwen2.5-7b", "qwen2.5-7b-instruct"),
                       c("llama3.1-8b", "llama3.1-8b-instruct"))
```

```{r}
#| label: load-data
ell <- read_csv(file.path(RESULTS, "ellipse_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

toe <- read_csv(file.path(RESULTS, "toefl_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

# ELLIPSE as an [essay x config] surprisal matrix + aligned criterion vector,
# so every configuration shares the same essays (dependent, essay-level resampling).
ell_wide <- ell |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(ell, text_id, overall), by = "text_id")

config_cols <- setdiff(names(ell_wide), c("text_id", "overall"))

# TOEFL as an [essay x config] surprisal matrix + L1 / proficiency-level vectors.
toe_wide <- toe |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(toe, text_id, l1, level), by = "text_id")

stopifnot(identical(config_cols, setdiff(names(toe_wide), c("text_id", "l1", "level"))))

# Alignment is verified on the full matrix, then excluded configurations drop.
dropped     <- intersect(config_cols, EXCLUDED_CONFIGS)
config_cols <- setdiff(config_cols, EXCLUDED_CONFIGS)

S_ell <- as.matrix(ell_wide[, config_cols])   # essays x configs
y_ell <- ell_wide$overall
M_toe <- as.matrix(toe_wide[, config_cols])   # essays x configs
l1_toe    <- toe_wide$l1
level_toe <- toe_wide$level

cat(sprintf("configs: %d | ELLIPSE essays: %d | TOEFL essays: %d\n",
            length(config_cols), nrow(S_ell), nrow(M_toe)))
if (length(dropped)) cat("excluded as instrument failure:", paste(dropped, collapse = ", "), "\n")
```

# Section 2 — Validity (ELLIPSE)

Per configuration, the Pearson correlation between mean surprisal and the
holistic `overall` score, reported as the magnitude `|r|`.

```{r}
#| label: validity
# cor(S, y) returns one correlation per column of S -> vector over configs.
r_hat   <- cor(S_ell, y_ell)[, 1]
validity <- tibble(config = names(r_hat), r = r_hat, abs_r = abs(r_hat))
```

# Section 3 — Fairness (TOEFL 11)

For each configuration and each focal L1 group *g*: within each proficiency
stratum *s* (low / medium / high), Cohen's *d* in surprisal between *g* and all
remaining learners, standardized by the **pooled within-stratum SD**; the three
stratum *d*'s are combined into a focal-group **weighted average** (weighted by
*g*'s stratum sizes). The configuration's fairness score is **max |d|** across
the 11 groups (worst-case). We also count how many groups exceed the |d| ≥ 0.10
flag.

```{r}
#| label: fairness-fns
# Fast per-column variance (avoids apply): var of each column of X.
col_var <- function(X) {
  n <- nrow(X); m <- colMeans(X)
  (colSums(X * X) - n * m * m) / (n - 1)
}

# Weighted-average Cohen's d per (group) for every config column of M.
# Returns a [group x config] matrix of weighted-average d.
group_d_matrix <- function(M, l1, level,
                            groups = sort(unique(l1)),
                            strata = c("low", "medium", "high")) {
  d_wavg <- matrix(NA_real_, nrow = length(groups), ncol = ncol(M),
                   dimnames = list(groups, colnames(M)))
  for (g in groups) {
    num  <- rep(0, ncol(M)); wsum <- 0
    for (s in strata) {
      in_s  <- level == s
      foc   <- in_s & (l1 == g)
      rst   <- in_s & (l1 != g)
      n1 <- sum(foc); n2 <- sum(rst)
      if (n1 < 2L || n2 < 2L) next
      Mf <- M[foc, , drop = FALSE]; Mr <- M[rst, , drop = FALSE]
      m1 <- colMeans(Mf); m2 <- colMeans(Mr)
      v1 <- col_var(Mf);  v2 <- col_var(Mr)
      sp <- sqrt(((n1 - 1) * v1 + (n2 - 1) * v2) / (n1 + n2 - 2))
      d  <- (m1 - m2) / sp
      num  <- num + n1 * d          # weight stratum by focal-group size
      wsum <- wsum + n1
    }
    d_wavg[g, ] <- num / wsum
  }
  d_wavg
}

# Fairness summaries per config: max|d|, the worst L1, and #groups over the flag.
fairness_summary <- function(M, l1, level) {
  D <- group_d_matrix(M, l1, level)
  absD <- abs(D)
  tibble(
    config    = colnames(M),
    max_abs_d = apply(absD, 2, max, na.rm = TRUE),
    worst_l1  = rownames(D)[apply(absD, 2, which.max)],
    n_flag    = colSums(absD >= D_FLAG, na.rm = TRUE)
  )
}
```

```{r}
#| label: fairness
fairness <- fairness_summary(M_toe, l1_toe, level_toe)
```

# Section 4 — Pareto Front

```{r}
#| label: assemble-front
is_front <- function(abs_r, max_abs_d) {
  # config i is dominated if some j is >= on validity AND <= on fairness,
  # with at least one strict.
  n <- length(abs_r); keep <- rep(TRUE, n)
  for (i in seq_len(n)) {
    dom <- (abs_r >= abs_r[i]) & (max_abs_d <= max_abs_d[i]) &
           ((abs_r > abs_r[i]) | (max_abs_d < max_abs_d[i]))
    if (any(dom)) keep[i] <- FALSE
  }
  keep
}

points <- validity |>
  left_join(fairness, by = "config") |>
  separate(config, into = c("model", "window"), sep = "/", remove = FALSE) |>
  mutate(window = factor(window, levels = WINDOWS)) |>
  left_join(model_meta, by = "model") |>
  mutate(eligible = !instruct, front = FALSE)

# Domination runs over the eligible set only (see Overview). Contrast
# configurations keep front = FALSE and take no part in the comparison.
points$front[points$eligible] <- is_front(points$abs_r[points$eligible],
                                          points$max_abs_d[points$eligible])

# Column mask over S_ell / M_toe, so the bootstrap can apply the same restriction.
stopifnot(identical(points$config, config_cols))
ELIGIBLE <- points$eligible

cat(sprintf("eligible: %d | contrast (excluded from domination): %d | on front: %d\n",
            sum(points$eligible), sum(!points$eligible), sum(points$front)))

points |>
  arrange(desc(front), desc(abs_r)) |>
  select(model, window, arch, instruct, r, abs_r, max_abs_d, worst_l1, n_flag, front) |>
  mutate(across(c(r, abs_r, max_abs_d), \(x) round(x, 3)),
         front = ifelse(instruct, NA, front)) |>
  knitr::kable(caption = sprintf(
    paste("All %d configurations (front first, then by validity). Front is computed over the",
          "%d eligible configurations; the %d contrast configurations show NA."),
    length(config_cols), sum(points$eligible), sum(!points$eligible)))
```

## Bootstrap confidence regions

2,000 iterations. ELLIPSE essays are resampled with replacement for `|r|`; TOEFL
essays are resampled **stratified by proficiency level** for max `|d|` (the two
corpora are independent samples, so they are resampled independently). We report
per-axis 95% percentile intervals and, for each eligible configuration, its
**bootstrap front-membership probability** (share of iterations in which it is
non-dominated). Front membership is evaluated **within the eligible set on every
iteration**, matching the point-estimate front; contrast configurations are
resampled and get intervals on both axes, but no front probability.

```{r}
#| label: bootstrap
n_ell   <- nrow(S_ell)
lvl_idx <- split(seq_len(nrow(M_toe)), level_toe)   # row indices per stratum

boot_absr <- matrix(NA_real_, N_BOOT, ncol(S_ell), dimnames = list(NULL, colnames(S_ell)))
boot_maxd <- matrix(NA_real_, N_BOOT, ncol(M_toe), dimnames = list(NULL, colnames(M_toe)))
boot_front <- matrix(NA,      N_BOOT, ncol(S_ell), dimnames = list(NULL, colnames(S_ell)))

for (b in seq_len(N_BOOT)) {
  # validity: essay-level resample shared across configs
  ei <- sample.int(n_ell, n_ell, replace = TRUE)
  ab <- abs(cor(S_ell[ei, ], y_ell[ei])[, 1])

  # fairness: stratified resample
  ti <- unlist(lapply(lvl_idx, \(ix) sample(ix, length(ix), replace = TRUE)),
               use.names = FALSE)
  D  <- group_d_matrix(M_toe[ti, ], l1_toe[ti], level_toe[ti])
  md <- apply(abs(D), 2, max, na.rm = TRUE)

  boot_absr[b, ] <- ab
  boot_maxd[b, ] <- md
  boot_front[b, ELIGIBLE] <- is_front(ab[ELIGIBLE], md[ELIGIBLE])
}

q025 <- \(x) quantile(x, 0.025, names = FALSE)
q975 <- \(x) quantile(x, 0.975, names = FALSE)

boot_ci <- tibble(
  config      = colnames(S_ell),
  absr_lo     = apply(boot_absr, 2, q025),
  absr_hi     = apply(boot_absr, 2, q975),
  maxd_lo     = apply(boot_maxd, 2, q025),
  maxd_hi     = apply(boot_maxd, 2, q975),
  front_prob  = colMeans(boot_front)
)

points <- left_join(points, boot_ci, by = "config")
```

## The front

Every configuration on the front is eligible by construction, so the table
carries no instruct column.

```{r}
#| label: front-table
front_tbl <- points |> filter(front) |> arrange(desc(abs_r))

front_tbl |>
  select(model, window, arch, abs_r, max_abs_d, worst_l1, n_flag, front_prob) |>
  mutate(across(c(abs_r, max_abs_d, front_prob), \(x) round(x, 3))) |>
  knitr::kable(caption = sprintf(
    "Non-dominated eligible configurations (%d of %d). Contrast configurations take no part in the domination comparison.",
    nrow(front_tbl), sum(points$eligible)))
```

## Selected configuration

The front is a trade-off curve, not a shortlist of acceptable configurations.
Every point on it flags at least five L1 groups at `|d| ≥ 0.10`, so choosing the
front's validity endpoint would buy a worst-case displacement of 0.442 for a
validity gain measured in thousandths. Selection therefore runs on the
**configuration class** instead.

The class is identified empirically, as the architecture-by-window cell with the
highest mean eligible validity, and the choice within it is made on grounds the
Pareto plane does not encode: whether the pretraining corpus is publicly
released, and how expensive the model is to run.

```{r}
#| label: selection
class_rank <- points |> filter(eligible) |>
  group_by(arch, window) |>
  summarise(n = n(), mean_abs_r = mean(abs_r), max_abs_r = max(abs_r), .groups = "drop") |>
  arrange(desc(mean_abs_r))

class_rank |>
  mutate(across(c(mean_abs_r, max_abs_r), \(x) round(x, 3))) |>
  knitr::kable(caption = "Eligible configuration classes, by mean ELLIPSE validity.")

top_class <- slice(class_rank, 1)

# Within the top class: publicly released pretraining corpus, then smallest model.
selected <- points |>
  filter(eligible, arch == top_class$arch, window == top_class$window, open_data) |>
  arrange(params_b) |>
  slice(1)

points |>
  filter(eligible, arch == top_class$arch, window == top_class$window) |>
  arrange(desc(abs_r)) |>
  select(model, params_b, open_data, abs_r, max_abs_d, worst_l1, n_flag, front, front_prob) |>
  mutate(across(c(abs_r, max_abs_d, front_prob), \(x) round(x, 3))) |>
  knitr::kable(caption = sprintf("The %s / window %s class, from which the default is drawn.",
                                 top_class$arch, top_class$window))

cat(sprintf(
  "\nHighest-validity class: %s models at window %s (%d configurations, mean |r| = %.3f).\n",
  top_class$arch, top_class$window, top_class$n, top_class$mean_abs_r))
cat(sprintf(
  "Selected default (smallest open-pretraining-data model in that class):\n  %s / window %s  |  %.0fB params  |  |r| = %.3f  |  max|d| = %.3f  |  worst L1 = %s  |  flagged = %d  |  front prob = %.2f\n",
  selected$model, selected$window, selected$params_b, selected$abs_r,
  selected$max_abs_d, selected$worst_l1, selected$n_flag, selected$front_prob))

# The selection is not required to land on the front; say so explicitly when it
# does not, and name what dominates it, rather than leaving it to the figure.
if (!selected$front) {
  dom <- points |>
    filter(eligible, config != selected$config,
           abs_r >= selected$abs_r, max_abs_d <= selected$max_abs_d) |>
    arrange(desc(abs_r))
  cat(sprintf(
    "Not on the front. Dominated by %s (|r| = %.3f vs %.3f, max|d| = %.3f vs %.3f) — the same pretraining corpus at %.0fx the parameter count.\n",
    paste(dom$config, collapse = ", "), dom$abs_r[1], selected$abs_r,
    dom$max_abs_d[1], selected$max_abs_d, dom$params_b[1] / selected$params_b))
}
```

## Cross-corpus validity check

The Pareto plane draws its validity axis from ELLIPSE and its fairness axis from
TOEFL 11. That split is forced by the data, since ELLIPSE carries no L1 labels,
but it is only defensible if the validity ordering is a property of the
configurations rather than of the corpus that supplied the criterion. Were the
selected configuration not also near the top on TOEFL 11, the selection would be
an artefact of which corpus provided the y-axis.

TOEFL 11 proficiency is a three-level ordinal label, so validity there is a
Spearman correlation against the band rank. Two versions are reported. The
pooled estimate uses every essay. The within-L1 estimate computes the
correlation inside each L1 group and averages by group size, which matters
because band and L1 are strongly confounded in this corpus (see
`proficiency-confound-sensitivity.qmd`).

Magnitudes are **not** comparable across the two corpora. A three-category
criterion with heavy ties attenuates the coefficient relative to ELLIPSE's
continuous score, so only the ordering carries information.

```{r}
#| label: toefl-validity
BANDS     <- c("low", "medium", "high")
level_num <- as.integer(factor(level_toe, levels = BANDS))

# Spearman is Pearson on ranks. Ranking the surprisal matrix once here also lets
# the bootstrap below run on fixed rank scores instead of re-ranking every draw
# (verified identical to cor(..., method = "spearman") on the full sample).
R_toe  <- apply(M_toe, 2, rank)
rl_toe <- rank(level_num)

rho_pooled <- cor(R_toe, rl_toe)[, 1]

# Within-L1, size-weighted. Band and L1 are confounded, so the pooled estimate
# carries some L1 composition with it.
rho_within <- local({
  num <- rep(0, ncol(M_toe)); w <- 0
  for (g in sort(unique(l1_toe))) {
    ix  <- which(l1_toe == g)
    num <- num + length(ix) * cor(M_toe[ix, ], level_num[ix], method = "spearman")[, 1]
    w   <- w + length(ix)
  }
  num / w
})

validity_toefl <- tibble(config = colnames(M_toe),
                         rho = rho_pooled, abs_rho = abs(rho_pooled),
                         rho_within = rho_within, abs_rho_within = abs(rho_within))
points <- left_join(points, validity_toefl, by = "config")

xc <- points |> mutate(rank_ellipse = rank(-abs_r), rank_toefl = rank(-abs_rho))

xc |>
  arrange(rank_ellipse) |>
  select(config, instruct, abs_r, rank_ellipse, abs_rho, abs_rho_within, rank_toefl) |>
  head(12) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Top 12 configurations by ELLIPSE validity, with their TOEFL 11 ranks.")

sel_cfg <- paste(selected$model, selected$window, sep = "/")
sel_row <- xc |> filter(config == sel_cfg)
cat(sprintf("Selected %s: ELLIPSE rank %.0f/%d (|r| = %.3f) | TOEFL rank %.0f/%d (|rho| = %.3f, within-L1 %.3f)\n",
            sel_cfg, sel_row$rank_ellipse, nrow(xc), sel_row$abs_r,
            sel_row$rank_toefl, nrow(xc), sel_row$abs_rho, sel_row$abs_rho_within))
```

```{r}
#| label: cross-corpus-agreement
#| fig-width: 6.5
#| fig-height: 5
# TOEFL is resampled UNstratified here. The stratified resample used for the
# fairness axis holds band composition fixed, which would remove the very
# variation this correlation measures.
n_toe      <- nrow(M_toe)
boot_agree <- numeric(N_BOOT)
for (b in seq_len(N_BOOT)) {
  ei <- sample.int(n_ell, n_ell, replace = TRUE)
  ab <- abs(cor(S_ell[ei, ], y_ell[ei])[, 1])
  ti <- sample.int(n_toe, n_toe, replace = TRUE)
  at <- abs(cor(R_toe[ti, ], rl_toe[ti])[, 1])
  boot_agree[b] <- cor(ab, at, method = "spearman")
}

agree_hat <- cor(points$abs_r, points$abs_rho, method = "spearman")
cat(sprintf("rank agreement of the two validity orderings (%d configs): rho = %.3f, 95%% CI [%.3f, %.3f]\n",
            nrow(points), agree_hat, q025(boot_agree), q975(boot_agree)))
cat(sprintf("pooled vs within-L1 TOEFL ordering                       : rho = %.3f\n",
            cor(points$abs_rho, points$abs_rho_within, method = "spearman")))

pxc_df <- points

p_xc <- ggplot(pxc_df, aes(abs_rho, abs_r)) +
  geom_point(aes(colour = window, shape = instruct), size = 2.6) +
  ggrepel::geom_text_repel(
    data = filter(pxc_df, front | config == sel_cfg),
    aes(label = config), size = 2.5, max.overlaps = Inf, box.padding = 0.5,
    min.segment.length = 0, segment.colour = "grey55", seed = 42) +
  geom_point(data = filter(pxc_df, config == sel_cfg),
             shape = 21, size = 5, stroke = 1, colour = "black", fill = NA) +
  labs(x = "TOEFL 11 validity (|Spearman rho| vs proficiency band)",
       y = "ELLIPSE validity (|Pearson r| vs holistic score)",
       colour = "Window", shape = "Instruct",
       title = "Does the validity ordering survive changing the criterion corpus?",
       subtitle = paste("Circled point is the selected default. Axes are not on a common scale.",
                        sep = "\n")) +
  theme_minimal(base_size = 10)
p_xc
for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}
save_fig(p_xc, "validity_cross_corpus", width = 6.5, height = 5)
write_csv(select(xc, config, abs_r, rank_ellipse, abs_rho, abs_rho_within, rank_toefl),
          file.path(RESULTS, "validity_cross_corpus.csv"))
```

The ordering transfers. Rank agreement across the configurations is high, with a
tight bootstrap interval, so the ELLIPSE ranking is a property of the
configurations rather than of the corpus that supplied the criterion — the
assumption the two-corpus design rests on.

The selected default is not near the top on both, and the reason is the
architecture-by-window interaction rather than anything about the selection.
`olmo2-1b/8` is 5th of 32 on ELLIPSE and 12th on TOEFL 11, because the masked
models at window 64 lead the TOEFL ordering while the causal window-8 class
leads on ELLIPSE. Within its own class it is 2nd of 6 on TOEFL, behind
`llama3.1-8b/8` and ahead of both `olmo2-7b/8` (14th) and `qwen2.5-7b/8`
(16th) — so the cross-corpus check does not favour any of the alternatives the
class rule had to choose between.

Three further readings. The pooled and within-L1 TOEFL orderings agree at 0.966,
so the band-by-L1 confounding documented in
`proficiency-confound-sensitivity.qmd` does not drive the validity ordering.
The window effect replicates across corpora, with mean validity falling from
window 8 to 64 to full on both (0.626, 0.574, 0.527 on ELLIPSE; 0.499, 0.462,
0.394 on TOEFL), although it is architecture-dependent, since masked models are
level or better at window 64 while causal models are clearly best at window 8.
The excluded `modernbert-base/full` configuration reached |rho| = 0.034 on
TOEFL 11 before it was dropped, near zero on a corpus and criterion entirely
independent of the evidence used to diagnose it.

## Validity summary and truncation

Plan §3.1 asks for a tidy table over model, window and corpus carrying *r*, its
interval, *n*, and the truncation rate. Truncation affects only the models whose
context is shorter than the longest essays, namely BERT at 512 and the GPT-2
family at 1024. ModernBERT (8192) and every 4k-plus causal model truncate
nothing.

A truncation rate alone does not establish that truncation is harmless. The
test is whether `|r|` differs between the truncated and the complete essays
*within* an affected configuration. Those two subsets are disjoint samples, so
the comparison uses `cocor`'s independent-groups Zou interval rather than the
dependent-overlapping form used elsewhere in this document.

```{r}
#| label: validity-truncation
trunc_rate <- ell |>
  group_by(config) |>
  summarise(n = n(), trunc_rate = mean(truncated), .groups = "drop")

# Carries the fairness columns and the eligible-set front flag as well, so the
# manuscript's summary table is a straight read of one file.
validity_table <- points |>
  select(config, model, window, arch, instruct, abs_r, absr_lo, absr_hi, abs_rho,
         max_abs_d, worst_l1, n_flag, eligible, front) |>
  # NA rather than FALSE for the contrast rows, matching front_prob: those
  # configurations were not evaluated for domination, they did not fail it.
  mutate(front = ifelse(eligible, front, NA)) |>
  left_join(trunc_rate, by = "config") |>
  arrange(desc(abs_r))

validity_table |>
  mutate(across(c(abs_r, absr_lo, absr_hi, abs_rho, max_abs_d), \(x) round(x, 3)),
         front = ifelse(instruct, NA, front),
         trunc_rate = sprintf("%.1f%%", 100 * trunc_rate)) |>
  knitr::kable(caption = paste(
    "Per-configuration validity and fairness on both corpora, with ELLIPSE truncation rate.",
    "Front is the eligible-set flag; contrast configurations show NA."))
write_csv(validity_table, file.path(RESULTS, "validity_table.csv"))

# Truncated vs complete essays within each affected configuration.
split_r <- ell |>
  filter(config %in% (trunc_rate |> filter(trunc_rate > 0) |> pull(config))) |>
  group_by(config) |>
  group_modify(\(d, k) {
    tr <- filter(d, truncated); co <- filter(d, !truncated)
    r1 <- abs(cor(tr$mean_surprisal_bits, tr$overall))
    r2 <- abs(cor(co$mean_surprisal_bits, co$overall))
    ci <- cocor.indep.groups(r1.jk = r1, r2.hm = r2, n1 = nrow(tr), n2 = nrow(co),
                             test = "zou2007")@zou2007$conf.int
    tibble(n_truncated = nrow(tr), n_complete = nrow(co),
           r_truncated = r1, r_complete = r2, diff = r1 - r2, lo = ci[1], hi = ci[2])
  }) |>
  ungroup() |>
  mutate(excludes_0 = (lo > 0) | (hi < 0))

split_r |>
  mutate(across(c(r_truncated, r_complete, diff, lo, hi), \(x) round(x, 4))) |>
  knitr::kable(caption = "Validity on truncated vs complete essays, within affected configurations (independent-groups Zou intervals).")
```

Truncation is a footnote rather than a confound, but for two different reasons.
`bert-base` truncates 38.2% of ELLIPSE, yet the truncated and complete subsets
give almost the same `|r|`, within 0.01 at every window. The GPT-2 family shows
a larger split, with truncated essays correlating more strongly, but those
essays are only 4.6% of the corpus, so the aggregate estimate moves by well
under 0.01. The split is also confounded with length by construction, since the
truncated subset is exactly the longest essays, so it indexes essay length at
least as much as truncation itself.

## Validity heat map

Plan §3.1 calls for a heat map over model and context window, faceted by corpus.
Fill encodes **within-corpus rank** rather than the raw coefficient, because a
Pearson *r* against a continuous score and a Spearman rho against a
three-category band are not on a common scale. Cell labels carry the raw values.

```{r}
#| label: validity-heatmap
#| fig-width: 8
#| fig-height: 6
hm <- points |>
  select(config, model, window, abs_r, abs_rho) |>
  pivot_longer(c(abs_r, abs_rho), names_to = "corpus", values_to = "value") |>
  mutate(corpus = factor(corpus, levels = c("abs_r", "abs_rho"),
                         labels = c("ELLIPSE (|Pearson r|)", "TOEFL 11 (|Spearman rho|)")),
         model  = factor(model, levels = rev(MODELS)),
         window = factor(window, levels = WINDOWS)) |>
  group_by(corpus) |>
  mutate(fill_rank = rank(value) / n()) |>
  ungroup()

p_hm <- ggplot(hm, aes(window, model, fill = fill_rank)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.3f", value), colour = fill_rank > 0.55), size = 2.7) +
  scale_colour_manual(values = c("FALSE" = "grey15", "TRUE" = "white"), guide = "none") +
  facet_wrap(~corpus) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "Within-corpus\nrank",
                       labels = scales::percent_format(accuracy = 1)) +
  labs(x = "Context window", y = NULL,
       title = "Concurrent validity by model and context window",
       subtitle = "Fill is rank within corpus; the two corpora are not on a common scale. Labels are raw coefficients.") +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(), plot.subtitle = element_text(size = 8, colour = "grey30"))
p_hm
save_fig(p_hm, "validity_heatmap", width = 8, height = 6)
```

The map makes the architecture-by-window interaction visible directly. The three
masked models hold their validity from window 8 to 64 on both corpora, while
every causal model falls away sharply over the same step, which is the pattern
the targeted comparisons below quantify.

## Targeted comparisons (plan §3.4)

Every configuration is estimated on the same 6,482 ELLIPSE essays against the
same criterion, so the correlations are dependent and overlapping and cannot be
compared by checking whether their marginal intervals touch. Differences are
therefore reported as **Zou (2007) intervals for dependent overlapping
correlations** (plan §3.1), computed with `cocor`.

The criterion is sign-flipped before each comparison so that intervals are on
the difference in `|r|` directly. Every correlation here is negative, and Zou's
method is invariant to flipping the shared variable.

Three families are pre-specified in plan §3.4 (architecture at the ~125M tier,
context window within each model, and base versus instruct). A fourth,
selection-critical set is **post-hoc**, defined by the outcome, and is labelled
as such. Intervals are reported throughout. Holm-adjusted *p*-values from
`hittner2003` are included per family for readers who want them, but the
intervals are the intended reading.

```{r}
#| label: targeted-setup
# Zou interval on the difference in |r| for two configurations, plus a
# Holm-adjustable p-value from an independent test in the same family.
zou_pair <- function(a, b) {
  r1 <- -cor(S_ell[, a], y_ell); r2 <- -cor(S_ell[, b], y_ell)
  z <- cocor.dep.groups.overlap(r.jk = r1, r.jh = r2, r.kh = cor(S_ell[, a], S_ell[, b]),
                                n = nrow(S_ell), test = c("zou2007", "hittner2003"))
  tibble(pair = paste(a, "vs", b), abs_r_a = r1, abs_r_b = r2, diff = r1 - r2,
         lo = z@zou2007$conf.int[1], hi = z@zou2007$conf.int[2],
         p = z@hittner2003$p.value)
}

# Paired bootstrap difference, reusing the draws already computed above. Works
# for max |d| too, where no Zou analogue exists.
boot_pair <- function(M, a, b) {
  d <- M[, a] - M[, b]
  tibble(pair = paste(a, "vs", b), diff = mean(M[, a]) - mean(M[, b]),
         lo = q025(d), hi = q975(d))
}

fam_table <- function(tb, caption) {
  tb |>
    mutate(p_holm = p.adjust(p, method = "holm"),
           excludes_0 = (lo > 0) | (hi < 0)) |>
    mutate(across(c(abs_r_a, abs_r_b, diff, lo, hi), \(x) round(x, 4)),
           across(c(p, p_holm), \(x) signif(x, 3))) |>
    knitr::kable(caption = caption)
}
```

### Architecture at the ~125M tier

`bert-base` (110M, masked) and `modernbert-base` (149M, masked) against `gpt2`
(124M, generative), within each window.

```{r}
#| label: fam-architecture
have <- function(...) all(c(...) %in% config_cols)

fam_arch <- expand_grid(masked = c("bert-base", "modernbert-base"), window = WINDOWS) |>
  purrr::pmap(\(masked, window) {
    a <- paste0(masked, "/", window); b <- paste0("gpt2/", window)
    if (have(a, b)) zou_pair(a, b) else NULL
  }) |>
  bind_rows()
fam_table(fam_arch, "Masked vs generative at the ~125M tier. Positive diff favours the masked model.")
```

### Context window, within each model

```{r}
#| label: fam-window
fam_win <- MODELS |>
  purrr::map(\(m) purrr::map(list(c("8", "64"), c("8", "full"), c("64", "full")), \(w) {
    a <- paste0(m, "/", w[1]); b <- paste0(m, "/", w[2])
    if (have(a, b)) zou_pair(a, b) else NULL
  }) |> bind_rows()) |>
  bind_rows()
fam_table(fam_win, "Window contrasts within each model. Positive diff favours the shorter window.")
```

### Post-training, base vs instruct

Compared on both axes, as plan §3.4 requires. Validity uses Zou; max |d| has no
Zou analogue and uses the paired bootstrap draws.

```{r}
#| label: fam-instruct
fam_inst <- purrr::map(INSTRUCT_PAIRS, \(pr)
    purrr::map(WINDOWS, \(w) zou_pair(paste0(pr[1], "/", w), paste0(pr[2], "/", w))) |> bind_rows()) |>
  bind_rows()
fam_table(fam_inst, "Base vs instruct, validity. Positive diff favours the base model.")

purrr::map(INSTRUCT_PAIRS, \(pr)
    purrr::map(WINDOWS, \(w) boot_pair(boot_maxd, paste0(pr[1], "/", w), paste0(pr[2], "/", w))) |> bind_rows()) |>
  bind_rows() |>
  mutate(excludes_0 = (lo > 0) | (hi < 0), across(c(diff, lo, hi), \(x) round(x, 4))) |>
  knitr::kable(caption = "Base vs instruct, max |d| (paired bootstrap). Positive diff means the base model is LESS fair.")
```

### Selection-critical set (post-hoc)

Defined by the outcome rather than pre-specified, so these intervals are
descriptive and carry no multiplicity adjustment. The comparison is the selected
default against every eligible configuration that outranks it on validity, which
under the class-based rule is the set the selection has to answer for.

```{r}
#| label: fam-selection
better <- points |>
  filter(eligible, abs_r > points$abs_r[points$config == sel_cfg]) |>
  arrange(desc(abs_r)) |> pull(config)

purrr::map(better, \(b) zou_pair(sel_cfg, b)) |> bind_rows() |>
  mutate(other = better,
         other_on_front = points$front[match(better, points$config)],
         other_max_d    = points$max_abs_d[match(better, points$config)],
         excludes_0     = (lo > 0) | (hi < 0)) |>
  select(pair, diff, lo, hi, excludes_0, other_on_front, other_max_d) |>
  mutate(across(c(diff, lo, hi, other_max_d), \(x) round(x, 4))) |>
  knitr::kable(caption = paste("Selected default vs higher-validity eligible configurations.",
                               "Negative diff means the other configuration has higher |r|."))
```

### What the targeted comparisons show

**Architecture is window-dependent, not uniform.** The plan (§4) expects masked
to beat comparable generative at this tier, and at window 64 it does clearly,
with both masked models ahead of `gpt2` by about 0.047. At window 8 the
advantage disappears, with `modernbert-base` tied against `gpt2`
([-0.0096, +0.0010]) and `bert-base` actually behind by 0.0075. The full-window
row for `modernbert-base` is absent because that configuration is excluded.

**Every window contrast is resolved.** All 31 intervals exclude zero — 33 within-
model pairs less the two that `modernbert-base/full` would have entered — so with
6,482 essays and heavily dependent configurations the window effect is estimated
precisely enough that no within-model pair is ambiguous. This is the contrast
that makes the flatness of the top of the validity ranking informative rather
than merely underpowered: the same design that resolves every window contrast
cannot separate the four leading configurations.

**Instruct models are consistently more valid, which is worth reporting.** All
six intervals exclude zero and all six favour the **instruct** model, by 0.004
to 0.023. Plan §4 lists the opposite as an expected direction, but §4 is a set
of sanity checks on output rather than a hypothesis, and the grounds for
excluding instruct models are about construct validity rather than predictive
performance. Post-training optimizes for dialogue and problem-solving, so an
instruct model's token probabilities no longer estimate how probable a text is
in natural English, which is the quantity Studies 2 and 3 require. The
configurations remain a contrast condition rather than eligible defaults (§1).
That they nevertheless track proficiency better is a substantive finding about
what surprisal measures in a post-trained model.

**Nothing in the selected class is distinguishable on validity.** Two eligible
configurations carry nominally higher `|r|` than the selected default, and
neither difference is resolved: `qwen2.5-7b/8` at -0.0021 [-0.0044, +0.0001] and
`olmo2-7b/8` at -0.0007 [-0.0025, +0.0012]. Both intervals include zero, on
6,482 essays with heavily dependent correlations, which is the regime in which
this design resolves every one of the window contrasts. The top of the eligible
ranking is a flat 0.004 of `|r|` spread across four configurations from four
different pretraining corpora, and the data do not order them.

This is what makes the class-based rule the honest one. Were the validity
differences real, the front's validity endpoint would have a claim; they are
not, so the choice falls to criteria the plane does not encode. The cost is on
the fairness axis and is stated rather than argued away: the selected default
carries max |d| = 0.422 against `olmo2-7b/8`'s 0.409, a paired-bootstrap
difference of +0.013 [+0.003, +0.024] that does exclude zero. It is dominated by
that configuration, and the reason for preferring it anyway is that OLMo-2 1B
shares OLMo-2 7B's publicly released pretraining corpus at a seventh of the
parameter count. A 0.013 increase in worst-case displacement, on an axis where
every configuration is already flagging six or more L1 groups, is a smaller
practical cost than an order of magnitude in compute for a validity difference
of 0.0007.

# Section 5 — Pareto Plot

Windows are stacked rather than placed side by side. The figure is reproduced at
roughly a single text column in the dissertation, and three panels across that
width leaves the point labels below legible size; one panel per row keeps the
full width for each.

```{r}
#| label: pareto-plot
#| fig-width: 6.5
#| fig-height: 9.4
# Shared visual language (palette keyed to model family, theme, save helper).
for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}

pa <- points |>
  mutate(family = factor(model_family(model), levels = FAMILY_LEVELS),
         lbl    = short_model(model),
         window = factor(window, levels = c("8", "64", "full")))
win_labs <- c("8" = "window 8", "64" = "window 64", "full" = "window full")
selected <- selected |> mutate(window = factor(window, levels = c("8", "64", "full")))

p <- ggplot(pa, aes(max_abs_d, abs_r)) +
  geom_vline(xintercept = D_FLAG, linetype = "dotted", colour = "grey55") +
  # bootstrap CIs shown only for front configs (uncertainty that changes the pick)
  geom_errorbar(data = filter(pa, front), aes(ymin = absr_lo, ymax = absr_hi),
                width = 0, alpha = 0.35, colour = "grey55") +
  geom_errorbar(data = filter(pa, front), aes(xmin = maxd_lo, xmax = maxd_hi),
                orientation = "y", width = 0, alpha = 0.35, colour = "grey55") +
  # Contrast configurations are drawn (triangles, lighter) but took no part in
  # the domination comparison, so they are never circled.
  geom_point(aes(colour = family, shape = instruct, alpha = instruct), size = 3.1) +
  geom_point(data = filter(pa, front), shape = 21, size = 5.6, stroke = 1,
             colour = "black", fill = NA) +
  # The selection is not a front member, so it needs a marker the eye separates
  # from the circles rather than one drawn on top of them: a filled black
  # diamond, called out by name.
  geom_point(data = selected, shape = 23, size = 3.4, stroke = 0.9,
             colour = "black", fill = "black") +
  ggrepel::geom_text_repel(aes(label = lbl, colour = family, alpha = instruct),
                           size = 3.1, max.overlaps = Inf, box.padding = 0.4,
                           min.segment.length = 0, segment.colour = "grey65",
                           seed = 42, show.legend = FALSE) +
  ggrepel::geom_label_repel(
    data = mutate(selected, tag = "selected default"), aes(label = tag),
    size = 3.0, fontface = "bold", colour = "black", fill = "white",
    label.size = 0.3, label.padding = grid::unit(0.14, "lines"),
    box.padding = 0.3, point.padding = 0.4, nudge_y = -0.072, nudge_x = 0.02,
    direction = "y", min.segment.length = 0, segment.colour = "black",
    segment.size = 0.4, seed = 42) +
  annotate("text", x = D_FLAG, y = 0.503, label = "|d| = 0.10", hjust = -0.06,
           size = 3.1, colour = "grey45") +
  facet_wrap(~window, ncol = 1, labeller = labeller(window = win_labs)) +
  scale_colour_manual(values = FAMILY_COLORS, drop = FALSE) +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17),
                     labels = c("pretrained (eligible)", "instruct (contrast)")) +
  scale_alpha_manual(values = c(`FALSE` = 1, `TRUE` = 0.55), guide = "none") +
  coord_cartesian(xlim = c(D_FLAG, 0.66), ylim = c(0.50, 0.66)) +
  labs(
    x = "Unfairness — max |Cohen's d| across L1 groups, TOEFL 11 (worse →)",
    y = "Validity — |Pearson r| with holistic score, ELLIPSE (better ↑)",
    colour = "Model family", shape = NULL,
    title = "Fairness–validity plane, by context window",
    subtitle = paste("Circled = non-dominated within the eligible set. Contrast models are plotted",
                     "but excluded from the domination comparison. CIs (front only) = 95%",
                     "bootstrap; dotted = |d| = 0.10 flag.",
                     sep = "\n")) +
  theme_study1(12) +
  theme(legend.position = "bottom", legend.box = "vertical",
        legend.margin = margin(t = 0, b = 0), legend.spacing.y = grid::unit(0, "pt"),
        legend.key.spacing.x = grid::unit(4, "pt"),
        plot.subtitle = element_text(size = 9),
        plot.margin = margin(5.5, 11, 5.5, 5.5)) +
  guides(colour = guide_legend(nrow = 2, order = 1, title.position = "top",
                               override.aes = list(size = 2.6, alpha = 1)),
         shape  = guide_legend(nrow = 1, order = 2,
                               override.aes = list(size = 2.6, alpha = 1)))

p
save_fig(p, "pareto_front", width = 6.5, height = 9.4)
```

# Section 6 — Outputs & Notes

```{r}
#| label: write-table
out <- points |>
  mutate(selected = config == sel_cfg) |>
  arrange(desc(front), desc(abs_r)) |>
  mutate(front = ifelse(eligible, front, NA)) |>
  select(model, window, arch, instruct, eligible, r, abs_r, max_abs_d, worst_l1,
         n_flag, front, front_prob, selected, absr_lo, absr_hi, maxd_lo, maxd_hi)
write_csv(out, file.path(RESULTS, "pareto_points.csv"))
cat("wrote", file.path(RESULTS, "pareto_points.csv"), "and pareto_front.png\n")
```

**Interpretation caveats (for the write-up):**

- **All correlations are negative** (more predictable text ↔ higher proficiency);
  validity is reported as the magnitude `|r|`.
- **Fairness violations are pervasive.** Every configuration flags multiple L1
  groups at |d| ≥ 0.10, so the front reflects *relative*, not absolute, fairness.
  Report the selected configuration's max |d| and flagged-group count explicitly.
- **Thin low-proficiency strata for some L1s** (e.g. GER-low n≈14, HIN-low n≈25)
  make those stratum-level *d*'s noisy; the bootstrap envelopes capture this.
- **Instruct models are a contrast condition**, excluded from the domination
  comparison as well as from selection. `front` and `front_prob` are defined
  over the eligible set only and are NA for those six configurations.
- **The selected default is not on the front.** The class-based rule prefers the
  smallest open-pretraining-data model in the highest-validity class, and that
  configuration is dominated by its 7B sibling on a validity difference the data
  cannot resolve. Report it as a deliberate departure, with the max |d| cost.
- **Truncation:** BERT (512) and GPT-2 (1024) truncate some long essays at the
  full window (flagged per-row upstream); relevant only for those specific points.

# Reproducibility

```{r}
#| label: sessioninfo
cat(sprintf("N_BOOT = %d | seed = 42\n", N_BOOT))
sessionInfo()
```
