---
title: "Study 1. Within-Band Proficiency as a Confound for Cross-L1 Displacement"
subtitle: "Bounding how much of the observed fairness gap the three-level proficiency scale could be hiding"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

The fairness axis in `pareto-analysis.qmd` conditions on the TOEFL 11
proficiency label, which has only three levels (low, medium, high).
Conditioning on a three-category coarsening of a continuous variable does not
remove the variation inside a category, so a cross-L1 displacement that survives
stratification could still be a proficiency effect rather than an L1 effect.
Residualizing surprisal on the band label instead of stratifying by it does not
help, since both use the same three categories and nothing finer.

The objection is legitimate and quantifiable. Cochran (1968) showed that
subclassification on a normally distributed confounder removes roughly 64, 79,
86 and 90 percent of its bias for two, three, four and five strata. Three bands
therefore leave on the order of **one fifth** of the proficiency bias in place.

TOEFL 11 carries no finer proficiency measure, so the residual cannot be
controlled directly. It can be bounded. Section 2 gives the bound as three lines
of arithmetic. Section 3 corroborates it with `sensemakr`, an implementation of
the omitted-variable-bias framework of Cinelli and Hazlett (2020). Appendix A
reaches the same conclusion through a latent-variable model, and is included
only because it was run; nothing in Sections 2 or 3 depends on it.

All three agree. For the worst-case L1 group in the selected configuration, the
residual confounding available under Cochran's rule of thumb falls short of the
observed displacement by a wide margin, and no configuration in the matrix is
explained away at its worst-case group.

# Setup and Data

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ordinal)      # Christensen — cumulative link models with a scale term
library(truncnorm)    # Mersmann et al. — truncated normal moments
library(sensemakr)    # Cinelli & Hazlett — omitted variable bias bounds

BANDS    <- c("low", "medium", "high")
WINDOWS  <- c("8", "64", "full")
SELECTED <- "olmo2-1b/8"      # the Study 2/3 default from pareto-analysis.qmd
# Configurations dropped as instrument failures rather than data. See "Known
# issues" in this directory's README for the diagnosis and the evidence.
EXCLUDED_CONFIGS <- c("modernbert-base/full")

# Cochran (1968): three strata on a normal confounder remove about 79% of its
# bias, so about 21% survives. COCHRAN_CONSERVATIVE is carried alongside because
# Cochran's figure assumes equal-size strata and these bands are far from equal.
COCHRAN_RESIDUAL     <- 0.21
COCHRAN_CONSERVATIVE <- 0.35

find_results <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    cand <- file.path(d, "results", "predictability")
    if (dir.exists(cand)) return(cand)
    d <- dirname(d)
  }
  stop("results/predictability not found — run assemble_analysis_data.py first")
}
RESULTS <- find_results()
```

```{r}
#| label: load-data
toe <- read_csv(file.path(RESULTS, "toefl_long.csv.gz"),
                col_types = cols(window = col_character())) |>
  mutate(config = paste(model, window, sep = "/"))

toe_wide <- toe |>
  select(text_id, config, mean_surprisal_bits) |>
  pivot_wider(names_from = config, values_from = mean_surprisal_bits) |>
  left_join(distinct(toe, text_id, l1, level), by = "text_id")

config_cols <- setdiff(names(toe_wide), c("text_id", "l1", "level"))
config_cols <- setdiff(config_cols, EXCLUDED_CONFIGS)
M_toe     <- as.matrix(toe_wide[, config_cols])
l1_toe    <- toe_wide$l1
level_toe <- factor(toe_wide$level, levels = BANDS)
GROUPS    <- sort(unique(l1_toe))

cat(sprintf("configs: %d | TOEFL essays: %d | L1 groups: %d\n",
            length(config_cols), nrow(M_toe), length(GROUPS)))
```

# Section 1. The Objection, and Why It Is Legitimate

Two facts make within-band proficiency a candidate explanation rather than a
theoretical worry. First, proficiency composition differs sharply across L1
groups. Second, the three-level control already absorbs a large share of the
raw L1 differences, which is what a variable does when it is doing real work.

```{r}
#| label: composition
knitr::kable(round(prop.table(table(l1_toe, level_toe), margin = 1), 3),
             caption = "Proficiency composition within each L1 group (row proportions).")
```

```{r}
#| label: shrinkage
d_unadjusted <- function(x, l1) vapply(GROUPS, function(g) {
  f <- x[l1 == g]; r <- x[l1 != g]; n1 <- length(f); n2 <- length(r)
  (mean(f) - mean(r)) / sqrt(((n1 - 1) * var(f) + (n2 - 1) * var(r)) / (n1 + n2 - 2))
}, numeric(1))

# Focal-size-weighted Cohen's d within proficiency strata: the fairness axis.
d_stratified <- function(x, l1, level) vapply(GROUPS, function(g) {
  num <- 0; wsum <- 0
  for (s in BANDS) {
    in_s <- level == s
    f <- x[in_s & l1 == g]; r <- x[in_s & l1 != g]
    n1 <- length(f); n2 <- length(r)
    if (n1 < 2L || n2 < 2L) next
    sp <- sqrt(((n1 - 1) * var(f) + (n2 - 1) * var(r)) / (n1 + n2 - 2))
    num <- num + n1 * ((mean(f) - mean(r)) / sp); wsum <- wsum + n1
  }
  num / wsum
}, numeric(1))

x_sel <- M_toe[, SELECTED]
shrink <- tibble(l1 = GROUPS,
                 d_unadjusted = d_unadjusted(x_sel, l1_toe),
                 d_stratified = d_stratified(x_sel, l1_toe, level_toe)) |>
  mutate(absorbed_pct = 100 * (1 - abs(d_stratified) / abs(d_unadjusted))) |>
  arrange(desc(abs(d_unadjusted)))

knitr::kable(shrink, digits = 3,
             caption = paste0("Effect of the three-level control on cross-L1 d (", SELECTED, ")."))
cat(sprintf("max |d| unadjusted %.3f (%s) -> stratified %.3f (%s); median absorbed %.1f%%\n",
            max(abs(shrink$d_unadjusted)), shrink$l1[which.max(abs(shrink$d_unadjusted))],
            max(abs(shrink$d_stratified)), shrink$l1[which.max(abs(shrink$d_stratified))],
            median(shrink$absorbed_pct)))
```

Cochran's result puts a number on what is left over. Three strata remove about
79 percent of a normal confounder's bias, so roughly 21 percent survives. The
question for Sections 2 and 3 is whether that remainder is large enough to
account for the displacements that set the fairness axis.

# Section 2. A Rule-of-Thumb Bound

Three steps, each a single quantity.

## Step 1. How far apart are the L1 groups in proficiency?

The band composition identifies it. A cumulative probit with a scale term
(`ordinal::clm`, following McCullagh 1980) fits each group's latent proficiency
against the rest of the sample, in standard deviation units.

```{r}
#| label: step1-gap
latent_gap <- lapply(GROUPS, function(g) {
  d <- tibble(level = factor(level_toe, levels = BANDS, ordered = TRUE),
              focal = factor(ifelse(l1_toe == g, "focal", "rest"), levels = c("rest", "focal")))
  m <- clm(level ~ focal, scale = ~ focal, link = "probit", data = d)
  tibble(l1 = g,
         mu_gap    = unname(coef(m)[["focalfocal"]]),
         mu_se     = unname(sqrt(diag(vcov(m)))[["focalfocal"]]),
         scale_ratio = unname(exp(m$zeta[["focalfocal"]])))
}) |> bind_rows() |> arrange(desc(abs(mu_gap)))

knitr::kable(latent_gap, digits = 3,
             caption = "Latent proficiency gap versus the rest of the sample, in theta SD (ordinal::clm, probit).")
mu_gap <- setNames(latent_gap$mu_gap, latent_gap$l1)
```

The scale ratios sit near 1, so the groups differ mainly in location. The
largest gap in the corpus is GER, and that is the group that gives the confound
the most room to work.

## Step 2. How much of that gap survives the three-level control?

Cochran's figure: about 21 percent.

```{r}
#| label: step2-residual
residual_gap <- COCHRAN_RESIDUAL * mu_gap
cat(sprintf("largest latent gap: %s at %+.3f theta SD -> residual after 3 strata %+.3f theta SD\n",
            names(which.max(abs(mu_gap))), mu_gap[[which.max(abs(mu_gap))]],
            residual_gap[[which.max(abs(mu_gap))]]))
```

## Step 3. What is a proficiency SD worth in surprisal?

The slope is estimated **within each L1 group** and then pooled, so that L1
composition cannot contaminate the very quantity used to test it. Band mean
proficiency comes from `truncnorm::etruncnorm` on the cutpoints implied by the
observed band shares.

```{r}
#| label: step3-slope
band_p    <- prop.table(table(level_toe))
cuts      <- qnorm(cumsum(c(0, as.numeric(band_p))))
theta_bar <- vapply(seq_along(BANDS), function(i) etruncnorm(cuts[i], cuts[i + 1], 0, 1), numeric(1))
theta_var <- vapply(seq_along(BANDS), function(i) vtruncnorm(cuts[i], cuts[i + 1], 0, 1), numeric(1))
names(theta_bar) <- names(theta_var) <- BANDS

n_by_l1 <- as.numeric(table(factor(l1_toe, levels = GROUPS))); names(n_by_l1) <- GROUPS
beta_by_l1 <- matrix(NA_real_, length(GROUPS), ncol(M_toe), dimnames = list(GROUPS, colnames(M_toe)))
for (g in GROUPS) {
  use <- character(0); ns <- numeric(0); Y <- NULL
  for (s in BANDS) {
    idx <- which(l1_toe == g & level_toe == s)
    if (length(idx) < 2L) next
    use <- c(use, s); ns <- c(ns, length(idx)); Y <- rbind(Y, colMeans(M_toe[idx, , drop = FALSE]))
  }
  if (length(use) < 2L) next
  x  <- theta_bar[use]; xb <- weighted.mean(x, ns); yb <- colSums(ns * Y) / sum(ns)
  beta_by_l1[g, ] <- colSums(ns * (x - xb) * sweep(Y, 2, yb)) / sum(ns * (x - xb)^2)
}
ok   <- !is.na(beta_by_l1[, 1])
beta <- colSums(n_by_l1[ok] * beta_by_l1[ok, , drop = FALSE]) / sum(n_by_l1[ok])

cat(sprintf("%s: %.4f bits per theta SD (within-L1 range %.3f to %.3f)\n",
            SELECTED, beta[[SELECTED]], min(beta_by_l1[, SELECTED]), max(beta_by_l1[, SELECTED])))
```

## Putting the three together

The residual proficiency gap times the slope gives the surprisal difference
that within-band confounding could produce. Compare it with what was actually
observed, as a focal-size-weighted within-band gap in bits, on the same estimand
as the fairness axis.

```{r}
#| label: rule-of-thumb
observed_gap_bits <- function(M, l1, level) {
  D <- matrix(NA_real_, length(GROUPS), ncol(M), dimnames = list(GROUPS, colnames(M)))
  for (g in GROUPS) {
    num <- rep(0, ncol(M)); w <- 0
    for (s in BANDS) {
      in_s <- level == s
      foc <- in_s & (l1 == g); rst <- in_s & (l1 != g)
      if (sum(foc) < 2L || sum(rst) < 2L) next
      num <- num + sum(foc) * (colMeans(M[foc, , drop = FALSE]) - colMeans(M[rst, , drop = FALSE]))
      w   <- w + sum(foc)
    }
    D[g, ] <- num / w
  }
  D
}

OBS <- observed_gap_bits(M_toe, l1_toe, level_toe)
# Worst-case L1 per config is defined on the standardized fairness axis, so that
# this table refers to the same group that sets max |d|.
absD    <- abs(apply(M_toe, 2, function(x) d_stratified(x, l1_toe, level_toe)))
worst_i <- apply(absD, 2, which.max)
pick    <- cbind(worst_i, seq_len(ncol(M_toe)))

rot <- tibble(
  config     = colnames(M_toe),
  worst_l1   = GROUPS[worst_i],
  max_abs_d  = absD[pick],
  obs_bits   = abs(OBS[pick]),
  allow_bits = abs(residual_gap[GROUPS[worst_i]] * beta),
  allow_bits_conservative =
    abs((COCHRAN_CONSERVATIVE / COCHRAN_RESIDUAL) * residual_gap[GROUPS[worst_i]] * beta)
) |>
  mutate(ratio        = obs_bits / allow_bits,
         ratio_conservative = obs_bits / allow_bits_conservative,
         window       = factor(sub(".*/", "", config), levels = WINDOWS),
         selected     = config == SELECTED) |>
  arrange(ratio)

knitr::kable(select(rot, config, worst_l1, max_abs_d, obs_bits, allow_bits,
                    ratio, ratio_conservative),
             digits = 3,
             caption = paste("Observed within-band gap versus the gap residual confounding could produce.",
                             "ratio > 1 means confounding cannot account for it."))

cat(sprintf("configs where confounding suffices (ratio < 1): %d of %d\n",
            sum(rot$ratio < 1), nrow(rot)))
cat(sprintf("under the conservative 35%% residual              : %d of %d\n",
            sum(rot$ratio_conservative < 1), nrow(rot)))
write_csv(rot, file.path(RESULTS, "proficiency_confound_ruleofthumb.csv"))
```

```{r}
#| label: rot-plot
#| fig-width: 8.5
#| fig-height: 8
plot_df <- mutate(rot, config = factor(config, levels = config[order(ratio)]))

p <- ggplot(plot_df, aes(y = config)) +
  geom_segment(aes(x = allow_bits, xend = obs_bits, yend = config),
               colour = "grey78", linewidth = 0.7) +
  geom_point(aes(x = allow_bits), shape = 23, fill = "white", colour = "black", size = 2) +
  geom_point(aes(x = obs_bits, colour = window, shape = window), size = 2.8) +
  geom_text(aes(x = obs_bits + 0.012, label = worst_l1), size = 2.5, hjust = 0, colour = "grey30") +
  geom_point(data = filter(plot_df, selected), aes(x = obs_bits),
             shape = 21, size = 5.5, stroke = 1, colour = "black", fill = NA) +
  labs(x = "Within-band gap for the worst-case L1 (bits of surprisal)", y = NULL,
       colour = "Window", shape = "Window",
       title = "Observed displacement versus what residual confounding allows",
       subtitle = paste("White diamond: gap that within-band proficiency could produce (Cochran 21% residual).",
                        "Coloured point: gap actually observed. Circled point is the selected default.",
                        sep = "\n")) +
  scale_x_continuous(expand = expansion(mult = c(0.02, 0.10))) +
  theme_minimal(base_size = 10) +
  theme(panel.grid.major.y = element_line(colour = "grey93"),
        plot.subtitle = element_text(size = 8, colour = "grey30"))
p
# Shared save helper (vector PDF + 400 dpi PNG), as in the Pareto reports.
for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}
save_fig(p, "proficiency_confound_ruleofthumb", width = 8.5, height = 8)
```

## What this does and does not say

For the configurations whose worst-case group is TEL, including the selected
default, the observed gap is an order of magnitude beyond what residual
confounding allows, because Telugu writers' proficiency composition is close to
the sample average and a group that looks average between bands cannot be
extreme within them. For GER the margin is roughly threefold, and GER is the
binding case because it has the largest latent gap in the corpus.

The honest qualification is that this does **not** hold for every group. ARA,
HIN and JPN have large latent gaps and small observed displacements, and for
those groups residual within-band confounding is an entirely adequate
explanation. None of them sets `max |d|` for any configuration, so the fairness
axis is unaffected, but the claim that survives is specific. Residual
confounding plausibly explains the small displacements. It does not explain the
ones that determine the axis.

# Section 3. Corroboration with sensemakr

Section 2 rests on Cochran's rule of thumb and a linear slope. `sensemakr`
reaches the question from a different direction, treating within-band
proficiency as an omitted variable in a regression of surprisal on the focal L1
indicator with the band label controlled, and asking how strong that omitted
variable would have to be relative to the band label itself.

The benchmark has a natural value here. By the law of total variance on the
coarsening, the within-band remainder of proficiency carries a fixed share of
the variance that the band label captures.

```{r}
#| label: plausible-k
within_var  <- sum(as.numeric(band_p) * theta_var)
k_plausible <- within_var / (1 - within_var)
cat(sprintf("within-band theta variance %.3f | between-band %.3f | plausible k = %.3f\n",
            within_var, 1 - within_var, k_plausible))
```

So a confounder about **0.3 times** as strong as the observed band label is the
substantively plausible case. Anything much beyond that is asserting the
coarsening lost more than it kept. (`sensemakr`'s *k* is a ratio of partial
R-squared rather than of variance shares, so this is a calibration rather than
an identity.)

```{r}
#| label: sensemakr
# Smallest k at which a confounder k times as strong as `level` nullifies the effect.
tipping_k <- function(m, grid = seq(0.05, 6, by = 0.05)) {
  e0 <- coef(m)[["focal"]]
  b <- suppressWarnings(ovb_bounds(
    m, treatment = "focal",
    benchmark_covariates = list(level = c("levelmedium", "levelhigh")), kd = grid))
  hit <- which(sign(b$adjusted_estimate) != sign(e0))
  if (!length(hit)) return(NA_real_)
  grid[min(hit)]
}

# One regression per configuration, with that configuration's own worst-case L1
# as the focal group, so this table lines up row for row with Section 2.
sens <- lapply(seq_len(ncol(M_toe)), function(j) {
  g <- GROUPS[worst_i[j]]
  d <- tibble(y = M_toe[, j], level = level_toe, focal = as.numeric(l1_toe == g))
  m <- lm(y ~ focal + level, data = d)
  tibble(config = colnames(M_toe)[j], worst_l1 = g,
         estimate_bits    = unname(coef(m)[["focal"]]),
         robustness_value = robustness_value(m, covariate = "focal"),
         tipping_k        = tipping_k(m))
}) |> bind_rows() |>
  mutate(margin   = tipping_k / k_plausible,
         selected = config == SELECTED) |>
  arrange(margin)

knitr::kable(select(sens, config, worst_l1, estimate_bits, robustness_value, tipping_k, margin),
             digits = 3,
             caption = paste0("sensemakr bounds, one regression per configuration on its own worst-case L1. ",
                              "tipping_k is in units of the observed band label; plausible k is ",
                              round(k_plausible, 2), ". NA means not nullified at any k up to 6."))

n_tip <- sum(!is.na(sens$tipping_k))
cat(sprintf("configs nullified at some k <= 6      : %d of %d\n", n_tip, nrow(sens)))
cat(sprintf("configs nullified at the plausible k  : %d of %d\n",
            sum(sens$margin < 1, na.rm = TRUE), nrow(sens)))
write_csv(sens, file.path(RESULTS, "proficiency_confound_sensemakr.csv"))
```

Configurations whose worst-case group is TEL are never nullified within the
searched range, so no confounder of this kind explains them. The binding cases
are the GER configurations, which tip at roughly 0.85 times the band label
against a plausible 0.3, a margin near threefold. That matches the rule-of-thumb
ratio for the same configurations in Section 2. The two methods share the
coarsening model but almost nothing else, so their agreement on the binding case
carries weight.

# Section 4. Agreement Between the Two Methods

Sections 2 and 3 share the coarsening model and the assumption that proficiency
is unidimensional. Beyond that they have little in common. Section 2 works in
bits through a calibrated slope and Cochran's bias-removal figure; Section 3
works in partial R-squared through an omitted-variable-bias bound and never uses
the slope at all. Their rankings can therefore be compared.

```{r}
#| label: agreement
agree <- rot |>
  select(config, worst_l1, ratio) |>
  left_join(select(sens, config, tipping_k, margin), by = "config")

cat(sprintf("Spearman correlation of the two margins: %.3f\n",
            cor(agree$ratio, agree$margin, method = "spearman", use = "complete.obs")))
cat(sprintf("configs cleared by both methods: %d of %d\n",
            sum(agree$ratio > 1 & (agree$margin > 1 | is.na(agree$margin))), nrow(agree)))
cat(sprintf("configs cleared by neither     : %d of %d\n",
            sum(agree$ratio <= 1 & agree$margin <= 1, na.rm = TRUE), nrow(agree)))

knitr::kable(head(arrange(agree, ratio), 8), digits = 3,
             caption = "The eight configurations where confounding has the most room, by both methods.")
```

# Section 5. Summary

- The objection is real. Three bands leave roughly a fifth of the proficiency
  bias unadjusted (Cochran 1968), and L1 composition across bands is sharply
  unequal.
- For the groups that set the fairness axis, the residual is far too small to
  account for the observed displacement. TEL by roughly two orders of magnitude,
  GER by roughly threefold, and no configuration is explained away at its
  worst-case group under either the 21 percent or the conservative 35 percent
  residual.
- `sensemakr` agrees on the binding case from a different set of assumptions,
  and the two methods rank the 32 configurations almost identically
  (Spearman 0.99), with all 32 cleared by both.
- For ARA, HIN and JPN the residual is an adequate explanation, and those
  displacements should not be read as L1 effects. They do not set `max |d|`.
- The bound assumes a monotone coarsening of a unidimensional proficiency
  variable and a linear relationship between proficiency and surprisal over the
  observed range. With three bands, curvature cannot be tested.

# Appendix A. Latent Truncation Model

Retained for completeness. This was the original form of the analysis, before
Cochran's rule of thumb made most of the machinery unnecessary. It fits each
group's latent proficiency distribution, derives the within-band displacement
that distribution implies, and compares it with the displacement the observed
gap would require. **Nothing in Sections 2 to 4 depends on it.**

```{r}
#| label: appendix-latent
col_var <- function(X) { n <- nrow(X); m <- colMeans(X); (colSums(X * X) - n * m * m) / (n - 1) }
theta_sd <- sqrt(theta_var)

# Within-band displacement implied by each group's fitted latent distribution.
implied_delta <- vapply(GROUPS, function(g) {
  d  <- tibble(level = factor(level_toe, levels = BANDS, ordered = TRUE),
               focal = factor(ifelse(l1_toe == g, "focal", "rest"), levels = c("rest", "focal")))
  m  <- clm(level ~ focal, scale = ~ focal, link = "probit", data = d)
  mu <- unname(coef(m)[["focalfocal"]]); sg <- unname(exp(m$zeta[["focalfocal"]]))
  num <- 0; w <- 0
  for (i in seq_along(BANDS)) {
    s  <- BANDS[i]
    mg <- etruncnorm(cuts[i], cuts[i + 1], mu, sg)
    mr <- etruncnorm(cuts[i], cuts[i + 1], 0, 1)
    n1 <- sum(l1_toe == g & level_toe == s)
    num <- num + n1 * ((mg - mr) / theta_sd[[s]]); w <- w + n1
  }
  num / w
}, numeric(1))

# Displacement the observed gap would require, on the same scale.
required_delta <- vapply(seq_along(GROUPS), function(gi) {
  g <- GROUPS[gi]; num <- 0; w <- 0
  for (i in seq_along(BANDS)) {
    s <- BANDS[i]; in_s <- level_toe == s
    foc <- in_s & (l1_toe == g); rst <- in_s & (l1_toe != g)
    if (sum(foc) < 2L || sum(rst) < 2L) next
    gap <- mean(M_toe[foc, SELECTED]) - mean(M_toe[rst, SELECTED])
    num <- num + sum(foc) * ((gap / beta[[SELECTED]]) / theta_sd[[s]]); w <- w + sum(foc)
  }
  num / w
}, numeric(1))
names(required_delta) <- GROUPS

app <- tibble(l1 = GROUPS, delta_implied = implied_delta, delta_required = required_delta) |>
  mutate(ratio = abs(delta_required) / abs(delta_implied)) |>
  arrange(desc(ratio))
knitr::kable(app, digits = 3,
             caption = paste0("Appendix: implied versus required within-band displacement (", SELECTED, ")."))
write_csv(app, file.path(RESULTS, "proficiency_confound_latent_appendix.csv"))
```

The ordering and the conclusion match Section 2. TEL is the most robust group,
GER the binding one, and the groups with small displacements and large latent
gaps are the ones confounding can explain.

# References

Cinelli, C., & Hazlett, C. (2020). Making sense of sensitivity: Extending
omitted variable bias. *Journal of the Royal Statistical Society: Series B*,
82(1), 39–67.

Cochran, W. G. (1968). The effectiveness of adjustment by subclassification in
removing bias in observational studies. *Biometrics*, 24(2), 295–313.

Cornfield, J., Haenszel, W., Hammond, E. C., Lilienfeld, A. M., Shimkin, M. B.,
& Wynder, E. L. (1959). Smoking and lung cancer: Recent evidence and a
discussion of some questions. *Journal of the National Cancer Institute*,
22(1), 173–203.

Frank, K. A., Maroulis, S. J., Duong, M. Q., & Kelcey, B. M. (2013). What would
it take to change an inference? Using Rubin's causal model to interpret the
robustness of causal inferences. *Educational Evaluation and Policy Analysis*,
35(4), 437–460.

Liddell, T. M., & Kruschke, J. K. (2018). Analyzing ordinal data with metric
models: What could possibly go wrong? *Journal of Experimental Social
Psychology*, 79, 328–348.

McCullagh, P. (1980). Regression models for ordinal data. *Journal of the Royal
Statistical Society: Series B*, 42(2), 109–142.

Rosenbaum, P. R., & Rubin, D. B. (1983). Assessing sensitivity to an unobserved
binary covariate in an observational study with binary outcome. *Journal of the
Royal Statistical Society: Series B*, 45(2), 212–218.

VanderWeele, T. J., & Ding, P. (2017). Sensitivity analysis in observational
research: Introducing the E-value. *Annals of Internal Medicine*, 167(4),
268–274.

Williamson, D. M., Xi, X., & Breyer, F. J. (2012). A framework for evaluation
and use of automated scoring. *Educational Measurement: Issues and Practice*,
31(1), 2–13.

R packages: `ordinal` (Christensen), `truncnorm` (Mersmann et al.), `sensemakr`
(Cinelli, Ferwerda & Hazlett).
