---
title: "Study 1 — Why Validity Falls as the Context Window Grows"
subtitle: "A worked example of self-conditioning on a writer's own non-standard forms"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    self-contained: true
execute:
  warning: false
  message: false
---

# Overview

Hypothesis 2 predicted that a wider context window would raise concurrent
validity, on the reasoning that a wider window adds sensitivity to
discourse-level coherence. The benchmark finds the opposite. Validity declines
monotonically from the 8-token window to the full essay, and the decline is
steepest for the generative models, which condition only on the writer's
preceding text.

The conjecture offered for that reversal is that conditioning on a long span of
a writer's own production yields a partly writer-specific estimate. As the
window grows it comes to include more of what the writer has already written,
including any non-standard forms that production contains, and a form that has
already appeared becomes cheaper when it appears again. On this account a
writer who is consistently unusual becomes progressively more predictable, so
what the measure indexes shifts from conventionality with respect to the target
language toward internal self-consistency.

This notebook is not a test of that conjecture across the corpus. It is a
worked example that shows the mechanism operating in one TOEFL 11 response, at
the token level, with the selected Study 1 configuration. Its purpose is to
make the proposed mechanism concrete and to put a number on how much of the
measure's discriminating power the wide window gives up.

Input is `results/predictability/self_conditioning_tokens.csv`, written by
`self_conditioning_example.py`.

# Section 1. Setup

```{r}
#| label: setup
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)

for (hp in c("_viz_helpers.R", "src/1-predictability-benchmark/_viz_helpers.R")) {
  if (file.exists(hp)) { source(hp); break }
}
for (th in c("../util/text_heatmap.R", "src/util/text_heatmap.R")) {
  if (file.exists(th)) { source(th); break }
}

find_results <- function() {
  d <- normalizePath(getwd())
  for (i in 1:6) {
    cand <- file.path(d, "results", "predictability")
    if (dir.exists(cand)) return(cand)
    d <- dirname(d)
  }
  stop("results/predictability not found — run self_conditioning_example.py first")
}
RESULTS <- find_results()

# The essay carrying the figure. It is the one response in the scan that uses
# both the non-standard form and its conventional counterpart, which is what
# lets the two be compared inside a single text with prompt, topic, and writer
# all held constant.
EXAMPLE <- "704896.txt"

# Last token of the excerpt the heatmap renders. Chosen at a sentence boundary
# that closes the opening third of the essay, by which point the writer has used
# the standard form three times and the non-standard form three times. The
# occurrence trend below is computed over the whole essay, not the excerpt.
EXCERPT_END <- 111

WIN_LABELS <- c("8" = "8-token window", "64" = "64-token window",
                "full" = "Full essay as context")
WIN_COLORS <- c("8-token window" = "#0072B2", "64-token window" = "#999999",
                "Full essay as context" = "#CC79A7")

tok <- read_csv(file.path(RESULTS, "self_conditioning_tokens.csv"),
                show_col_types = FALSE) |>
  mutate(window_lab = factor(WIN_LABELS[window], levels = unname(WIN_LABELS)))

ex <- filter(tok, text_id == EXAMPLE)
TARGET   <- "peoples"
STANDARD <- "people"

ex |>
  distinct(text_id, l1, level, prompt) |>
  knitr::kable(caption = "The example response")
```

# Section 2. What the model charges for the error

The essay pluralises *people* as *peoples* twelve times, and also uses the
standard form three times. Both forms can therefore be scored under identical
conditions, and the difference between them is what the measure is charging for
the error.

```{r}
#| label: gap-table
gap <- ex |>
  filter(form != "other") |>
  # The first occurrence of the target has no prior instance to condition on, so
  # it measures something different from the later ones and is held out here.
  filter(!(form == "target" & occurrence == 1)) |>
  group_by(window_lab, form) |>
  summarise(mean_bits = mean(surprisal_bits), n = n(), .groups = "drop") |>
  pivot_wider(names_from = form, values_from = c(mean_bits, n)) |>
  mutate(gap_bits = mean_bits_target - mean_bits_standard)

gap |>
  select(`Context window` = window_lab,
         `"peoples" (later occurrences)` = mean_bits_target,
         `"people" (standard)` = mean_bits_standard,
         `Gap` = gap_bits) |>
  knitr::kable(digits = 2,
               caption = "Mean surprisal in bits, OLMo-2 1B. The gap is what the measure charges for the non-standard form.")
```

```{r}
#| label: baseline-table
# The comparison that rules out the trivial reading. Every token gets cheaper
# with more context, so the question is whether the repeated non-standard form
# gets cheaper faster than the essay's own baseline.
ex |>
  mutate(grp = case_when(
    form == "target" & occurrence == 1 ~ "Target, first occurrence",
    form == "target"                   ~ "Target, later occurrences",
    TRUE                               ~ "All other tokens")) |>
  group_by(grp, window) |>
  summarise(bits = mean(surprisal_bits), .groups = "drop") |>
  pivot_wider(names_from = window, values_from = bits) |>
  mutate(`Drop, 8 to full` = `8` - full) |>
  select(Tokens = grp, `Window 8` = `8`, `Window 64` = `64`,
         `Full essay` = full, `Drop, 8 to full`) |>
  arrange(Tokens) |>
  knitr::kable(digits = 2,
               caption = "Mean surprisal in bits by token group.")
```

# Section 3. Figure

```{r}
#| label: fig-self-conditioning
#| fig-width: 9
#| fig-height: 8.6
heat_dat <- ex |>
  filter(spacy_idx <= EXCERPT_END, window %in% c("8", "full")) |>
  arrange(window_lab, spacy_idx)

# One fill scale for both panels, since the comparison the figure asks the
# reader to make is between the same word in two panels. The top of the range is
# squished rather than extended so the ordinary variation stays legible.
LIMITS <- c(0, 16)

p_heat <- text_heatmap(
  heat_dat, token = "token", value = "surprisal_bits",
  panel = "window_lab", space_after = "space_after",
  highlight = c(TARGET, STANDARD),
  width = 95, limits = LIMITS, label_size = 3.3,
  legend_title = "Surprisal\n(bits)"
) +
  labs(title = "The same words, scored twice",
       subtitle = paste0(
         "Opening third of TOEFL 11 response ", EXAMPLE, ". Outlined tokens are\n",
         "“peoples” (non-standard) and “people” (standard). The first “peoples” is\n",
         "costly under both windows, since nothing earlier in the essay predicts it.\n",
         "The next two are costly only under the 8-token window."))

occ <- ex |>
  filter(form == "target") |>
  select(occurrence, window_lab, surprisal_bits)

p_occ <- ggplot(occ, aes(occurrence, surprisal_bits,
                         colour = window_lab, group = window_lab)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.9) +
  scale_colour_manual(values = WIN_COLORS, name = NULL) +
  scale_x_continuous(breaks = seq(1, max(occ$occurrence), 1)) +
  labs(x = paste0("Occurrence of “", TARGET, "” in reading order"),
       y = "Surprisal (bits)",
       title = "Each time the form reappears",
       subtitle = paste(
         "All twelve occurrences, whole essay. The 8-token window never reaches back far enough to\n",
         "include an earlier use, so it charges close to full price every time. Occurrence 1 has no\n",
         "earlier use for any window, so the gap there is ordinary context benefit rather than repetition.")) +
  theme_study1(base_size = 10) +
  theme(legend.position = "bottom",
        plot.subtitle = element_text(size = 8, colour = "grey30"))

p <- p_heat / p_occ + plot_layout(heights = c(2.4, 1)) +
  plot_annotation(tag_levels = "A")
p
save_fig(p, "self_conditioning_example", width = 9, height = 8.6)
```

# Section 4. Reading the figure

Panel A shows the opening third of the response twice, once scored with an
8-token window and once with the whole essay as context. Outlined tokens are the
non-standard *peoples* and the standard *people*. The excerpt holds the first
three uses of *peoples*. The first stays dark in both panels, since no earlier
use predicts it under either window. The second and third are dark under the
8-token window and take the same pale value as the standard *people* beside them
once the whole essay is available, which is the shift the figure is about.

Panel B follows every occurrence of *peoples* through the essay. Under the
8-token window the twelve occurrences stay between 9.9 and 21.8 bits with no
trend, because the window never reaches back far enough to include an earlier
use. Under the full essay the first occurrence costs 12.9 bits and the later
ones fall towards the floor, reaching 0.6 bits at three separate points.

Two features of the example are worth stating rather than leaving to the reader.

First, the repeated form is also a lexical repetition, and repetition makes any
unpredictable word cheaper. Topic words in this essay fall by comparable
amounts. The claim the example supports is therefore not that errors are treated
specially. It is that standard forms begin cheap and have little room to fall,
while non-standard forms begin expensive and fall furthest, so what erodes as
the window widens is specifically the price of being unconventional.

Second, the three standard *people* tokens sit early in the response while the
*peoples* tokens run to the end, so the two sets are not matched on position.
The gap closes because the non-standard form becomes cheap, not because the
standard form becomes expensive, and the table in Section 2 shows the standard
form barely moving across windows. The asymmetry does not produce the effect,
though it does mean the example is an illustration of a mechanism rather than an
estimate of its size.
